### 2023-06-26 paper.tex
primo salvataggio

%  LaTeX support: latex@mdpi.com 
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

%=================================================================
\documentclass[journal,article,submit,pdftex,moreauthors]{Definitions/mdpi}

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, adolescents, aerobiology, aerospace, agriculture, agriengineering, agrochemicals, agronomy, ai, air, algorithms, allergies, alloys, analytica, analytics, anatomia, animals, antibiotics, antibodies, antioxidants, applbiosci, appliedchem, appliedmath, applmech, applmicrobiol, applnano, applsci, aquacj, architecture, arm, arthropoda, arts, asc, asi, astronomy, atmosphere, atoms, audiolres, automation, axioms, bacteria, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomass, biomechanics, biomed, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, blsf, brainsci, breath, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, chips, cimb, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coasts, coatings, colloids, colorants, commodities, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, covid, crops, cryptography, crystals, csmf, ctn, curroncol, cyber, dairy, data, ddc, dentistry, dermato, dermatopathology, designs, devices, diabetology, diagnostics, dietetics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, entomology, entropy, environments, environsciproc, epidemiologia, epigenomes, est, fermentation, fibers, fintech, fire, fishes, fluids, foods, forecasting, forensicsci, forests, foundations, fractalfract, fuels, future, futureinternet, futurepharmacol, futurephys, futuretransp, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, grasses, gucdd, hazardousmatters, healthcare, hearts, hemato, hematolrep, heritage, higheredu, highthroughput, histories, horticulturae, hospitals, humanities, humans, hydrobiology, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijpb, ijtm, ijtpp, ime, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jal, jcdd, jcm, jcp, jcs, jcto, jdb, jeta, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jvd, jzbg, kidneydial, kinasesphosphatases, knowledge, land, languages, laws, life, liquids, literature, livers, logics, logistics, lubricants, lymphatics, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, merits, metabolites, metals, meteorology, methane, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, microplastics, minerals, mining, modelling, molbank, molecules, mps, msf, mti, muscles, nanoenergyadv, nanomanufacturing,\gdef\@continuouspages{yes}} nanomaterials, ncrna, ndt, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, %%nri, nursrep, nutraceuticals, nutrients, obesities, oceans, ohbm, onco, %oncopathology, optics, oral, organics, organoids, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacoepidemiology,\gdef\@ISSN{2813-0618}\gdef\@continuous pharmacy, philosophies, photochem, photonics, phycology, physchem, physics, physiologia, plants, plasma, platforms, pollutants, polymers, polysaccharides, poultry, powders, preprints, proceedings, processes, prosthesis, proteomes, psf, psych, psychiatryint, psychoactives, publications, quantumrep, quaternary, qubs, radiation, reactions, receptors, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, rheumato, risks, robotics, ruminants, safety, sci, scipharm, sclerosis, seeds, sensors, separations, sexes, signals, sinusitis, skins, smartcities, sna, societies, socsci, software, soilsystems, solar, solids, spectroscj, sports, standards, stats, std, stresses, surfaces, surgeries, suschem, sustainability, symmetry, synbio, systems, targets, taxonomy, technologies, telecom, test, textiles, thalassrep, thermo, tomography, tourismhosp, toxics, toxins, transplantology, transportation, traumacare, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, venereology, vetsci, vibration, virtualworlds, viruses, vision, waste, water, wem, wevj, wind, women, world, youth, zoonoticdis 
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal. Changing "submit" to "accept" before posting will remove line numbers.

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by: 
% abstract, addendum, article, book, bookreview, briefreport, casereport, comment, commentary, communication, conferenceproceedings, correction, conferencereport, entry, expressionofconcern, extendedabstract, datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, systematicreview, supfile, technicalnote, viewpoint, guidelines, registeredreport, tutorial
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. Remove "pdftex" for (1) compiling with LaTeX & dvi2pdf (if eps figures are used) or for (2) compiling with XeLaTeX.

%=================================================================
% MDPI internal commands - do not modify
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2023}
\copyrightyear{2023}
%\externaleditor{Academic Editor: Firstname Lastname}
\datereceived{ } 
\daterevised{ } % Comment out if no revised date
\dateaccepted{ } 
\datepublished{ } 
%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
%\dateretracted{} % For corrected papers: "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%\pdfoutput=1 % Uncommented for upload to arXiv.org

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}

%=================================================================
% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{A sustainable approach to tourist signage on (historical) trails}

% MDPI internal command: Title for citation in the left column
\TitleCitation{A sustainable approach to tourist signage on historical trails}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{Firstname Lastname $^{1,\dagger,\ddagger}$\orcidA{}, Firstname Lastname $^{2,\ddagger}$ and Firstname Lastname $^{2,}$*}

%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{Firstname Lastname, Firstname Lastname and Firstname Lastname}

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}

% Contact information of the corresponding author
\corres{Correspondence: e-mail@e-mail.com; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}

% Current address and/or shared authorship
\firstnote{Current address: Affiliation 3.} 
\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

%\simplesumm{} % Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{
Guiding the visitor to appreciate all aspects of tourist attractions is paramount for building a memorable experience. In the literature, we find many articles exploring the use of sophisticated technologies towards such a goal. The case of wilderness resources exhibits a peculiar aspect, since the presence of human artifacts and the signage, may, in fact, damage the experience, and introduce pollutants as well. In this paper, we consider the case of a natural trail targeted to niche tourism to understand the available options for signage aiming to provide directions, guide the exploration, and warn about dangers. The footprints of the various options are evaluated, and we describe the outcomes of an experimental setup implemented to verify our conclusions' applicability.
}
%\abstract{A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.}

% Keywords
\keyword{keyword 1; keyword 2; keyword 3 (List three to ten pertinent keywords specific to the article; yet reasonably common within the subject discipline.)}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data
%\dataset{DOI number or link to the deposited data set if the data set is published separately. If the data set shall be published as a supplement to this paper, this field will be filled by the journal editors. In this case, please submit the data set as a supplement.}
%\datasetlicense{License under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Encyclopedia
%\encyclopediadef{For entry manuscripts only: please provide a brief overview of the entry title instead of an abstract.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Advances in Respiratory Medicine
%\addhighlights{yes}
%\renewcommand{\addhighlights}{%

%\noindent This is an obligatory section in “Advances in Respiratory Medicine”, whose goal is to increase the discoverability and readability of the article via search engines and other scholars. Highlights should not be a copy of the abstract, but a simple text allowing the reader to quickly and simplified find out what the article is about and what can be cited from it. Each of these parts should be devoted up to 2~bullet points.\vspace{3pt}\\
%\textbf{What are the main findings?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}\vspace{3pt}
%\textbf{What is the implication of the main finding?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\subsection{The role of tourist signage}

\section{Tourist signage sustainability}

\subsection{Working hypothesis}

\subsection{Evaluation parameters}

\subsection{Evaluation of sample alternatives}

\section{Application to a concrete use case}

\subsection{Trail selection}

\subsection{Social and political aspects}

\subsection{Signage design}

\subsection{Monitoring results}

\section{Conclusions}

\setcounter{section}{-1} %% Remove this when starting to work on the template.
\section{How to Use this Template}

The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions on the authors' page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX-related questions please contact latex@mdpi.com.%\endnote{This is an endnote.} % To use endnotes, please un-comment \printendnotes below (before References). Only journal Laws uses \footnote.

% The order of the section titles is different for some journals. Please refer to the "Instructions for Authors” on the journal homepage.

\section{Introduction}

Although tourism persistently targets most renowned sites, like art cities and parks, there is an increasing interest for targets not exposed to mass tourism, that may provide a more intense and inner experience. Suh an exclusive flavor is appreciated not only by the occasional visitor, which helps the economy of the corner, but also by those that have their homes in such privileged location. The key to sustain such a difficult equilibrium is to guide the visitor to enjoy the place with discrete messages that he finds on his way.

%The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the principal conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research. Citing a journal paper \cite{ref-journal}. Now citing a book reference \cite{ref-book1,ref-book2} or other reference types \cite{ref-unpublish,ref-communication,ref-proceeding}. Please use the command \citep{ref-thesis,ref-url} for the following MDPI journals, which use author--date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, Journal of Intelligence, Journalism and Media, JRFM, Languages, Laws, Religions, Risks, Social Sciences, Literature.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

In our study we have considered a range of solutions that may help towards our goal of a sustainable support to a sustainable tourism, and specifically to the dynamic provision of information during a visit. To this end, the hosting organization (the {\em host} for short) implements a system of devices from which the visitor (or {\em user} ) extracts useful and enjoyable information.

The purpose of such an installation is to guide the {\em user} in a tour that includes hurban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research). The task of the host organization is to provide the user with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.

A traditional approach makes use of physical information boards with graphical or textual contents. Our study starts from finding the issues related with such a solution:

\begin{itemize}
\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
\item installation logistics: depending on the location, the placement of a plaque may require a basement or other sorts of supports
\item environmental impact: in order to be effective, the plaque is prominent, and this may negatively affect the quality of the site 
\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
\item update limits: to update the content, the plaque must be replaced
\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues like those found during the installation
\end{itemize}

Such considerations motivate an interest for an alternative way of communication.

During the last decades we have witnessed the diffusion of smartphone devices, and we have no reasons to expect a change in such a thread. Smartphones empower the communication capabilities of an individual, allowing to receive sound, visual and tatcile interactions. It is therefore straightforward to consider the way to exploit such capabilities in our context. Also in this case we start from the limits of such technology.

Concentrate on outdoor activities

One is related with the diffusion of such a technology, which, although very diffused, is not equally familiar to everybody. The second is that several functions depends on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach an enabling infrastructure.

Starting from the two aspects above we envision the directions towards a successful strategy, keeping in mind the overall sustainability of the approach.

Regarding usability, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of installing new applications.

The second issue is more complex, especially when considering outdoor activities. In such cases the provision of a networking infrastructure may be impossibile without a severe environmental impact. Consider the installation of antennas to cover a wide area and power supply for the radios.

Starting from previous experiences documented in the bibliography we explore a few broad categories, to find weaknesses and strengths for each of them.

One strategy concentrates on the wireless capabilities of the smartphone, although avoiding the coverage of the whole region. We consider the implementation of several small servers operating in isolation that provide a few HTML pages. The approach requires a modest investment and a marginal environmental impact: a small device based on the ESP8266 single chip computer has a coverage fo tens of meters, and a volume in the order of the $cm^3$. They operate as WiFi Access Points, so the user has access to the data in the device by joining the AP and browsing the URL. The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image.

The dependency from power supply is a serious limitation of such kind of solution. A radio transmitter is a rather power consuming device, in the range of Watts. Even if operation is intermittent, the operation of the transmitter cannot be guaranteed for long periods and the host organizaton should consider the installation of a power generator, which contributes to the economical cost and environmental footprint of the device,

An alternative consists in the utilization of passive devices, like Near Field trasmitters. The device is flat with a surface ranging from a credit card to a coin. The receiver must be very close to enable the device to transmits its content. The power needed to operate the radio is drained from the smartphone, so batteries are not needed. The capacity is in the range of the KB, one page on this journal.

The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application. Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is trasferred to the smartphone as a chink of text, and can be treated as such. The device can read it aloud to meet the user's inabilities or a translator. For both functions off-line apps exist to compensate the lack of network connection.

Another solution in the family of passive devices is the QR-code. Such technique does not require radio communication, but uses the smartphone camera to acquire a graphical code that can be translated into a text. The capacity of a QR-code depends on the number of dots in the code, which in turn depends on the size of the code and the smartphone camera. We may consider that the capacity of a QR-code roughly equals that of an NFC tag with comparable size but a simpler manufacturing.

Such a minimalistic solution compares surprisingly well when compared with other solutions that make use of the user's smartphone. There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words are sufficient to convey the description of a site, or provide directions for the the visit. In exchange, the QR-code solution does not require any power supply, is extremely easy to use, has a limited impact on the laandscape, is not exposed to theft, has a negligible cost, is durable and produces a limited quantity of garbage when disposed, is readable at a moderate distance and can be read aloud, translated or stored by the user's device. It significantly outperforms any active device, except for the capacity, and the NFC compares favorably only when capacity and size are prominent.

The adoption of QR-codes compares favorably with traditional boards, considering the strengths and limits as listed above.

\begin{itemize}
\item dimensions: a 2-300 characters QR-code has the dimensione of a poker card
\item logistics: QR-code board can be installed on any sort of preexisting support
\item impact: the board has a minimal interference with the landscape and may easely go un-noticed if not properly advertised
\item accessibility for the visually impaired: the text can be read aloud
\item international: the text can be translated
\item update: the board can be easely replaced when the content turns obsolete
\item disposal: the card releases a limited quantity of pollutants related with ink support (paper or plastic)
\end{itemize}

One of the issues the host must resolve is the availability of support for electronic devices. The provisioon of such support has a footprint that ha

may be present, but, in the spirit of our investigation, power supply should be address renewable sources, like wind, sunlight and similar.

One option discriminates the use of an electronic device on the side of the user.

In case we assume the user does not have a electronic device with her, the 
One option is to introduce active electronics: the touristic site administration provides a

%Materials and Methods should be described with sufficient details to allow others to replicate and build on published results. Please note that publication of your manuscript implicates that you must make all materials, data, computer code, and protocols associated with the publication available to readers. Please disclose at the submission stage any restrictions on the availability of materials or information. New methods and protocols should be described in detail while well-established methods can be briefly described and appropriately cited.

%Research manuscripts reporting large datasets that are deposited in a publicly avail-able database should specify where the data have been deposited and provide the relevant accession numbers. If the accession numbers have not yet been obtained at the time of submission, please state that they will be provided during review. They must be provided prior to publication.

%Interventionary studies involving animals or humans, and other studies require ethical approval must list the authority that provided approval and the corresponding ethical approval code.
%\begin{quote}
%This is an example of a quote.
%\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

This section may be divided by subheadings. It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
\subsection{Subsection}
\subsubsection{Subsubsection}

Bulleted lists look like this:
\begin{itemize}
\item	First bullet;
\item	Second bullet;
\item	Third bullet.
\end{itemize}

Numbered lists can be added as follows:
\begin{enumerate}
\item	First item; 
\item	Second item;
\item	Third item.
\end{enumerate}

The text continues here.

\subsection{Figures, Tables and Schemes}

All figures and tables should be cited in the main text as Figure~\ref{fig1}, Table~\ref{tab1}, etc.

\begin{figure}[H]
\includegraphics[width=10.5 cm]{Definitions/logo-mdpi}
\caption{This is a figure. Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.\label{fig1}}
\end{figure}   
\unskip

\begin{table}[H] 
\caption{This is a table caption. Tables should be placed in the main text near to the first time they are~cited.\label{tab1}}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CCC}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
Entry 1		& Data			& Data\\
Entry 2		& Data			& Data \textsuperscript{1}\\
\bottomrule
\end{tabularx}
\noindent{\footnotesize{\textsuperscript{1} Tables may have a footer.}}
\end{table}

The text continues here (Figure~\ref{fig2} and Table~\ref{tab2}).

% Example of a figure that spans the whole page width. The same concept works for tables, too.
\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=15.5cm]{Definitions/logo-mdpi}
\end{adjustwidth}
\caption{This is a wide figure.\label{fig2}}
\end{figure}

\begin{table}[H]
\caption{This is a wide table.\label{tab2}}
	\begin{adjustwidth}{-\extralength}{0cm}
		\newcolumntype{C}{>{\centering\arraybackslash}X}
		\begin{tabularx}{\fulllength}{CCCC}
			\toprule
			\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}     & \textbf{Title 4}\\
			\midrule
\multirow[m]{3}{*}{Entry 1 *}	& Data			& Data			& Data\\
			  	                   & Data			& Data			& Data\\
			             	      & Data			& Data			& Data\\
                   \midrule
\multirow[m]{3}{*}{Entry 2}    & Data			& Data			& Data\\
			  	                  & Data			& Data			& Data\\
			             	     & Data			& Data			& Data\\
                   \midrule
\multirow[m]{3}{*}{Entry 3}    & Data			& Data			& Data\\
			  	                 & Data			& Data			& Data\\
			             	    & Data			& Data			& Data\\
                  \midrule
\multirow[m]{3}{*}{Entry 4}   & Data			& Data			& Data\\
			  	                 & Data			& Data			& Data\\
			             	    & Data			& Data			& Data\\
			\bottomrule
		\end{tabularx}
	\end{adjustwidth}
	\noindent{\footnotesize{* Tables may have a footer.}}
\end{table}

%\begin{listing}[H]
%\caption{Title of the listing}
%\rule{\columnwidth}{1pt}
%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
%\rule{\columnwidth}{1pt}
%\end{listing}

Text.

Text.

\subsection{Formatting of Mathematical Components}

This is the example 1 of equation:
\begin{linenomath}
\begin{equation}
a = 1,
\end{equation}
\end{linenomath}
the text following an equation need not be a new paragraph. Please punctuate equations as regular text.
%% If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments). This ensures correct linenumbering. The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.

This is the example 2 of equation:
\begin{adjustwidth}{-\extralength}{0cm}
\begin{equation}
a = b + c + d + e + f + g + h + i + j + k + l + m + n + o + p + q + r + s + t + u + v + w + x + y + z
\end{equation}
\end{adjustwidth}

% Example of a page in landscape format (with table and table footnote).
%\startlandscape
%\begin{table}[H] %% Table in wide page
%\caption{This is a very wide table.\label{tab3}}
%	\begin{tabularx}{\textwidth}{CCCC}
%		\toprule
%		\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}	& \textbf{Title 4}\\
%		\midrule
%		Entry 1		& Data			& Data			& This cell has some longer content that runs over two lines.\\
%		Entry 2		& Data			& Data			& Data\textsuperscript{1}\\
%		\bottomrule
%	\end{tabularx}
%	\begin{adjustwidth}{+\extralength}{0cm}
%		\noindent\footnotesize{\textsuperscript{1} This is a table footnote.}
%	\end{adjustwidth}
%\end{table}
%\finishlandscape

Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows:
%% Example of a theorem:
\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

%% Example of a proof:
\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase ``of Theorem 1'' is optional if it is clear which theorem is being referred to.
\end{proof}
The text continues here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

Authors should discuss the results and how they can be interpreted from the perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Patents}

This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\supplementary{The following supporting information can be downloaded at:  \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.}

% Only for the journal Methods and Protocols:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title. A supporting video article is available at doi: link.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{For research articles with several authors, a short paragraph specifying their individual contributions must be provided. The following statements should be used ``Conceptualization, X.X. and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y. and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing---original draft preparation, X.X.; writing---review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y. All authors have read and agreed to the published version of the manuscript.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation. Authorship must be limited to those who have contributed substantially to the work~reported.}

\funding{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''. Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}

\institutionalreview{In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study. You might choose to exclude this statement if the study did not require ethical approval. Please note that the Editorial Office might ask you for further information. Please add “The study was conducted in accordance with the Declaration of Helsinki, and approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving humans. OR “The animal study protocol was approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving animals. OR “Ethical review and approval were waived for this study due to REASON (please provide a detailed justification).” OR “Not applicable” for studies not involving humans or animals.}

\informedconsent{Any research article describing a study involving humans should contain this statement. Please add ``Informed consent was obtained from all subjects involved in the study.'' OR ``Patient consent was waived due to REASON (please provide a detailed justification).'' OR ``Not applicable'' for studies not involving humans. You might also choose to exclude this statement if the study did not involve humans.

Written informed consent for publication must be obtained from participating patients who can be identified (including by the patients themselves). Please state ``Written informed consent has been obtained from the patient(s) to publish this paper'' if applicable.}

\dataavailability{We encourage all authors of articles published in MDPI journals to share their research data. In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Where no new data were created, or where data is unavailable due to privacy or ethical re-strictions, a statement is still required. Suggested Data Availability Statements are available in section “MDPI Research Data Policies” at \url{https://www.mdpi.com/ethics}.}

\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections. This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}

\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflict of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results. Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section. If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the~results''.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional
\sampleavailability{Samples of the compounds ... are available from the authors.}

%% Only for journal Encyclopedia
%\entrylink{The Link to this entry published on the encyclopedia platform.}

\abbreviations{Abbreviations}{
The following abbreviations are used in this manuscript:\\

\noindent 
\begin{tabular}{@{}ll}
MDPI & Multidisciplinary Digital Publishing Institute\\
DOAJ & Directory of open access journals\\
TLA & Three letter acronym\\
LD & Linear dichroism
\end{tabular}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional
\appendixtitles{no} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendixstart
\appendix
\section[\appendixname~\thesection]{}
\subsection[\appendixname~\thesubsection]{}
The appendix is an optional section that can contain details and data supplemental to the main text---for example, explanations of experimental details that would disrupt the flow of the main text but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data are shown in the main text can be added here if brief, or as Supplementary Data. Mathematical proofs of results not central to the paper can be added as an appendix.

\begin{table}[H] 
\caption{This is a table caption.\label{tab5}}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CCC}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
Entry 1		& Data			& Data\\
Entry 2		& Data			& Data\\
\bottomrule
\end{tabularx}
\end{table}

\section[\appendixname~\thesection]{}
All appendix sections must be cited in the main text. In the appendices, Figures, Tables, etc. should be labeled, starting with ``A''---e.g., Figure A1, Figure A2, etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{adjustwidth}{-\extralength}{0cm}
%\printendnotes[custom] % Un-comment to print a list of endnotes

\reftitle{References}

% Please provide either the correct journal abbreviation (e.g. according to the “List of Title Word Abbreviations” http://www.issn.org/services/online-services/access-to-the-ltwa/) or the full name of the journal.
% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.

%=====================================
% References, variant A: external bibliography
%=====================================
%\bibliography{your_external_BibTeX_file}

%=====================================
% References, variant B: internal bibliography
%=====================================
\begin{thebibliography}{999}
% Reference 1
\bibitem[Author1(year)]{ref-journal}
Author~1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
% Reference 2
\bibitem[Author2(year)]{ref-book1}
Author~2, L. The title of the cited contribution. In {\em The Book Title}; Editor 1, F., Editor 2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
% Reference 3
\bibitem[Author3(year)]{ref-book2}
Author 1, A.; Author 2, B. \textit{Book Title}, 3rd ed.; Publisher: Publisher Location, Country, 2008; pp. 154--196.
% Reference 4
\bibitem[Author4(year)]{ref-unpublish}
Author 1, A.B.; Author 2, C. Title of Unpublished Work. \textit{Abbreviated Journal Name} year, \textit{phrase indicating stage of publication (submitted; accepted; in press)}.
% Reference 5
\bibitem[Author5(year)]{ref-communication}
Author 1, A.B. (University, City, State, Country); Author 2, C. (Institute, City, State, Country). Personal communication, 2012.
% Reference 6
\bibitem[Author6(year)]{ref-proceeding}
Author 1, A.B.; Author 2, C.D.; Author 3, E.F. Title of presentation. In Proceedings of the Name of the Conference, Location of Conference, Country, Date of Conference (Day Month Year); Abstract Number (optional), Pagination (optional).
% Reference 7
\bibitem[Author7(year)]{ref-thesis}
Author 1, A.B. Title of Thesis. Level of Thesis, Degree-Granting University, Location of University, Date of Completion.
% Reference 8
\bibitem[Author8(year)]{ref-url}
Title of Site. Available online: URL (accessed on Day Month Year).
\end{thebibliography}

% If authors have biography, please use the format below
%\section*{Short Biography of Authors}
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author1.pdf}}}
%{\textbf{Firstname Lastname} Biography of first author}
%
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author2.jpg}}}
%{\textbf{Firstname Lastname} Biography of second author}

% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% for journal Sci
%\reviewreports{\\
%Reviewer 1 comments and authors’ response\\
%Reviewer 2 comments and authors’ response\\
%Reviewer 3 comments and authors’ response
%}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\PublishersNote{}
\end{adjustwidth}

\end{document}

### 2023-06-29 paper.tex
Fatta prima versione bibliografia

@@
 %=================================================================
-\documentclass[journal,article,submit,pdftex,moreauthors]{Definitions/mdpi}
+\documentclass[sustainability,article,submit,pdftex,moreauthors]{Definitions/mdpi}
 
@@
 
-Although tourism persistently targets most renowned sites, like art cities and parks, there is an increasing interest for targets not exposed to mass tourism, that may provide a more intense and inner experience.
-Suh an exclusive flavor is appreciated not only by the occasional visitor, which helps the economy of the corner, but also by those that have their homes in such privileged location.
-The key to sustain such a difficult equilibrium is to guide the visitor to enjoy the place with discrete messages that he finds on his way.
+Although an official decision is still pending, there is a sound proposal of naming the time we are living as "anthropocene".
+Behind such a proposal is the recognition that humans have the potential to significantly alter the environment they live in at a global scale.
 
-%The introduction should briefly place the study in a broad context and highlight why it is important.
-It should define the purpose of the work and its significance.
-The current state of the research field should be reviewed carefully and key publications cited.
-Please highlight controversial and diverging hypotheses when necessary.
-Finally, briefly mention the main aim of the work and highlight the principal conclusions.
-As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research.
-Citing a journal paper \cite{ref-journal}.
-Now citing a book reference \cite{ref-book1,ref-book2} or other reference types \cite{ref-unpublish,ref-communication,ref-proceeding}.
-Please use the command \citep{ref-thesis,ref-url} for the following MDPI journals, which use author--date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, Journal of Intelligence, Journalism and Media, JRFM, Languages, Laws, Religions, Risks, Social Sciences, Literature.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The problem behind the above fact is that we humans are not able to understand the consequences of the deployment of such a potential, and not even to keep it under control.
+Even worse, such capabilities are often used aiming at short-term local effects, ignoring long-term global ones, with the risk of a long-term global deterioration.
+Such an attitude pervades all activities, starting from the most basic ones like the provision of food and homes, and is frequenly biased by economical profit.
+
+Leisure activities play a relevant role, since they bring profit to niche businesses.
+This creates a reinforcement effect which follows the negative dynamic seen above: a leisure activity that per se is neutral (e.g.
+skying) may have a relevant footprint due to induced effects (ski lifts) which reinforces the success of the activity (easier to practice) bringing more investments and impact.
+Unlike beneficial effects, which is usually well localized, the negative impact has a very wide spectrum encompassing environmental, social and finally economical domains, as discussed by Patthey in \cite{pat08a}.
+
+In an effort to manage such a situation, international agencies promote the "sustainability" keyword as a guideline to the conservation of our habitat.
+Jeffry Ramsey et al.
+\cite{ram15a} consider the concept behind the word is "vague and contested but not meaningless" and that it cannot be used without a concrete framework, especially in a normative context.
+To solve to problem of measuring sustainability, Tom Kuhlman et al.
+point out a inherent tension in the term between growth and stability.
+A tension that has been softed by a shift in meaning of the word, which moved from focusing on preservation for an indefinite future to a concept encompassing three "pillars", society, economy and environment, as determinants that need to be protected in the present in view of a future sustainable development.
+The authors put into relationship man-made and environmental assets, discriminating the two cases in which a loss in the latter may or may not be compensated by an increment of the former, introducinging the terms of {\em weak} and {\em strong} sustainability.
+
+The promotion of the human cultural heritage has its stand in such a complex scenario.
+As noted by Jorge Otero in \cite{ote22a}, local communities take advantage of heritage promotion: the preservation activity generates social growth, while tourism provides an extra income that may be re-invested in promotion.
+However, just like in the sky resorts case mentioned above, the risk of compromising the resource itself exists.
+Such an interplay rises a question of sustainability even for cultural preservation initiatives, two topics that are usually regarded as in synergy.
+As in the "World Heritage and Sustainable Development" document by Unesco \cite{lab17a} that calls for actions that "harness the potential of World Heritage properties and heritage in general, to contribute to sustainable development".
+
+This paper originates from such a painful contradiction trying to give an answer to the question: is there a way to unlock the potential of a cultural resource limiting the damage to a social, economic and environmental equilibrium?
+
+The question arises in the course of a project funded by the italian ministry of reasearch aiming at the application of cutting edge technologies to the investigation of karst caves that may have been inhabited in the past, from prehistory to today.
+Among the results expected from the project there is the definition of a strategy to disseminate research results, and allow non-specialists to visit and enjoy the findings, thus contributing to the economy of rural areas that are severely exposed to de-population.
+A sustainable strategy is even more crucial given that the geological enviroment of interest is notably fragile, as noted by Aleksander Antic\`c et al.
+in \cite{ant20a}.
+
+This paper provides a concept solution of a specific aspect.
+The solution is conceptual in that it tries to follow a guideline that takes into account as many side details as possible in a way that can be used in other milieu, but very practical in its unfolding.
+Given the tighly interconnected nature of the situation, we observe how a vary specific aspect, the touristic signage, exhibits effects on the society, economicy and environment.
+
+\section{Unlocking site potentials through signage}
+
+The signage used to guide and inform the visitor is of paramount importance and must be carefully designed in order to accomplish its functions while following the guidelines of a sustainable approach.
+
+Its functions are twofold
+\begin {itemize}
+\item an effective signage must guide the tourist across the resort.
+A map represents a starting point to this end, but not all visitors feel comfortable reading a map.
+The ideal signage should include visible reference points and explicit actions, like {\em turn right after crossing the stream}.
+A good example of such a signage is the one used by the Italian Alpine Club (CAI): it consist of a colored stripes marked on tree bark.
+The marks are placed in such a way that, standing nearby one of them, the hiker is able to see the successive on the trail
+\item a sign should explain the reasons of interest of a site.
+This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit.
+In this sense, the CAI signal above is not of use.
+A simple board with a site name may be sufficient only when the site is sufficiently reknown.
+Otherwise, a more structured message should be used to explain the reasons why the feature is regarded as relevant.
+\end{itemize}
+
+In our study we consider a range of solutions that help towards a sustainable support to a sustainable tourism, and specifically to the dynamic provision of information during a visit along a natural trail, not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}).
+We consider the presence of two stakeholders for the signage itaself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.
+
+The purpose of such an installation is guiding the {\em user} in a tour that includes urban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research).
+The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.
+
+A traditional approach makes use of physical information boards with graphical or textual contents. Our study starts from finding the issues related with such a solution:
+
+\begin{itemize}
+	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
+	\item installation logistics: depending on the location, the transportation and placement of a plaque may require a basement or other sorts of supports
+	\item environmental impact: in order to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site 
+	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
+	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
+	\item update limits: to update the content, the plaque must be replaced
+	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues similar to those found during the installation
+\end{itemize}
+
+Such considerations motivate an interest for an alternative way of communication.
+
+\subsection{Technology to minimize intrusion}
+
+We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention on the habitat, but the impact of such intervention must be better than that related with traditional signs.
+To this end we include in our solution tools that are not part of the site but remain with the user.
+
+During the last decades, we witnessed the diffusion of smartphone devices, and we have no reasons to expect a change in such a trend.
+Smartphones empower individual communication capabilities, allowing to receive sound, visual and tactile interactions.
+The relevance of such capabilities for the improvement of a tourist experience have been widely investigated, etiher in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a}.
+to investigate ways to exploit such tools in our context.
+Also in this case we start from the limits of such technology, especially in the case of outdoor activities.
+
+One is related with the diffusion of such a technology, which, although widely available, is not equally familiar to everybody.
+This is related with {\em usability}, with a term borrowed from P.
+Wan in \cite{wan22a}.
+The second is that several functions depends on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach an enabling infrastructure.
+So the {\em applicability limits} of a solution need to be defined.
+
+Starting from the two aspects above we envision the directions towards a successful smartphone-based strategy, keeping in mind the overall sustainability of the approach.
+
+Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of installing new applications.
+
+The definition of its {\em applicability limits} is more complex, especially for outdoor activities.
+In such cases, the provision of a networking infrastructure incurs a severe environmental impact.
+Consider the installation of antennas to cover a wide area, and the power supply for the radios.
+
+Circular economy
+
+\subsection{Related works}
+
+The research literature marginally covers the utilization of smartphones for tourist signage purposes.
+
+An exhaustive solution is described by P.
+Liu in \cite{liu16a}, which provide details for an infrastructure that guides the visitors inside an urban milieu.
+In that case the presence of a pervasive networking facilities is a cornerstone for the whole architecture, that deeply depends on communication.
+
+Wan \cite{wan22a} evaluates the quality signage, without making reference to a specific technology, but with many example showing phisical boards, using a formal reference, the Universal Design Principles \cite{udi97a}.
+
+The number of research papers explodes when we extend the range to articles that investigate the impact of smartphones in tourism.
+The topic {\em smart tourism} topic is very popular, and covered by several review papers that provide a framework to a vast literature.
+
+A relevant research direction covers the social aspects of related with the use of the smartphone.
+W.
+Tan \cite{tan17a} covers all aspects of a touristic experience related with the smartphone from the definition of travel destination to assistance during the visit.
+Much attention is dedicated to the network of connections that is established thanks to the smartphone, which again is considered in the Internet despite the title indicates a nature-based destination, where notably the Internet is not always reachable.
+
+On the other hand, roaming in places not reached by the Internet, or {\em dead zones} using the words of Pearce and Gretzel in \cite{pea12a}, may evoke constrasting feelings, from rewarding to threatening.
+
+More recently, the smartphone has been considered not strictly related with communication.
+In 2021 A.Slavec et al.
+investigate the use of cameras \cite{sla21a} while on travel in locations with a relevant cultural heritage in order to sustain its preservation and engage the tourist using location based games, similar to Pokemon Go or geo-caching.
+
+In 2023 V.
+Rodrigues et al.
+publish a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}.
+Although the title evokes a one-way contribution, i.e., the impact of digitalization is assumed to be positive on the quality and the sustainability of touristic offers, in the conclusions the authors reveal awareness for the necessity to address {\em "the preservation of tourism attractions/sites"} and call for for a {\em "a holistic approach ...
+to support a concept that still lacks conceptual and empirical clarification"}.
+
+In that direction we meet the phenomenon of {\em overtourism}, covered by a significant literature reviewed by Dodds and Butler \cite{dod19a}, which focusses on urban tourism and its social consequences.
+The impact of {\em overtourism} on resorts that trade on their natural resources is investigated in the case of the Hawaii Islands \cite{lin22a} or CostaRica \cite{mat10a} stressing the imact on the social fabric.
+
+The present paper wants to fill this gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined aspect of tourism support taking into account sustainability by design.
+Once a range of relevant solutions has been identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies feasibility of a specific solution.
+
+We stress that this article does not aim now at quantifying user satisfaction or the economic revenue associated with the specific solution.
+Such a target is outside the scope of our research, and indeed the figures that would represent the success of a strategy deserve further investigation.
+We aim at isolating an issue, propose a strategy for its solution and implement a proof of concept for it.
+
+% Other papers examine the topic from the point of view of the user satisfaction, with an accent on the accessibility of the contents from impaired users.
+
+% The use of smartphone is support of a tourist experience is mostly regarded from the point of view of the tourist satisfaction.
+
+% This is the case of
+
+% Cannot depend on communication
+
+\subsection{A simple, low impact solution}
+
+%Starting from previous experiences documented in the bibliography we explore a few broad categories, to find weaknesses and strengths for each of them.
+
+We aim at designing a strategy for a sustainable and effective smartphone-based signage.
+
+A straightforward approach concentrates on the wireless capabilities of such a device.
+Given the premise that the Internet is not reachable, the host might provide a local network of low power radios covering the region of interest.
+Small servers connected to the network might provide specific Web services.
+The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single chip computer (SCC) has a coverage fo tens of meters, and a volume in the order of the $cm^3$.
+They are able to provide WiFi Access Points as well as Web contents.
+The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image.
+Other SCCs, like the ESP32, are more powerful but exhibit a higher power consumption.
+
+The dependency from power supply is a serious limitation of such kind of solution.
+A radio transmitter is a rather power consuming device, in the range of Watts.
+Even if operation is intermittent, the operation of the transmitter cannot be guaranteed for long periods and the host organization should consider the installation of a power generator, which contributes to the economical cost and environmental footprint of the device.
+
+For this reason we do not consider a solution based on an ad-hoc network as a valid competitor against traditional board-based design for reasons related with poor sustainability.
+
+An alternative consists in the utilization of passive devices, like Near Field trasmitters.
+The device is flat with a surface ranging from a credit card to a coin.
+The receiver must be very close to enable the device to transmits its content.
+The power needed to operate the radio is drained from the smartphone, so batteries are not needed.
+The capacity is in the range of the KB, one page on this journal.
+
+The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application.
+Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the smartphone as a chink of text, and can be treated as such.
+The device can read it aloud to meet the user's inabilities or a translator.
+For both functions off-line apps exist.
+
+Another solution in the family of passive devices is the QR-code.
+Such technique does not require radio communication, but uses the smartphone camera to acquire a graphical code that is translated into text.
+The capacity of a QR-code depends on the number of dots in the code, which in turn depends on the size of the code and the smartphone camera characteristics.
+We may consider that the capacity of a QR-code roughly equals that of an NFC tag with comparable size but a simpler manufacturing.
+
+% TODO: \usepackage{graphicx} required
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\linewidth]{figure/qr+NFC+coin}
+	\caption[Passive devices dimensions]{An NFC coin, a QR-code card used in the project, and a 1 Euro coin by comparison on a paper with a square of 1cm}
+	\label{fig:qrnfccoin}
+\end{figure}
+
+We first check the above strategies against the Universal Design Principles \cite{udi97a}:
+
+\begin{enumerate}
+	\item Equitable Use
+	\item Flexibility in Use
+	\item Simple and Intuitive
+	\item Perceptible Information
+	\item Tolerance for Error
+	\item Low Physical Effort
+	\item Size and Space for Approach and Use
+\end{enumerate}
+
+The labels are self-descriptive, and we address the reader to the discussion in the referenced document for a more detailed explanation.
+
+Such minimalistic solutions (se figure \ref{fig:qrnfccoin}) compares surprisingly well when compared with other solutions that make use of the user's smartphone.
+There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words is sufficient to convey the description of the site, or provide directions for the the visit.
+In exchange, a solution besed on NFC or QR-codes:
+\begin{itemize} 
+\item does not require any power supply
+\item is easy to use
+\item has a limited impact on the landscape
+\item is not exposed to theft
+\item has a negligible cost
+\item is durable
+\item produces a limited quantity of waste when disposed
+\item is readable at a moderate distance, a feature that differentiates NFP and QR-codes
+\item can be read aloud or translated
+\item can be stored by the user's device.
+\end{itemize}
+
+Checking passive devices solutions against the Universal Design principles, we assess their accessibility.
+
+\begin{itemize}
+	\item {\em Equitable use} is tighly related to the smartphone technology, which is itself considered a vehicle for equitability,
+	\item {\em Flexibility in use} is allowed by the computing capabilities of the device, which allows to listen to the info instead or reading, or translate the information in a different language,
+	\item {\em Simple and intuitive use} holds since the operation requires an application, possibly already installed since useful in many occasions, and tag reading requires
+	\item{\em Perceptible information} is a critical requirement, which contrast with the aim of keeping low the environmental intrusion.
+This point will be further discussed in the section devoted to the implementation
+	\item{\em Tolerance for error}: there are no margins to use the device in a way that comprimizes user safety.
+The deliberate or accidental release of the passive device into the environment determines a minor pollution
+	\item{\em Low physical effort} holds, although the user needs to carry the smartphone
+	\item{Size and Space for approach and use} have to be carefully considered, since in the case of the NFC the smartphone need to be almost in touch with the passive device, while the QR-code must be in line of sight and frameable without effort
+\end{itemize}
+
+It significantly outperforms any active device, except for the capacity, and the NFC compares favorably only when capacity and size are prominent.
+
+The adoption of QR-codes compares favorably with traditional boards, considering the strengths and limits as listed above.
+
+\begin{itemize}
+	\item dimensions: a 2-300 characters QR-code has the dimensione of a poker card
+	\item logistics: QR-code board can be installed on any sort of preexisting support
+	\item impact: the board has a minimal interference with the landscape and may easely go un-noticed if not properly advertised
+	\item accessibility for the visually impaired: the text can be read aloud
+	\item international: the text can be translated
+	\item update: the board can be easely replaced when the content turns obsolete
+	\item disposal: the card releases a limited quantity of pollutants related with ink support (paper or plastic)
+\end{itemize}
+
+		One of the issues the host must resolve is the availability of support for electronic devices. The provisioon of such support has a footprint that ha
+
+		may be present, but, in the spirit of our investigation, power supply should be address renewable sources, like wind, sunlight and similar.
+
+		One option discriminates the use of an electronic device on the side of the user.
+
+		In case we assume the user does not have a electronic device with her, the 
+		One option is to introduce active electronics: the touristic site administration provides a
+
 \section{Materials and Methods}
@@
 
-%=====================================
-% References, variant A: external bibliography
-%=====================================
-%\bibliography{your_external_BibTeX_file}
-
-%=====================================
-% References, variant B: internal bibliography
-%=====================================
-\begin{thebibliography}{999}
-% Reference 1
-\bibitem[Author1(year)]{ref-journal}
-Author~1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
-% Reference 2
-\bibitem[Author2(year)]{ref-book1}
-Author~2, L. The title of the cited contribution. In {\em The Book Title}; Editor 1, F., Editor 2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
-% Reference 3
-\bibitem[Author3(year)]{ref-book2}
-Author 1, A.; Author 2, B. \textit{Book Title}, 3rd ed.; Publisher: Publisher Location, Country, 2008; pp. 154--196.
-% Reference 4
-\bibitem[Author4(year)]{ref-unpublish}
-Author 1, A.B.; Author 2, C. Title of Unpublished Work. \textit{Abbreviated Journal Name} year, \textit{phrase indicating stage of publication (submitted; accepted; in press)}.
-% Reference 5
-\bibitem[Author5(year)]{ref-communication}
-Author 1, A.B. (University, City, State, Country); Author 2, C. (Institute, City, State, Country). Personal communication, 2012.
-% Reference 6
-\bibitem[Author6(year)]{ref-proceeding}
-Author 1, A.B.; Author 2, C.D.; Author 3, E.F.
-Title of presentation.
-In Proceedings of the Name of the Conference, Location of Conference, Country, Date of Conference (Day Month Year); Abstract Number (optional), Pagination (optional).
-% Reference 7
-\bibitem[Author7(year)]{ref-thesis}
-Author 1, A.B. Title of Thesis. Level of Thesis, Degree-Granting University, Location of University, Date of Completion.
-% Reference 8
-\bibitem[Author8(year)]{ref-url}
-Title of Site. Available online: URL (accessed on Day Month Year).
-\end{thebibliography}
+\bibliography{biblio/paper.bib}
 

### 2023-07-17 paper.tex
Avanzamento

@@
 
-We first check the above strategies against the Universal Design Principles \cite{udi97a}:
+A preliminary check verifies compliance with the Universal Design Principles \cite{udi97a}:
 
-\begin{enumerate}
-	\item Equitable Use
-	\item Flexibility in Use
-	\item Simple and Intuitive
-	\item Perceptible Information
-	\item Tolerance for Error
-	\item Low Physical Effort
-	\item Size and Space for Approach and Use
-\end{enumerate}
+%\begin{enumerate}
+%	\item Equitable Use
+%	\item Flexibility in Use
+%	\item Simple and Intuitive
+%	\item Perceptible Information
+%	\item Tolerance for Error
+%	\item Low Physical Effort
+%	\item Size and Space for Approach and Use
+%\end{enumerate}
 
-The labels are self-descriptive, and we address the reader to the discussion in the referenced document for a more detailed explanation.
+%The labels are self-descriptive, and we address the reader to the discussion in the referenced document for a more detailed explanation.
+%Checking passive devices solutions against the Universal Design principles, we assess their accessibility.
 
-Such minimalistic solutions (se figure \ref{fig:qrnfccoin}) compares surprisingly well when compared with other solutions that make use of the user's smartphone.
+\begin{itemize}
+	\item {\em Equitable use} is tighly related to the smartphone technology, which is itself considered a vehicle for equitability,
+	\item {\em Flexibility in use} is enabled by the computing capabilities of the device, which allows to listen to the info instead or reading, or translate the information in a different language,
+	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single touch
+	\item{\em Perceptible information} is a critical requirement, which contrasts with the aim of keeping low the environmental intrusion.
+This point will be further discussed in the section devoted to the implementation
+	\item{\em Tolerance for error}: there are no margins to use the device in a way that compromises user safety.
+The deliberate or accidental release of the passive device into the environment determines a minor pollution
+	\item{\em Low physical effort} holds, although the user needs to carry the smartphone
+	\item{Size and Space for approach and use} have to be carefully considered.
+In the case of the NFC the smartphone needs to be nearly in touch with the passive device, while the QR-code must be in line of sight and frameable without effort
+\end{itemize}
+
+Such a minimalistic solutions (se figure \ref{fig:qrnfccoin}) compares well compared with other solutions that make use of the user's smartphone.
 There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words is sufficient to convey the description of the site, or provide directions for the the visit.
-In exchange, a solution besed on NFC or QR-codes:
+If capacity is not an issue a solution based on NFC or QR-codes exhibits a number of advantages:
 \begin{itemize} 
 \item does not require any power supply
-\item is easy to use
 \item has a limited impact on the landscape
@@
 \item produces a limited quantity of waste when disposed
-\item is readable at a moderate distance, a feature that differentiates NFP and QR-codes
-\item can be read aloud or translated
-\item can be stored by the user's device.
+\item the content can be stored by the user's device for successive utilization
 \end{itemize}
 
-Checking passive devices solutions against the Universal Design principles, we assess their accessibility.
+%It significantly outperforms any active device, except for the capacity, and the NFC compares favorably only when capacity and size are prominent.
+
+A solution based on the user smartphone associated with passive devices compares favorably with traditional signage boards.
+Considering the limits of the latter approach listed in table \ref{tab:board}:
+
+%The adoption of passive devices compares favorably with traditional boards, considering the strengths and limits as listed above.
 
 \begin{itemize}
-	\item {\em Equitable use} is tighly related to the smartphone technology, which is itself considered a vehicle for equitability,
-	\item {\em Flexibility in use} is allowed by the computing capabilities of the device, which allows to listen to the info instead or reading, or translate the information in a different language,
-	\item {\em Simple and intuitive use} holds since the operation requires an application, possibly already installed since useful in many occasions, and tag reading requires
-	\item{\em Perceptible information} is a critical requirement, which contrast with the aim of keeping low the environmental intrusion.
-This point will be further discussed in the section devoted to the implementation
-	\item{\em Tolerance for error}: there are no margins to use the device in a way that comprimizes user safety.
-The deliberate or accidental release of the passive device into the environment determines a minor pollution
-	\item{\em Low physical effort} holds, although the user needs to carry the smartphone
-	\item{Size and Space for approach and use} have to be carefully considered, since in the case of the NFC the smartphone need to be almost in touch with the passive device, while the QR-code must be in line of sight and frameable without effort
-\end{itemize}
-
-It significantly outperforms any active device, except for the capacity, and the NFC compares favorably only when capacity and size are prominent.
-
-The adoption of QR-codes compares favorably with traditional boards, considering the strengths and limits as listed above.
-
-\begin{itemize}
-	\item dimensions: a 2-300 characters QR-code has the dimensione of a poker card
-	\item logistics: QR-code board can be installed on any sort of preexisting support
+	\item dimensions: a 2-300 characters QR-code has a dimension in the order of 100 $cm^2$
+	\item logistics: QR-code board can be installed on any sort of pre-existent or natural support
 	\item impact: the board has a minimal interference with the landscape and may easely go un-noticed if not properly advertised
@@
 	\item international: the text can be translated
-	\item update: the board can be easely replaced when the content turns obsolete
+	\item update: the board can be easely replaced when the content becomes obsolete
 	\item disposal: the card releases a limited quantity of pollutants related with ink support (paper or plastic)
@@
 
-		One of the issues the host must resolve is the availability of support for electronic devices. The provisioon of such support has a footprint that ha
+%One of the issues the host must resolve is the availability of support for electronic devices. The provisioon of such support has a footprint that ha		
+%may be present, but, in the spirit of our investigation, power supply should be address renewable sources, like wind, sunlight and similar.
+%One option discriminates the use of an electronic device on the side of the user. 
+%In case we assume the user does not have a electronic device with her, the 
+%One option is to introduce active electronics: the touristic site administration provides a
 
-		may be present, but, in the spirit of our investigation, power supply should be address renewable sources, like wind, sunlight and similar.
-
-		One option discriminates the use of an electronic device on the side of the user.
-
-		In case we assume the user does not have a electronic device with her, the 
-		One option is to introduce active electronics: the touristic site administration provides a
+There are two relevant trade-offs that the host needs to resolve. One is related with the visibility of the tag. The trade-off is among the
 

### 2023-07-20 paper.tex
Prima revisione quasi competa, aggiunte immagini, tabella e mappa tag, eliminato frammento duplicato

@@
 % cleveref: load \crefname definitions after \begin{document}
+\usepackage{float}
+\usepackage[caption = false]{subfig}
 
@@
 
-\section{Introduction}
-
-\subsection{The role of tourist signage}
-
-\section{Tourist signage sustainability}
-
-\subsection{Working hypothesis}
-
-\subsection{Evaluation parameters}
-
-\subsection{Evaluation of sample alternatives}
-
-\section{Application to a concrete use case}
-
-\subsection{Trail selection}
-
-\subsection{Social and political aspects}
-
-\subsection{Signage design}
-
-\subsection{Monitoring results}
-
-\section{Conclusions}
-
-\setcounter{section}{-1} %% Remove this when starting to work on the template.
-\section{How to Use this Template}
-
-The template details the sections that can be used in a manuscript.
+%\setcounter{section}{-1} %% Remove this when starting to work on the template.
+%\section{How to Use this Template}
+
+%The template details the sections that can be used in a manuscript.
 Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section).
@@
 The solution is conceptual in that it tries to follow a guideline that takes into account as many side details as possible in a way that can be used in other milieu, but very practical in its unfolding.
-Given the tighly interconnected nature of the situation, we observe how a vary specific aspect, the touristic signage, exhibits effects on the society, economicy and environment.
+Given the tighly interconnected nature of the situation, we observe how a very specific aspect, the touristic signage, exhibits effects on the society, economy and environment.
 
@@
 
-Its functions are twofold
+The functions of signage are twofold
 \begin {itemize}
@@
 A map represents a starting point to this end, but not all visitors feel comfortable reading a map.
-The ideal signage should include visible reference points and explicit actions, like {\em turn right after crossing the stream}.
-A good example of such a signage is the one used by the Italian Alpine Club (CAI): it consist of a colored stripes marked on tree bark.
-The marks are placed in such a way that, standing nearby one of them, the hiker is able to see the successive on the trail
+The ideal signage should include visible reference points and explicit actions, like "turn right after crossing the stream".
+A good example of such a signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark.
+The marks are placed in such a way that, standing nearby one of them, the hiker is able to spot the successive on the trail
 \item a sign should explain the reasons of interest of a site.
@@
 
+%In our study we have considered a range of solutions that may help towards our goal of a sustainable support to a sustainable tourism, and specifically to the dynamic provision of information during a visit.
+To this end, the hosting organization (the {\em host} for short) implements a system of devices from which the visitor (or {\em user} ) extracts useful and enjoyable information.
+
 The purpose of such an installation is guiding the {\em user} in a tour that includes urban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research).
 The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.
-
-A traditional approach makes use of physical information boards with graphical or textual contents. Our study starts from finding the issues related with such a solution:
+A traditional approach makes use of physical information boards with graphical or textual contents.
+
+%The purpose of such an installation is to guide the {\em user} in a tour that includes hurban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research).
+The task of the host organization is to provide the user with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.
+
+Our study starts from pointing out the issues related with such a solution:
 
@@
 
-\subsection{Technology to minimize intrusion}
-
-We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention on the habitat, but the impact of such intervention must be better than that related with traditional signs.
+%A traditional approach makes use of physical information boards with graphical or textual contents. Our study starts from finding the issues related with such a solution:
+
+%\begin{itemize}
+%	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
+%	\item installation logistics: depending on the location, the placement of a plaque may require a basement or other sorts of supports
+%	\item environmental impact: in order to be effective, the plaque is prominent, and this may negatively affect the quality of the site 
+%	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
+%	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
+%	\item update limits: to update the content, the plaque must be replaced
+%	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues like those found during the installation
+%\end{itemize}
+
+%Such considerations motivate an interest for an alternative way of communication.
+
+\subsection{Technology to minimize intrusion \label{sec:minimize}}
+
+We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention on the environment, but the impact of such intervention must be better than that related with traditional signs.
 To this end we include in our solution tools that are not part of the site but remain with the user.
@@
 The relevance of such capabilities for the improvement of a tourist experience have been widely investigated, etiher in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a}.
-to investigate ways to exploit such tools in our context.
+to find ways to exploit such tools in our context.
 Also in this case we start from the limits of such technology, especially in the case of outdoor activities.
@@
 
+% The second issue is more complex, especially when considering outdoor activities.
+In such cases the provision of a networking infrastructure may be impossibile without a severe environmental impact.
+Consider the installation of antennas to cover a wide area and power supply for the radios.
+
 Starting from the two aspects above we envision the directions towards a successful smartphone-based strategy, keeping in mind the overall sustainability of the approach.
 
-Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of installing new applications.
+Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of familiarizing with new applications.
 
@@
 
-Circular economy
+% Circular economy
 
@@
 An exhaustive solution is described by P.
-Liu in \cite{liu16a}, which provide details for an infrastructure that guides the visitors inside an urban milieu.
-In that case the presence of a pervasive networking facilities is a cornerstone for the whole architecture, that deeply depends on communication.
-
-Wan \cite{wan22a} evaluates the quality signage, without making reference to a specific technology, but with many example showing phisical boards, using a formal reference, the Universal Design Principles \cite{udi97a}.
+Liu in \cite{liu16a}, which details an infrastructure that guides the visitors inside an urban milieu.
+In that case the presence of a pervasive networking facilities is a cornerstone for the whole architecture, that deeply depends on Internet connectivity.
+
+Wan \cite{wan22a} evaluates the quality of signage, without making reference to a specific technology, but with many examples showing physical boards, using as a formal reference the Universal Design Principles \cite{udi97a}.
 
 The number of research papers explodes when we extend the range to articles that investigate the impact of smartphones in tourism.
-The topic {\em smart tourism} topic is very popular, and covered by several review papers that provide a framework to a vast literature.
-
-A relevant research direction covers the social aspects of related with the use of the smartphone.
+The topic {\em smart tourism} topic is very popular, and covered by several review papers that provide a framework to the vast literature.
+
+A popular research direction covers the social aspects related with the use of the smartphone.
 W.
-Tan \cite{tan17a} covers all aspects of a touristic experience related with the smartphone from the definition of travel destination to assistance during the visit.
-Much attention is dedicated to the network of connections that is established thanks to the smartphone, which again is considered in the Internet despite the title indicates a nature-based destination, where notably the Internet is not always reachable.
+Tan \cite{tan17a} covers all aspects of a touristic experience related with the smartphone, from the definition of travel destination to assistance during the visit.
+Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is considered in the Internet, despite the title indicates a nature-based destination, where notably the Internet is not always reachable.
 
@@
 In that direction we meet the phenomenon of {\em overtourism}, covered by a significant literature reviewed by Dodds and Butler \cite{dod19a}, which focusses on urban tourism and its social consequences.
-The impact of {\em overtourism} on resorts that trade on their natural resources is investigated in the case of the Hawaii Islands \cite{lin22a} or CostaRica \cite{mat10a} stressing the imact on the social fabric.
-
-The present paper wants to fill this gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined aspect of tourism support taking into account sustainability by design.
+The impact of {\em overtourism} on resorts that trade on their natural resources is investigated in the case of the Hawaii Islands \cite{lin22a} or CostaRica \cite{mat10a} stressing the impact on the social fabric.
+
+The present paper wants to fill the gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined tourism support aspect, taking into account by design its sustainability.
 Once a range of relevant solutions has been identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies feasibility of a specific solution.
@@
 We stress that this article does not aim now at quantifying user satisfaction or the economic revenue associated with the specific solution.
-Such a target is outside the scope of our research, and indeed the figures that would represent the success of a strategy deserve further investigation.
-We aim at isolating an issue, propose a strategy for its solution and implement a proof of concept for it.
+Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve further investigation.
+We aim at isolating an issue, proposing a strategy for its solution and implementing a proof of concept for it.
 
@@
 A straightforward approach concentrates on the wireless capabilities of such a device.
-Given the premise that the Internet is not reachable, the host might provide a local network of low power radios covering the region of interest.
+Given the premise that the Internet is not reachable, the {\em host} stakeholder may provide a local network of low power radios covering the region of interest.
 Small servers connected to the network might provide specific Web services.
-The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single chip computer (SCC) has a coverage fo tens of meters, and a volume in the order of the $cm^3$.
+The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single chip computer (SCC) has a coverage of tens of meters, and a volume in the order of the $cm^3$.
 They are able to provide WiFi Access Points as well as Web contents.
@@
 
-The dependency from power supply is a serious limitation of such kind of solution.
+% One strategy concentrates on the wireless capabilities of the smartphone, although avoiding the coverage of the whole region.
+We consider the implementation of several small servers operating in isolation that provide a few HTML pages.
+The approach requires a modest investment and a marginal environmental impact: a small device based on the ESP8266 single chip computer has a coverage fo tens of meters, and a volume in the order of the $cm^3$.
+They operate as WiFi Access Points, so the user has access to the data in the device by joining the AP and browsing the URL.
+The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image.
+
+Power supply dependency is a serious limitation for such kind of solution.
 A radio transmitter is a rather power consuming device, in the range of Watts.
@@
 An alternative consists in the utilization of passive devices, like Near Field trasmitters.
-The device is flat with a surface ranging from a credit card to a coin.
-The receiver must be very close to enable the device to transmits its content.
-The power needed to operate the radio is drained from the smartphone, so batteries are not needed.
-The capacity is in the range of the KB, one page on this journal.
+The transmitting device is flat with the size of a coin and a cost of less than one dollar per piece.
+The receiving smartphone must be very close to receive the content.
+The power needed to operate the radio is drained from the smartphone, so that batteries are not needed.
+The NFC device capacity is in the range of the KBytes, nearly a page on this journal.
+Its manufacturing requires a smartphone with a specific app.
 
 The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application.
-Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the smartphone as a chink of text, and can be treated as such.
-The device can read it aloud to meet the user's inabilities or a translator.
-For both functions off-line apps exist.
-
-Another solution in the family of passive devices is the QR-code.
+Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the {\em user's} device as a chink of text which can be treated as such.
+The smartphone can read it aloud to compensate user's inabilities or a translator to cope with linguistic issues without need of an Internet connection.
+
+Another solution in the family of passive devices is the QR-tag.
 Such technique does not require radio communication, but uses the smartphone camera to acquire a graphical code that is translated into text.
-The capacity of a QR-code depends on the number of dots in the code, which in turn depends on the size of the code and the smartphone camera characteristics.
-We may consider that the capacity of a QR-code roughly equals that of an NFC tag with comparable size but a simpler manufacturing.
+The capacity of a QR-tag depends on the number of dots in the image, which in turn depends on the size of the code and the smartphone camera characteristics.
+We may consider that the capacity of a QR-code roughly equals that of an NFC tag.
+A QR code is larger than an NFC tag of comparable capacity, and manufacturing requires a printer.
 
@@
 	\item {\em Equitable use} is tighly related to the smartphone technology, which is itself considered a vehicle for equitability,
-	\item {\em Flexibility in use} is enabled by the computing capabilities of the device, which allows to listen to the info instead or reading, or translate the information in a different language,
+	\item {\em Flexibility in use} is enabled by the computing capabilities of the device, which allows listening to the info instead or reading, translating the information in a different language, or storage for later use,
 	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single touch
@@
 
-Such a minimalistic solutions (se figure \ref{fig:qrnfccoin}) compares well compared with other solutions that make use of the user's smartphone.
+Such minimalistic solutions (see figure \ref{fig:qrnfccoin}) compares well with other solutions that make use of the user's smartphone.
 There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words is sufficient to convey the description of the site, or provide directions for the the visit.
-If capacity is not an issue a solution based on NFC or QR-codes exhibits a number of advantages:
+If capacity limits are not an issue, a solution based on NFC or QR-codes exhibits a number of advantages:
 \begin{itemize} 
@@
 \item has a limited impact on the landscape
-\item is not exposed to theft
+\item does not entice theft
 \item has a negligible cost
@@
 \item produces a limited quantity of waste when disposed
-\item the content can be stored by the user's device for successive utilization
+\item content can be stored for usage when the user reaches a zone covered by the Internet
 \end{itemize}
 
+The two passive technologies of choice exhibit the following features, that may make one of them preferable for a specific application:
+
+\begin{itemize}
+	\item an NFC-tag is smaller than a QR-tag
+	\item writing an NFC-tag requires a smartphone, while the QR-tag needs a printer
+	\item reading an NFC-tag operates at near-to-contact distances, while QR-codes can be read from meters away
+\end{itemize}
+
 %It significantly outperforms any active device, except for the capacity, and the NFC compares favorably only when capacity and size are prominent.
 
-A solution based on the user smartphone associated with passive devices compares favorably with traditional signage boards.
-Considering the limits of the latter approach listed in table \ref{tab:board}:
+For a signage application the QR-tag is preferable because a noticeable size is needed. In addition, keeping the tag out of reach prevents vandalisms and misuse.
+
+In conclusion, we have reasons to select a QR-code based solution as a good candidate for a smartphone-based signage.
+We now consider how it it copes with the limits of a traditiona board-based approach (as listed in table \ref{tab:board}):
 
@@
 
-There are two relevant trade-offs that the host needs to resolve. One is related with the visibility of the tag. The trade-off is among the
+There are two relevant trade-offs that the host needs to resolve. One is related with the visibility of the tag. The trade-off is among visuali impact, and visibility.
+
+Il secondo?
 
@@
 
-In our study we have considered a range of solutions that may help towards our goal of a sustainable support to a sustainable tourism, and specifically to the dynamic provision of information during a visit.
-To this end, the hosting organization (the {\em host} for short) implements a system of devices from which the visitor (or {\em user} ) extracts useful and enjoyable information.
-
-The purpose of such an installation is to guide the {\em user} in a tour that includes hurban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research).
-The task of the host organization is to provide the user with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.
-
-A traditional approach makes use of physical information boards with graphical or textual contents. Our study starts from finding the issues related with such a solution:
-
+As anticipated in the introduction, our study wants to provide a proof of concept for a sustainable solution in a concrete setup.
+To comply with such an holistic approach, we need to start from the definition of the operation context.
+
+This section is devoted to the description of the economical and social context where we are going to operate, as well as the historical backgound representing the heritage resource we want to promote.
+Next we will discuss the technical detail of the solution.
+
+The geographical area of interest are the surroundings of Casoli, a small village in a mountaineous region in the north of Tuscany, in Italy.
+The area fall within the municipailty of Bagni di Lucca, in the province of Lucca.
+The archaeological team of our project formerly realized the municipal archaeological map in 2021 \cite{letizia2021}, thus assessing the reasons of interest for the heritage and the geomorphological features of the area.
+
+\subsection{The natural and cultural resources of Casoli and its vicinity}
+
+Bagni di Lucca is located on the north-eastern boundary of the Province of Lucca, in Val di Lima, and is part of the Media Valle del Serchio district.
+With its mountains it marks the historical border with the Modena and Pistoia area, it is very rich in potential for its naturalistic, archaeological and historical heritage, both expressed and as yet unexpressed.
+Thanks to the archaeological map the main sites of interest from prehistoric to contemporary times in this municipality are catalogued, photographed and georeferenced.
+
+From a historical-geographical point of view, this area is identified with the Lima stream and its tributaries, that impress to the area a particular geomorphology, very impervious despite the modest hilly elevations.
+
+The Lima stream has characterised the history of Bagni di Lucca since ever: some of the oldest evidence of human presence in the valley has
+been found along its ancient river terraces, such as caves and rock shelters frequented since the Palaeolithic age; the
+manufacturing industry (paper mill, flour mills and, in recent times, energy production) have exploited the since the Middle Age until the 1980s \cite{men76, bed05, ser21}.
+
+On the naturalistic side, the region of Bagni di Lucca encompasses an incredible concentration of biodiversity, counting no less than three sites of the Natura 2000 Network \cite{natura2000} European Economic Community (EEC) initiative.
+
+There are three SCIs-SACs (Sites of Community Interest and Special Areas of Conservation) located in the area north of the Lima stream, corresponding to the Apennine portion, covering 23\% of the municipal surface:
+\begin {itemize}
+\item the limestone areas of Val di Lima and Balzo Nero; 
+\item Monte Prato Fiorito-Monte Coronato-Valle dello
+Scesta; 
+\item Orrido di Botri.
+\end{itemize}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.3\linewidth]{figure/toscana-casoli}
+	\caption[The region of Casoli]{}
+	\label{fig:toscana-casoli}
+\end{figure}
+
+The latter is also a SPA (Special Protection Area) and the Orrido di Botri State Reserve is located within it \cite{nat00}.
+It is therefore a natural heritage with a fragile balance that needs to be preserved.
+
+In recent years, the main tourist attractions have concentrated on the Lima stream, with some associations and private entities promoting outdoor experiences, particularly fluvial sports, such as canyoning, rafting and sup.
+Others important tourist attraction are trekking and hiking, supported by a network of path tracked by local CAI (Club Alpino Italiano).
+
+In recent years the community opened two entirely new trails:
 \begin{itemize}
-\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
-\item installation logistics: depending on the location, the placement of a plaque may require a basement or other sorts of supports
-\item environmental impact: in order to be effective, the plaque is prominent, and this may negatively affect the quality of the site 
-\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
-\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
-\item update limits: to update the content, the plaque must be replaced
-\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues like those found during the installation
+	\item the Alta Via dei Pastori (2019), a ring around Monte Prato Fiorito that takes up the ancient grazing route, and
+	\item t the Sentiero degli Avi (2020), a ring that from Montefegatesi
+	reaches Monte Coronato \cite{pin21}
 \end{itemize}
 
-Such considerations motivate an interest for an alternative way of communication.
-
-During the last decades we have witnessed the diffusion of smartphone devices, and we have no reasons to expect a change in such a thread.
-Smartphones empower the communication capabilities of an individual, allowing to receive sound, visual and tatcile interactions.
-It is therefore straightforward to consider the way to exploit such capabilities in our context.
-Also in this case we start from the limits of such technology.
-
-Concentrate on outdoor activities
-
-One is related with the diffusion of such a technology, which, although very diffused, is not equally familiar to everybody.
-The second is that several functions depends on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach an enabling infrastructure.
-
-Starting from the two aspects above we envision the directions towards a successful strategy, keeping in mind the overall sustainability of the approach.
-
-Regarding usability, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of installing new applications.
-
-The second issue is more complex, especially when considering outdoor activities.
-In such cases the provision of a networking infrastructure may be impossibile without a severe environmental impact.
-Consider the installation of antennas to cover a wide area and power supply for the radios.
-
-Starting from previous experiences documented in the bibliography we explore a few broad categories, to find weaknesses and strengths for each of them.
-
-One strategy concentrates on the wireless capabilities of the smartphone, although avoiding the coverage of the whole region.
-We consider the implementation of several small servers operating in isolation that provide a few HTML pages.
-The approach requires a modest investment and a marginal environmental impact: a small device based on the ESP8266 single chip computer has a coverage fo tens of meters, and a volume in the order of the $cm^3$.
-They operate as WiFi Access Points, so the user has access to the data in the device by joining the AP and browsing the URL.
-The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image.
-
-The dependency from power supply is a serious limitation of such kind of solution.
-A radio transmitter is a rather power consuming device, in the range of Watts.
-Even if operation is intermittent, the operation of the transmitter cannot be guaranteed for long periods and the host organizaton should consider the installation of a power generator, which contributes to the economical cost and environmental footprint of the device,
-
-An alternative consists in the utilization of passive devices, like Near Field trasmitters.
-The device is flat with a surface ranging from a credit card to a coin.
-The receiver must be very close to enable the device to transmits its content.
-The power needed to operate the radio is drained from the smartphone, so batteries are not needed.
-The capacity is in the range of the KB, one page on this journal.
-
-The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application.
-Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is trasferred to the smartphone as a chink of text, and can be treated as such.
-The device can read it aloud to meet the user's inabilities or a translator.
-For both functions off-line apps exist to compensate the lack of network connection.
-
-Another solution in the family of passive devices is the QR-code.
-Such technique does not require radio communication, but uses the smartphone camera to acquire a graphical code that can be translated into a text.
-The capacity of a QR-code depends on the number of dots in the code, which in turn depends on the size of the code and the smartphone camera.
-We may consider that the capacity of a QR-code roughly equals that of an NFC tag with comparable size but a simpler manufacturing.
-
-Such a minimalistic solution compares surprisingly well when compared with other solutions that make use of the user's smartphone.
-There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words are sufficient to convey the description of a site, or provide directions for the the visit.
-In exchange, the QR-code solution does not require any power supply, is extremely easy to use, has a limited impact on the laandscape, is not exposed to theft, has a negligible cost, is durable and produces a limited quantity of garbage when disposed, is readable at a moderate distance and can be read aloud, translated or stored by the user's device.
-It significantly outperforms any active device, except for the capacity, and the NFC compares favorably only when capacity and size are prominent.
-
-The adoption of QR-codes compares favorably with traditional boards, considering the strengths and limits as listed above.
-
-\begin{itemize}
-\item dimensions: a 2-300 characters QR-code has the dimensione of a poker card
-\item logistics: QR-code board can be installed on any sort of preexisting support
-\item impact: the board has a minimal interference with the landscape and may easely go un-noticed if not properly advertised
-\item accessibility for the visually impaired: the text can be read aloud
-\item international: the text can be translated
-\item update: the board can be easely replaced when the content turns obsolete
-\item disposal: the card releases a limited quantity of pollutants related with ink support (paper or plastic)
-\end{itemize}
-
-One of the issues the host must resolve is the availability of support for electronic devices. The provisioon of such support has a footprint that ha
-
-may be present, but, in the spirit of our investigation, power supply should be address renewable sources, like wind, sunlight and similar.
-
-One option discriminates the use of an electronic device on the side of the user.
-
-In case we assume the user does not have a electronic device with her, the 
-One option is to introduce active electronics: the touristic site administration provides a
+Another recently developed project (2019) is the expansion of the Saint Bartholomew's Path, which runs through the territory of Pistoia, with a
+variant that from Popiglio continues in five stages in the municipality of Bagni di Lucca to Pieve di Controne, acting as the 'Lucca gateway' to the Path \cite{camsb}.
+
+Such initiatives led to a considerable revival of interest among Italian and international hikers for the area, especially for the Apennine side of the valley.
+
+We see the opportunity to improve the valorisation and enjoyment of this area by applying a slow tourism approach and involving the community, especially in the southern part of the municipality, which is still little frequented, more hilly and therefore less travelled by the trail network.
+We propose the creation of geo-itineraries characterised by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.
+
+\subsubsection{An historical perspective of an italian mountain site}
+
+% The village of Casoli is located on the Lima stream.
+To the south-east of the settlement lies a modest relief with a small glacial lake.
+The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age .
+These frequentations can be attributed to a seasonal exploitation of resources and, later, with the climatic improvement, as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.
+
+South-east of the today location of Casoli lies a modest relief and a small glacial lake.
+The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age and used also as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.
+
+Ceramic fragments dated between the 3rd century B.C.
+and the 1st-2nd century A.D.
+prove that Ligurian populations occupied the region scattered and in small nuclei.
+The dedication of the Latin colony of Lucca in 180 B.C.
+marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.
+
+We have little evidence of settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of Buca La Piella, investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside.
+In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found.
+
+% Just outside the Casoli settlement, ceramic fragments datable between the 3rd century B.C.
+and the 1st-2nd century A.D.
+emerged from an excavation for the construction of a bowling green, testifying, together with other documented evidence in the area of Bagni di Lucca, how Ligurian populations settled in Val di Lima at least from the 3rd century B.C.
+onwards.
+
+% In this period, the settlement must have been rather scattered and in small nuclei.
+The dedication of the Latin colony of Lucca in 180 B.C.
+marked a decisive turning point in the Romanisation of the area and within a few years there was a definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.
+
+% From the data currently available, it can be stated that the mountainous hinterland retained a relatively marginal role.
+Archaeological traces for the Roman era are scarce, apart from the material that emerged from Buca La Piella, located east of Casoli along the Solco del Monte.
+
+%The cave was discovered and investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside.
+In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found.
+
+% This confirms a rarefied or seasonal frequentation with a possible division in “fundi”, probably linked to the exploitation of the mountain's natural resources, the products of the forest and pasture, as well as having represented a connection route with the Po Valley area, through the Apennine passes \cite{gia96, men81, cia03}.
+
+In addition to the scarcity of archaeological remains, there is a profound silence in the historical and literary sources for the Val di Lima from Roman and Late Antique times, although it can be assumed that the Val di Lima followed the fate of the city of Lucca, first with the occupation of the Goths and the brief Byzantine rule and then with the Longobard invasion and the formation of the Duchy of Lucca.
+In the hinterland, the Lombard settlement was concentrated according to a system consisting of castles and villages, mostly evidenced by documents from the 8th and 9th centuries and the remains of burial areas.
+In Longobard and Carolingian times, Lucca's Val di Lima was one of the three administrative districts into which the mountains were divided and was called “fines Contronenses” \cite{qui02, cia06, cia11}.
+
+The only find from this period is the Grotta di Arzale, also known as the 'Grotta Murata', a rock shelter that opens up in the locality of Sevilucchio, north-west of Casoli along a historic road connecting Cocciglia to Casoli.
+Fictile artefacts, defined as being from the 'barbarian' era, have been found here; in reality, the bibliographic and oral sources collected have indicated fragments of an olla with threading, the attestation of which in the archaeological contexts of Lucca and northern Tuscany starts from the 6th-7th century and may reach as far as the end of the 11th century \cite{gia96}.
+
+The first attestation of the settlements of Casoli and Lacu (Lake) (around the area of the Lake Casoli) dates back to the 10th century, the communities of which appear to belong to the parish of Vico Pancellorum.
+In this period, Bagni di Lucca was subdivided into four plebates, to which numerous villages were attached \cite{bar41, gia96}.
+The demographic and material entity of these agglomerates are rather elusive, but they testify to a wide-ranging distribution of the settlement with a strengthening of the social function of the churches as seats of aggregation of the population.
+Most documents show a widespread allivation by the bishop of Lucca of many properties, tithes, annuities and offerings of the parishes, a practice that incentivised the fractioning and alienation of ecclesiastical property in favour of certain members of the city aristocracy.
+In this way, they secured an accumulation of funds and power, heritages that were to the nucleus of the wealth of the future lords over the following centuries, until they emerged in
+the territory by building their castles and constructing seigniorial domains centred on these areas \cite{qui02, gia96, for12, for15}.
+
+Written sources mention a castle in Casoli from 1180, the year in which the fortification is in the hands of the Bishop of Lucca, but its foundation must be earlier, probably built during the 11th- 12th centuries.
+It occupies the summit of Colle delle Tanette, towering to the north of the village.
+
+Between the 13th and 15th centuries, the fortification was at the centre, along with the other castles of the Val di Lima, of clashes between Lucca, Pisa and Florence, with alternating fortunes, as it was a strategic border area for the power of Lucca.
+Today, very little is preserved and the area has been partly inhabited.
+The fortified complex must have been equipped with
+two walls, a tower on the highest portion of the hill that is poorly preserved and two buildings.
+
+According to some oral sources, part of the castle stones were re-used to construct the surrounding buildings \cite{gia96, for12, rom16}.
+The same document from 1180 mentions the church of San Donato, located in the main square on the north-western edge of the village of Casoli.
+The current structure dates back to the 12th-13th centuries, with subsequent renovations and an enlargement of the naves that partially incorporated the base of the bell tower.
+The façade is surmounted by a plastered moulded tympanum, while the side aisles are adorned with moulded cornices.
+Initially dedicated only to San Donato, it acquired the double dedication after the abandonment of the church of Sant'Andrea de Lacu.
+The latter is located on a plateau to the east of Lake Casoli and is attested from 1260 in the Estimi of the diocese of Lucca as being subject to the parish of Vico Pancellorum \cite{ber18, gia96, cap17}, but already from the pastoral visit of 1358 it appears to have been united to the church of Casoli and from the pastoral visit of 1467 we learn of its state of abandonment \cite{gia96, con12}.
+This was probably due to the depopulation of the lake area and the simultaneous strengthening of the town of Casoli, which was fortified and better protected during this unstable phase.
+
+Today, the romanesque church of Sant'Andrea de Lacu is in a state of decay, with a rectangular plan, ending on the east side with a semi-circular apse; the masonry is characterised by squared lithic blocks arranged in regular rows.
+Inside, there is a reused element of the previous building walled into the northern perimeter, testifying to an older origin, probably contemporary with the settlement of Lacu (Lake) mentioned in sources from the 10th century and no longer visible today.
+
+Nearby are the remains of a quadrangular building with masonry in lithic blocks arranged in fairly regular rows, dating back to the Middle Ages (13th century), possibly a residential building.
+
+In the early modern age, villages that survived the demographic crisis of the 14th-15th centuries experienced a progressive architectural renewal between the 16th and 18th centuries, which still largely characterizes the settlements today.
+Within Casoli, a series of residential buildings with imposing dated portals are preserved, some with courtyards, and mansions that denote a discrete deployment of resources by wealthy social classes.
+There is also a well-preserved bread oven, typically used by the entire community, which remained in use at least for village festivals until a few decades ago.
+
+Two oratories in the village date back to the modern age (17th-18th centuries): one dedicated to the Madonna of Loreto, flanked by a wash-house, and the other called Madonna di Castello and characterised by an entrance enclosure, located along the cobbled road that retraces the ancient road linking the village with the summit area of the medieval castle \cite{gia96}.
+
+Casoli was connected to the valley floor by two main roads attested in the Borbonic Cadastre, which from the village follow the gentler slopes to the east and south-west to ford the Lima torrent thanks to the Ponte Maggio and Ponte Nero bridges, still preserved today.
+Towards the hilly area, an articulated road network is evidenced, which is still legible in the landscape and in part is taken up by the CAI trail system.
+In particular, from the village starts the road that led toward Lucchio crossing the area of Casoli Lake.
+Along it are the chapel of Madonna di Col di Piano and two “metati” referable to the contemporary age, used for drying chestnuts.
+The entire area of the lake is covered with crops of centuries-old chestnut trees, typical of the Lima Valley, as of the entire Apennine area.
+Referred to as the "bread tree," chestnut fruits and the flour derived from them were a staple of the mountain diet until the mid-1900s, and today they are increasingly less common due to the disease that has affected their specimens and the slow replacement by allochthonous species \cite{buc92, puc10}.
+
+\subsubsection{The socio-economical context of Casoli and its vicinity}
+
+......
+
+\subsubsection{Sites and trail’s state of art}
+
+%The study of the Casoli area made it possible to identify the historical-archaeological sites considered of greatest interest for the purposes of a geotouristic route.
+Once these sites had been located and the possible paths to reach them analysed, also by consulting the online sharing of itineraries by hikers and climbers, we carried out field surveys between April and June 2023 to check the status of the trails and sites for their enjoyment.
+(questo penso di metterlo dopo)
+
+As explained in the previous section, the area around Casoli has a long history witnessed by architectural and geological features: it is therefore a good candidate for our case study.
+In the rest of this section we define the features of interest, and report about their accessibility.
+
+%The site at the centre of the project's interest is Buca La Piella, a cave located along the Solco del Monte and off the conventional circuits, which is why it is less frequented and consequently less compromised for the scientific analyses that will be carried out there, compared to the caves located along the Lima stream, near the state highway.
+The first part of the route coincides for about 500 m with CAI path 386, which together with path 388 forms the Casoli-Lucchio ring around Monte Memoriante.
+Access to the path, which is poorly marked with a faded red and white simple signpost, is along the asphalt road that leads from Ponte Maggio to Casoli and starts from the slope with a short stone ramp.
+After a short climb, one walks at altitudealong the slope and then enters the forest, skirting the final part of the Solco del Monte, known as Rio di Castello.
+
+The cave named {\em Buca La Piella} is the foremost site from an archeological point of view.
+It is located in a fascinating environment outside conventional circuits, which is why it is less frequented and consequently less compromised.
+It is reachable leaving a marked path to follow the bed of the Lima river.
+The way is dotted with minor karsic caves, and the {\em Buca La Piella} is reachable climbing up on a steep slope in the wood on the right of the Lima river.
+
+%A faint track through the grass diverts from the path to climb up the side of the slope to the rock face where one of the many karstic formations that characterise the area opens up.
+This is a small cave, called the 'Antro dell'Ugola' in the Tuscan speleological register.
+It consists of two openings whose entrance has a conformation reminiscent of a uvula.
+There are no proven finds, but the space inside and the position could indicate a frequentation that will be verified with the comparison analyses that we will carry out.
+
+Not far from the {\em Buca La Piella} on the same side of the river is another karst cave named {\em Antro dell'Ugola}.
+It is reachable following a faint and intermittent track that departs from a marked path.
+It is craracterize by a suspended geological formation reminding an uvula, hence the name.
+
+Both caves are going to be analyzed in the course of the Underlandscape project \cite{Underlandscape}, and findings can further enhance the interest for both sites.
+
+\textbf{Qui aggiungerei notizie sugli altri punti in cui sono stati collocati i QR-code, correlandoli alla descrizione storica
+}
+
+%We return to the main path, this one looking over the Rio to proceed in the direction of Lucchio, while we abandon it to penetrate between the high rock walls carved over time by the Rio; mobile network coverage is absent, while the GPS has poor coverage and in many places is unreliable.
+When walking along the stream bed, great care is required, due to slippery pebbles and sections with wet foliage and moss.
+After about 300 m, follow a track that climbs up the slope to the right.
+When you reach a small plateau, you will find a rock shelter, which is not marked in the Tuscan cave cadastre and has not yet been investigated.
+Buca La Piella is located about 120 m further north.
+To reach it, climb up to the rock face by crossing a rather steep wooded slope.
+A great deal of caution is required, and for future use it would be a good idea to at least provide a safety rope.
+When you reach the wall, there is a short tunnel that leads into the cave.
+A very wide and high cavity opens up towards the east.
+The vault has a hole in the centre, and there are stalactites and stalagmites.
+The ground descends gently towards the opening, beyond which the slope becomes steep, making it necessary in the future to secure it with nets or fences.
+
+\subsection{Implementing a QR-based signage}
+
+To test our signage proposal, we placed on the field several cards carrying QR codes. Se the list in Table~\ref{tab:qrtags}.
+
+% TODO: \usepackage{graphicx} required
+\begin{figure}
+	\centering
+	\includegraphics[width=\linewidth]{figure/mappa}
+	\caption[Map with QR-tag positions]{}
+	\label{fig:mappa}
+\end{figure}
+
+\begin{table}
+	\begin{tabular}{|>{\raggedright\arraybackslash}p{3cm}|l|l|l|l|}
+\hline 
+\textbf{Name} & \textbf{Long} & \textbf{Lat} & \textbf{URL Key} & \textbf{Length} \\
+\hline
+Buca La Piella & 10.68361 & 44.03667 & t5ysrm & 476 \\
+\hline
+Calendario celtico & 10.66985 & 44.03986 & my0kp8 & 484 \\
+\hline
+Castello & 10.67062 & 44.04004 & 4l4r6y & 597 \\
+\hline
+Chiesa dei SS. Donato e Andrea & 10.66947 & 44.03928 & lwtyx6 & 632 \\
+\hline
+Iscrizione longobarda & 10.67244 & 44.03906 & 60m75s & 369 \\
+\hline
+Lago di Casoli & 10.67761 & 44.03468 & xqjpbk & 369 \\
+\hline
+Madonna di Castello & 10.66994 & 44.03960 & qlci89 & 395 \\
+\hline
+Madonna di Col di Piano & 10.67760 & 44.03173 & 3w44wr & 414 \\
+\hline
+Metato & 10.67764 & 44.03591 & sbgnl0 & 660 \\
+\hline
+Sant'Andrea de Lacu & 10.67592 & 44.03530 & fxq83v & 657 \\
+\hline
+Antro dell’Ugola & 10.68296 & 44.03871 & mprs0w & 226 \\
+\hline
+Deviazione per Antro dell’Ugola & 10.68343 & 44.03971 & e4n2js & 88 \\
+\hline
+Deviazione per Buca La Piella & 10.68427 & 44.03573 & hve4pj & 139 \\
+\hline
+Deviazione per Sant’Andrea de Lacu & 10.67663 & 44.03507 & 13uav3 & 200 \\
+\hline
+Ingresso Buca La Piella & 10.68358 & 44.03592 & 5ahvp8 & 190 \\
+\hline
+Riparo sottoroccia lungo percorso Piella & 10.68397 & 44.03704 & kwr1wx & 286 \\
+\hline
+	\end{tabular}
+\caption{The QR-tags placed around Casoli.
+The URL key is a random code that appears in the web page linked to the QR code.
+The Length refers to the description in the QR-code and is in characters.
+\label{tab:qrtags}}
+\end{table}
+
+A card sample is in Figure \ref{fig:toscana-casoli}.
+On the top there is the logo of the funding project, a descriptive title ({\em Castle} in english) and concise directions for use {\em Scan fo further information}).
+Under the QR-code is the notice that other QR-codes are present in the vicinity, and a reference to the funding project.
+
+The text in the QR-code (in figure \ref{fig:qrcodecontent}) contains the main text with 614 characters (100 words) and a URL 86 characters long.
+The URL is useful only if the Internet is reachable, and brings to a new page containing a map with the location of all the tags (Figure \ref{fig:linkmap}) and a short survey form.
+Note that, while the QR-tag content is always available, the URL is reachable in locations with broadband coverage.
+In our case, during a recon to the {\em Piella} cave none of the smartphones of the participants received sufficient broadband network signal.
+
+\begin{figure}
+\subfloat[QR-tag content]{
+	\includegraphics[width = 0.3 \linewidth]{figure/qr-code-content}
+	\label{fig:qrcodecontent}
+}
+\subfloat[Linked page: the map]{
+	\includegraphics[width = 0.3 \linewidth]{figure/sito-a}
+	\label{fig:linkmap}
+}
+\subfloat[Linked page: survey form]{
+	\includegraphics[width = 0.3 \linewidth]{figure/sito-b}
+	\label{fig:linksurvey}
+}
+\end{figure}
+
+We placed QR-tags both along the trail and at sites of interest; the former are placed at detours and, at more complex points, at close range so that from one you can see the next, as per the CAI standard \cite{cai10}.
+
+They can therefore have either the function of a simple signpost to indicate the continuity of the trail, or be scanned to read the written indications recorded within it, which are more accurate and precise.
+
+%QR codes placed at sites of interest provide historical, archaeological and naturalistic information through short text descriptions.
+They also give the possibility, where phone coverage allows, to deepen the project and discover more details about the proposed itinerary through a link to the dedicated site.
+
+This approach ensures that the tourist is guided through the visitor experience and, at the same time, that the impact of signage on the environment is reduced compared with traditional boards, as explained in section \ref{sec:minimize}.
+
+%, considering that there is no need for panels and vertical signage involving the placement of poles.
+In addition to being more impactful, classic vertical signage can more easily suffer degradation and needs more maintenance, especially the information paneling on the sites; whereas in our proposal the possible maintenance of the QR code would constitute very short time and rather low cost.
+
+Clearly, our methodology does not aim to replace CAI signage, which is recognized throughout the country and is very effective, especially in terms of horizontal typology, but at the very least it is repurposed to complement it in order to provide more detailed information at multiple levels of interest, from short text to in-depth information on the project site.
+In addition, being able to read the written directions along the path can be of great help to even the most inexperienced hikers, particularly on unclear and not-quite-visible trails such as the one we suggested to reach Buca La Piella.
+
+% ====
+
+% Concentrate on outdoor activities
+
+% Starting from previous experiences documented in the bibliography we explore a few broad categories, to find weaknesses and strengths for each of them.
+
+% One of the issues the host must resolve is the availability of support for electronic devices. The provisioon of such support has a footprint that ha
+
+% may be present, but, in the spirit of our investigation, power supply should be address renewable sources, like wind, sunlight and similar.
+
+% One option discriminates the use of an electronic device on the side of the user.
+
+% In case we assume the user does not have a electronic device with her, the 
+% One option is to introduce active electronics: the touristic site administration provides a
 
@@
 
-This section may be divided by subheadings.
+%This section may be divided by subheadings.
 It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
-\subsection{Subsection}
-\subsubsection{Subsubsection}
-
-Bulleted lists look like this:
-\begin{itemize}
-\item	First bullet;
-\item	Second bullet;
-\item	Third bullet.
-\end{itemize}
-
-Numbered lists can be added as follows:
-\begin{enumerate}
-\item	First item; 
-\item	Second item;
-\item	Third item.
-\end{enumerate}
-
-The text continues here.
-
-\subsection{Figures, Tables and Schemes}
-
-All figures and tables should be cited in the main text as Figure~\ref{fig1}, Table~\ref{tab1}, etc.
-
-\begin{figure}[H]
-\includegraphics[width=10.5 cm]{Definitions/logo-mdpi}
-\caption{This is a figure.
+%\subsection{Subsection}
+%\subsubsection{Subsubsection}
+
+%Bulleted lists look like this:
+%\begin{itemize}
+%\item	First bullet;
+%\item	Second bullet;
+%\item	Third bullet.
+%\end{itemize}
+
+%Numbered lists can be added as follows:
+%\begin{enumerate}
+%\item	First item; 
+%\item	Second item;
+%\item	Third item.
+%\end{enumerate}
+
+%The text continues here. 
+%
+%\subsection{Figures, Tables and Schemes}
+%
+%All figures and tables should be cited in the main text as Figure~\ref{fig1}, Table~\ref{tab1}, etc.
+%
+%\begin{figure}[H]
+%\includegraphics[width=10.5 cm]{Definitions/logo-mdpi}
+%\caption{This is a figure.
 Schemes follow the same formatting.
@@
 A caption on a single line should be centered.\label{fig1}}
-\end{figure}   
-\unskip
-
-\begin{table}[H] 
-\caption{This is a table caption. Tables should be placed in the main text near to the first time they are~cited.\label{tab1}}
-\newcolumntype{C}{>{\centering\arraybackslash}X}
-\begin{tabularx}{\textwidth}{CCC}
-\toprule
-\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
-\midrule
-Entry 1		& Data			& Data\\
-Entry 2		& Data			& Data \textsuperscript{1}\\
-\bottomrule
-\end{tabularx}
-\noindent{\footnotesize{\textsuperscript{1} Tables may have a footer.}}
-\end{table}
-
-The text continues here (Figure~\ref{fig2} and Table~\ref{tab2}).
-
-% Example of a figure that spans the whole page width. The same concept works for tables, too.
-\begin{figure}[H]
-\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-\includegraphics[width=15.5cm]{Definitions/logo-mdpi}
-\end{adjustwidth}
-\caption{This is a wide figure.\label{fig2}}
-\end{figure}
-
-\begin{table}[H]
-\caption{This is a wide table.\label{tab2}}
-	\begin{adjustwidth}{-\extralength}{0cm}
-		\newcolumntype{C}{>{\centering\arraybackslash}X}
-		\begin{tabularx}{\fulllength}{CCCC}
-			\toprule
-			\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}     & \textbf{Title 4}\\
-			\midrule
-\multirow[m]{3}{*}{Entry 1 *}	& Data			& Data			& Data\\
-			  	                   & Data			& Data			& Data\\
-			             	      & Data			& Data			& Data\\
-                   \midrule
-\multirow[m]{3}{*}{Entry 2}    & Data			& Data			& Data\\
-			  	                  & Data			& Data			& Data\\
-			             	     & Data			& Data			& Data\\
-                   \midrule
-\multirow[m]{3}{*}{Entry 3}    & Data			& Data			& Data\\
-			  	                 & Data			& Data			& Data\\
-			             	    & Data			& Data			& Data\\
-                  \midrule
-\multirow[m]{3}{*}{Entry 4}   & Data			& Data			& Data\\
-			  	                 & Data			& Data			& Data\\
-			             	    & Data			& Data			& Data\\
-			\bottomrule
-		\end{tabularx}
-	\end{adjustwidth}
-	\noindent{\footnotesize{* Tables may have a footer.}}
-\end{table}
+%\end{figure}   
+%\unskip
+%
+%\begin{table}[H] 
+%\caption{This is a table caption. Tables should be placed in the main text near to the first time they are~cited.\label{tab1}}
+%\newcolumntype{C}{>{\centering\arraybackslash}X}
+%\begin{tabularx}{\textwidth}{CCC}
+%\toprule
+%\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
+%\midrule
+%Entry 1		& Data			& Data\\
+%Entry 2		& Data			& Data \textsuperscript{1}\\
+%\bottomrule
+%\end{tabularx}
+%\noindent{\footnotesize{\textsuperscript{1} Tables may have a footer.}}
+%\end{table}
+%
+%The text continues here (Figure~\ref{fig2} and Table~\ref{tab2}).
+%
+%% Example of a figure that spans the whole page width. The same concept works for tables, too.
+%\begin{figure}[H]
+%\begin{adjustwidth}{-\extralength}{0cm}
+%\centering
+%\includegraphics[width=15.5cm]{Definitions/logo-mdpi}
+%\end{adjustwidth}
+%\caption{This is a wide figure.\label{fig2}}
+%\end{figure}  
+%
+%\begin{table}[H]
+%\caption{This is a wide table.\label{tab2}}
+%	\begin{adjustwidth}{-\extralength}{0cm}
+%		\newcolumntype{C}{>{\centering\arraybackslash}X}
+%		\begin{tabularx}{\fulllength}{CCCC}
+%			\toprule
+%			\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}     & \textbf{Title 4}\\
+%			\midrule
+%\multirow[m]{3}{*}{Entry 1 *}	& Data			& Data			& Data\\
+%			  	                   & Data			& Data			& Data\\
+%			             	      & Data			& Data			& Data\\
+%                   \midrule
+%\multirow[m]{3}{*}{Entry 2}    & Data			& Data			& Data\\
+%			  	                  & Data			& Data			& Data\\
+%			             	     & Data			& Data			& Data\\
+%                   \midrule
+%\multirow[m]{3}{*}{Entry 3}    & Data			& Data			& Data\\
+%			  	                 & Data			& Data			& Data\\
+%			             	    & Data			& Data			& Data\\
+%                  \midrule
+%\multirow[m]{3}{*}{Entry 4}   & Data			& Data			& Data\\
+%			  	                 & Data			& Data			& Data\\
+%			             	    & Data			& Data			& Data\\
+%			\bottomrule
+%		\end{tabularx}
+%	\end{adjustwidth}
+%	\noindent{\footnotesize{* Tables may have a footer.}}
+%\end{table}
 
@@
 
-Text.
-
-Text.
-
-\subsection{Formatting of Mathematical Components}
-
-This is the example 1 of equation:
-\begin{linenomath}
-\begin{equation}
-a = 1,
-\end{equation}
-\end{linenomath}
-the text following an equation need not be a new paragraph. Please punctuate equations as regular text.
-%% If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments).
+%Text.
+%
+%Text.
+%
+%\subsection{Formatting of Mathematical Components}
+%
+%This is the example 1 of equation:
+%\begin{linenomath}
+%\begin{equation}
+%a = 1,
+%\end{equation}
+%\end{linenomath}
+%the text following an equation need not be a new paragraph. Please punctuate equations as regular text.
+%%% If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments).
 This ensures correct linenumbering.
 The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.
-
-This is the example 2 of equation:
-\begin{adjustwidth}{-\extralength}{0cm}
-\begin{equation}
-a = b + c + d + e + f + g + h + i + j + k + l + m + n + o + p + q + r + s + t + u + v + w + x + y + z
-\end{equation}
-\end{adjustwidth}
-
-% Example of a page in landscape format (with table and table footnote).
-%\startlandscape
-%\begin{table}[H] %% Table in wide page
-%\caption{This is a very wide table.\label{tab3}}
-%	\begin{tabularx}{\textwidth}{CCCC}
-%		\toprule
-%		\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}	& \textbf{Title 4}\\
-%		\midrule
-%		Entry 1		& Data			& Data			& This cell has some longer content that runs over two lines.\\
-%		Entry 2		& Data			& Data			& Data\textsuperscript{1}\\
-%		\bottomrule
-%	\end{tabularx}
-%	\begin{adjustwidth}{+\extralength}{0cm}
-%		\noindent\footnotesize{\textsuperscript{1} This is a table footnote.}
-%	\end{adjustwidth}
-%\end{table}
-%\finishlandscape
-
-Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows:
-%% Example of a theorem:
-\begin{Theorem}
-Example text of a theorem.
-\end{Theorem}
-
-The text continues here. Proofs must be formatted as follows:
-
-%% Example of a proof:
-\begin{proof}[Proof of Theorem 1]
-Text of the proof. Note that the phrase ``of Theorem 1'' is optional if it is clear which theorem is being referred to.
-\end{proof}
-The text continues here.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-
-Authors should discuss the results and how they can be interpreted from the perspective of previous studies and of the working hypotheses.
+%
+%This is the example 2 of equation:
+%\begin{adjustwidth}{-\extralength}{0cm}
+%\begin{equation}
+%a = b + c + d + e + f + g + h + i + j + k + l + m + n + o + p + q + r + s + t + u + v + w + x + y + z
+%\end{equation}
+%\end{adjustwidth}
+%
+%% Example of a page in landscape format (with table and table footnote).
+%%\startlandscape
+%%\begin{table}[H] %% Table in wide page
+%%\caption{This is a very wide table.\label{tab3}}
+%%	\begin{tabularx}{\textwidth}{CCCC}
+%%		\toprule
+%%		\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}	& \textbf{Title 4}\\
+%%		\midrule
+%%		Entry 1		& Data			& Data			& This cell has some longer content that runs over two lines.\\
+%%		Entry 2		& Data			& Data			& Data\textsuperscript{1}\\
+%%		\bottomrule
+%%	\end{tabularx}
+%%	\begin{adjustwidth}{+\extralength}{0cm}
+%%		\noindent\footnotesize{\textsuperscript{1} This is a table footnote.}
+%%	\end{adjustwidth}
+%%\end{table}
+%%\finishlandscape
+%
+%
+%Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows:
+%%% Example of a theorem:
+%\begin{Theorem}
+%Example text of a theorem.
+%\end{Theorem}
+%
+%The text continues here. Proofs must be formatted as follows:
+%
+%%% Example of a proof:
+%\begin{proof}[Proof of Theorem 1]
+%Text of the proof. Note that the phrase ``of Theorem 1'' is optional if it is clear which theorem is being referred to.
+%\end{proof}
+%The text continues here.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Discussion}
+%
+%Authors should discuss the results and how they can be interpreted from the perspective of previous studies and of the working hypotheses.
 The findings and their implications should be discussed in the broadest context possible.
 Future research directions may also be highlighted.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
-
-This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Patents}
-
-This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Conclusions}
+%
+%This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Patents}
+%
+%This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Optional
-\appendixtitles{no} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
-\appendixstart
-\appendix
-\section[\appendixname~\thesection]{}
-\subsection[\appendixname~\thesubsection]{}
-The appendix is an optional section that can contain details and data supplemental to the main text---for example, explanations of experimental details that would disrupt the flow of the main text but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data are shown in the main text can be added here if brief, or as Supplementary Data.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% Optional
+%\appendixtitles{no} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A").
+If the appendix sections contain a heading then change the argument to "yes".
+%\appendixstart
+%\appendix
+%\section[\appendixname~\thesection]{}
+%\subsection[\appendixname~\thesubsection]{}
+%The appendix is an optional section that can contain details and data supplemental to the main text---for example, explanations of experimental details that would disrupt the flow of the main text but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data are shown in the main text can be added here if brief, or as Supplementary Data.
 Mathematical proofs of results not central to the paper can be added as an appendix.
-
-\begin{table}[H] 
-\caption{This is a table caption.\label{tab5}}
-\newcolumntype{C}{>{\centering\arraybackslash}X}
-\begin{tabularx}{\textwidth}{CCC}
-\toprule
-\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
-\midrule
-Entry 1		& Data			& Data\\
-Entry 2		& Data			& Data\\
-\bottomrule
-\end{tabularx}
-\end{table}
-
-\section[\appendixname~\thesection]{}
-All appendix sections must be cited in the main text. In the appendices, Figures, Tables, etc. should be labeled, starting with ``A''---e.g., Figure A1, Figure A2, etc.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%\begin{table}[H] 
+%\caption{This is a table caption.\label{tab5}}
+%\newcolumntype{C}{>{\centering\arraybackslash}X}
+%\begin{tabularx}{\textwidth}{CCC}
+%\toprule
+%\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
+%\midrule
+%Entry 1		& Data			& Data\\
+%Entry 2		& Data			& Data\\
+%\bottomrule
+%\end{tabularx}
+%\end{table}
+%
+%\section[\appendixname~\thesection]{}
+%All appendix sections must be cited in the main text. In the appendices, Figures, Tables, etc. should be labeled, starting with ``A''---e.g., Figure A1, Figure A2, etc.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{adjustwidth}{-\extralength}{0cm}
-%\printendnotes[custom] % Un-comment to print a list of endnotes
+%%\printendnotes[custom] % Un-comment to print a list of endnotes
 

### 2023-07-20 paper.tex
Sostituito storico compresso di Letizia (manca biblio)

@@
 
-% The village of Casoli is located on the Lima stream.
-To the south-east of the settlement lies a modest relief with a small glacial lake.
-The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age .
-These frequentations can be attributed to a seasonal exploitation of resources and, later, with the climatic improvement, as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.
-
-South-east of the today location of Casoli lies a modest relief and a small glacial lake.
+ The hill village of Casoli is located south of the Lima stream.
 The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age and used also as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.
@@
 
-We have little evidence of settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of Buca La Piella, investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside.
-In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found.
-
-% Just outside the Casoli settlement, ceramic fragments datable between the 3rd century B.C.
-and the 1st-2nd century A.D.
-emerged from an excavation for the construction of a bowling green, testifying, together with other documented evidence in the area of Bagni di Lucca, how Ligurian populations settled in Val di Lima at least from the 3rd century B.C.
-onwards.
-
-% In this period, the settlement must have been rather scattered and in small nuclei.
-The dedication of the Latin colony of Lucca in 180 B.C.
-marked a decisive turning point in the Romanisation of the area and within a few years there was a definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.
-
-% From the data currently available, it can be stated that the mountainous hinterland retained a relatively marginal role.
-Archaeological traces for the Roman era are scarce, apart from the material that emerged from Buca La Piella, located east of Casoli along the Solco del Monte.
-
-%The cave was discovered and investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside.
-In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found.
-
-% This confirms a rarefied or seasonal frequentation with a possible division in “fundi”, probably linked to the exploitation of the mountain's natural resources, the products of the forest and pasture, as well as having represented a connection route with the Po Valley area, through the Apennine passes \cite{gia96, men81, cia03}.
-
-In addition to the scarcity of archaeological remains, there is a profound silence in the historical and literary sources for the Val di Lima from Roman and Late Antique times, although it can be assumed that the Val di Lima followed the fate of the city of Lucca, first with the occupation of the Goths and the brief Byzantine rule and then with the Longobard invasion and the formation of the Duchy of Lucca.
-In the hinterland, the Lombard settlement was concentrated according to a system consisting of castles and villages, mostly evidenced by documents from the 8th and 9th centuries and the remains of burial areas.
+We have little evidence of roman settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of Buca La Piella, investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside.
+In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found \cite{gia96, men81, cia03}.
+
 In Longobard and Carolingian times, Lucca's Val di Lima was one of the three administrative districts into which the mountains were divided and was called “fines Contronenses” \cite{qui02, cia06, cia11}.
-
-The only find from this period is the Grotta di Arzale, also known as the 'Grotta Murata', a rock shelter that opens up in the locality of Sevilucchio, north-west of Casoli along a historic road connecting Cocciglia to Casoli.
-Fictile artefacts, defined as being from the 'barbarian' era, have been found here; in reality, the bibliographic and oral sources collected have indicated fragments of an olla with threading, the attestation of which in the archaeological contexts of Lucca and northern Tuscany starts from the 6th-7th century and may reach as far as the end of the 11th century \cite{gia96}.
-
-The first attestation of the settlements of Casoli and Lacu (Lake) (around the area of the Lake Casoli) dates back to the 10th century, the communities of which appear to belong to the parish of Vico Pancellorum.
-In this period, Bagni di Lucca was subdivided into four plebates, to which numerous villages were attached \cite{bar41, gia96}.
-The demographic and material entity of these agglomerates are rather elusive, but they testify to a wide-ranging distribution of the settlement with a strengthening of the social function of the churches as seats of aggregation of the population.
-Most documents show a widespread allivation by the bishop of Lucca of many properties, tithes, annuities and offerings of the parishes, a practice that incentivised the fractioning and alienation of ecclesiastical property in favour of certain members of the city aristocracy.
-In this way, they secured an accumulation of funds and power, heritages that were to the nucleus of the wealth of the future lords over the following centuries, until they emerged in
-the territory by building their castles and constructing seigniorial domains centred on these areas \cite{qui02, gia96, for12, for15}.
-
-Written sources mention a castle in Casoli from 1180, the year in which the fortification is in the hands of the Bishop of Lucca, but its foundation must be earlier, probably built during the 11th- 12th centuries.
-It occupies the summit of Colle delle Tanette, towering to the north of the village.
-
+The only find from this period is the Grotta di Arzale, a rock shelter that opens up north-west of Casoli \cite{gia96}.
+
+The first attestation of the settlements of Casoli and Lacu (Lake) dates back to the 10th century \cite{bar41, gia96}.
+Most documents of this period show the fractioning and alienation of ecclesiastical heritage in favour of the city aristocracy, securing an accumulation of funds and power that was to form the basis of the subsequent domains \cite{qui02, gia96, for12, for15}.
+
+Written sources mention a castle in Casoli from 1180, but its foundation must be earlier.
 Between the 13th and 15th centuries, the fortification was at the centre, along with the other castles of the Val di Lima, of clashes between Lucca, Pisa and Florence, with alternating fortunes, as it was a strategic border area for the power of Lucca.
-Today, very little is preserved and the area has been partly inhabited.
-The fortified complex must have been equipped with
-two walls, a tower on the highest portion of the hill that is poorly preserved and two buildings.
-
-According to some oral sources, part of the castle stones were re-used to construct the surrounding buildings \cite{gia96, for12, rom16}.
-The same document from 1180 mentions the church of San Donato, located in the main square on the north-western edge of the village of Casoli.
-The current structure dates back to the 12th-13th centuries, with subsequent renovations and an enlargement of the naves that partially incorporated the base of the bell tower.
-The façade is surmounted by a plastered moulded tympanum, while the side aisles are adorned with moulded cornices.
+Today, very little is preserved and the area has been partly inhabited \cite{gia96, for12, rom16}.
+
+The same document from 1180 mentions the church of San Donato, located in the main square of the village of Casoli.
+The current structure dates back to the 12th-13th centuries, with subsequent renovations.
 Initially dedicated only to San Donato, it acquired the double dedication after the abandonment of the church of Sant'Andrea de Lacu.
-The latter is located on a plateau to the east of Lake Casoli and is attested from 1260 in the Estimi of the diocese of Lucca as being subject to the parish of Vico Pancellorum \cite{ber18, gia96, cap17}, but already from the pastoral visit of 1358 it appears to have been united to the church of Casoli and from the pastoral visit of 1467 we learn of its state of abandonment \cite{gia96, con12}.
+The latter is located on a plateau to the east of Lake Casoli and is attested from 1260 \cite{ber18, gia96, cap17}, but already in 15th century we learn of its state of abandonment \cite{gia96, con12}.
 This was probably due to the depopulation of the lake area and the simultaneous strengthening of the town of Casoli, which was fortified and better protected during this unstable phase.
 
-Today, the romanesque church of Sant'Andrea de Lacu is in a state of decay, with a rectangular plan, ending on the east side with a semi-circular apse; the masonry is characterised by squared lithic blocks arranged in regular rows.
-Inside, there is a reused element of the previous building walled into the northern perimeter, testifying to an older origin, probably contemporary with the settlement of Lacu (Lake) mentioned in sources from the 10th century and no longer visible today.
-
-Nearby are the remains of a quadrangular building with masonry in lithic blocks arranged in fairly regular rows, dating back to the Middle Ages (13th century), possibly a residential building.
-
-In the early modern age, villages that survived the demographic crisis of the 14th-15th centuries experienced a progressive architectural renewal between the 16th and 18th centuries, which still largely characterizes the settlements today.
+Today, the romanesque church of Sant'Andrea de Lacu is in a state of decay, with a rectangular plan, ending on the east side with a semi-circular apse.
+Inside, there is a reused element of the previous building, testifying to an older origin, probably contemporary with the ancient settlement of Lacu (Lake) mentioned in sources from the 10th century and no longer visible today.
+
+In the early modern age, villages of the area experienced a progressive architectural renewal, which still largely characterizes the settlements today.
 Within Casoli, a series of residential buildings with imposing dated portals are preserved, some with courtyards, and mansions that denote a discrete deployment of resources by wealthy social classes.
-There is also a well-preserved bread oven, typically used by the entire community, which remained in use at least for village festivals until a few decades ago.
-
-Two oratories in the village date back to the modern age (17th-18th centuries): one dedicated to the Madonna of Loreto, flanked by a wash-house, and the other called Madonna di Castello and characterised by an entrance enclosure, located along the cobbled road that retraces the ancient road linking the village with the summit area of the medieval castle \cite{gia96}.
-
-Casoli was connected to the valley floor by two main roads attested in the Borbonic Cadastre, which from the village follow the gentler slopes to the east and south-west to ford the Lima torrent thanks to the Ponte Maggio and Ponte Nero bridges, still preserved today.
-Towards the hilly area, an articulated road network is evidenced, which is still legible in the landscape and in part is taken up by the CAI trail system.
-In particular, from the village starts the road that led toward Lucchio crossing the area of Casoli Lake.
-Along it are the chapel of Madonna di Col di Piano and two “metati” referable to the contemporary age, used for drying chestnuts.
-The entire area of the lake is covered with crops of centuries-old chestnut trees, typical of the Lima Valley, as of the entire Apennine area.
-Referred to as the "bread tree," chestnut fruits and the flour derived from them were a staple of the mountain diet until the mid-1900s, and today they are increasingly less common due to the disease that has affected their specimens and the slow replacement by allochthonous species \cite{buc92, puc10}.
+Also dating from the modern era is the Madonna di Castello Oratory, characterised by an entrance eclosure and located along the cobbled road that traces the ancient route between the village and the summit area of the medieval castle \cite{gia96}.
+
+Along the road that led toward Lucchio crossing the area of Casoli Lake there are the chapel of Madonna di Col di Piano and two “metati” referable to the contemporary age, used for drying chestnuts.
+Referred to as the "bread tree," chestnut fruits and the flour derived from them were a staple of the mountain diet until the mid-1900s \cite{buc92, puc10}.
 
@@
 	\includegraphics[width=\linewidth]{figure/mappa}
-	\caption[Map with QR-tag positions]{}
+	\caption[Map with QR-tag positions]{Map with QR-tag positions}
 	\label{fig:mappa}

### 2023-07-20 paper.tex
Inizio capitolo risultati

@@
 
-There are two relevant trade-offs that the host needs to resolve. One is related with the visibility of the tag. The trade-off is among visuali impact, and visibility.
-
-Il secondo?
+There are two relevant trade-offs that the host needs to resolve.
+One is related with the visibility of the tag.
+The trade-off is among visuali impact, and visibility.
+The other is related with monitoring tag utilization.
+This may be useful for all sorts of planning activities, and is discussed in section \ref{sec:results}.
 
@@
 
-The text in the QR-code (in figure \ref{fig:qrcodecontent}) contains the main text with 614 characters (100 words) and a URL 86 characters long.
+The tag is printed on Synaps™, a non-biodegradable polyester synthetic film by Agfa-Gevaert NV.
+Agfa documents the production process giving guarantees of sustainability \cite{agfa}.
+According to Agfa, Synaps is more sustainable than laminated paper.
+From previous experience it is also more durable.
+
+To keep the environmental footprint as leightweight as possible, the tags are tied to pre-existent structures, like trees or poles. The tag is tied with a thin biodegradable string.
+
+The text encode on the QR-tag (in figure \ref{fig:qrcodecontent}) contains the main text with 614 characters (100 words) and a URL 86 characters long.
 The URL is useful only if the Internet is reachable, and brings to a new page containing a map with the location of all the tags (Figure \ref{fig:linkmap}) and a short survey form.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
+\section{Results and discussion \label{sec:results}}
+
+Our point is to evaluate all aspects of a signage system addressing the guidance of visitors in a rural area with a cultural heritage to be protected and exploited at the same time.
+
+The selected site exhibits the aspects of interest: a history that goes back to the Paleolithic Era, human artifacts as well as natural formations that witness past events.
+The site is not included in mass turism circuits, but is not far from touristic site.
+It is exposed to depopulation by absence of productive activities.
+({\bf Mariagrazia per favore conferma o nega}).
+
+Studying site heritage we found reasons of interest by a motivated, non-casual visitor: a cave inhabited since the pre-history with a difficult but viable access, a village with historical buildings, nearby sites dating back to the Middle Ages.
+
+The task of guiding the visitor and tell the location story has been left in charge to a dozen of QR-code tags placed at key locations.
+The costo of the QR-codes amounts to a few Euros for the printing service, but finding a shop with an appropriate printing device was not easy.
+
+The production of the graphical file was based on a simple script taking a geojson file as input and producing an image with all the QR.codes ready for printing on an A0 sheet.
+
+The preparation an implementation of the signage was therefore simple and with a low cost, but required a solid and pre-existent knowledge of the site.
+
+We include three stakeholder communities in the evaluation of the response to the initiative.
+We do not speak about {\em success} since the initiative is deliberately soft, and an outcome is expected in years, and its evaluation as successful or not will pose new questions.
+So we are content to see if the stakeholders regarded the initiative as an intrusion, as helpful, or simply neutral.
+
+We considered four stakeholder communities:
+
+\begin{itemize}
+	\item {\em the residents} the local community that currently inhabits the site
+	\item {\em entrepreneurs} who currently have a business in the site
+	\item {\em administrators} that are in charge of managing the site resources and that, at due time, will respond to the two stakeholders above, and finally
+	\item {\em the users}, those that come to Casoli to visit the Village or La Piella and that we find the QR-codes on their way
+\end{itemize}
+
+The category of the users escapes an evaluation since the QR-tag does not record reading operations.
+We added to the text message in the QR-tag an external link, that we monitor for the number of hits on pages dedicated to the specific tag: however this feedback certainly underestimates the number of times the tag is read.
+In Figure \ref{tab:hits} we see the number of hits during from May to July 2023.
+
+\begin{table}
+\begin{tabular}{|l|c|c|}
+\hline
+\textbf{Pagina di destinazione} & \textbf{Sessioni} \\
+\hline
+3w44wr & 4 \\
+e4n2js & 4 \\
+fxq83v & 4 \\
+t5ysrm & 4 \\
+mprs0w & 3 \\
+4l4r6y & 2 \\
+kwr1wx & 2 \\
+lwtyx6 & 2 \\
+sbgnl0 & 2 \\
+BM6cAj & 1 \\
+8tml9x & 1 \\
+bm6caj & 1 \\
+xqjpbk & 1 \\
+\hline
+\end{tabular}
+\caption{Number of hits per URL linked to a tag \label{tab:hits}]}
+\end{table}
+
+\section{Conclusions}
+
+There is a demand for low-power devices able to manifest themselves to the visitor and to monitor...
 

### 2023-11-04 paper.tex
Commit riportato del 19 ottobre

--- 2023-07-20 paper.tex
@@
 \documentclass[sustainability,article,submit,pdftex,moreauthors]{Definitions/mdpi} 
-
 %--------------------
@@
 % For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal. Changing "submit" to "accept" before posting will remove line numbers.
-
 %---------
@@
 % supfile = supplementary materials
-
 %----------
@@
 Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
-
 %------------------
@@
 % If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
-
 %---------
@@
 % The option pdftex is for use with pdfLaTeX. Remove "pdftex" for (1) compiling with LaTeX & dvi2pdf (if eps figures are used) or for (2) compiling with XeLaTeX.
-
 %=================================================================
@@
 \usepackage[caption = false]{subfig}
+\usepackage{tablefootnote}
+\usepackage{multibib}
+\usepackage{listings}
+\usepackage{xcolor}
+
+\newcites{Web}{Sitography}
+
+\colorlet{punct}{red!60!black}
+\definecolor{background}{HTML}{EEEEEE}
+\definecolor{delim}{RGB}{20,105,176}
+\colorlet{numb}{magenta!60!black}
+
+\lstdefinelanguage{json}{
+    basicstyle=\normalfont\ttfamily,
+    numbers=left,
+    numberstyle=\scriptsize,
+    stepnumber=1,
+    numbersep=8pt,
+    showstringspaces=false,
+    breaklines=true,
+    frame=lines,
+    backgroundcolor=\color{background},
+    literate=
+     *{0}{{{\color{numb}0}}}{1}
+      {1}{{{\color{numb}1}}}{1}
+      {2}{{{\color{numb}2}}}{1}
+      {3}{{{\color{numb}3}}}{1}
+      {4}{{{\color{numb}4}}}{1}
+      {5}{{{\color{numb}5}}}{1}
+      {6}{{{\color{numb}6}}}{1}
+      {7}{{{\color{numb}7}}}{1}
+      {8}{{{\color{numb}8}}}{1}
+      {9}{{{\color{numb}9}}}{1}
+      {:}{{{\color{punct}{:}}}}{1}
+      {,}{{{\color{punct}{,}}}}{1}
+      {\{}{{{\color{delim}{\{}}}}{1}
+      {\}}{{{\color{delim}{\}}}}}{1}
+      {[}{{{\color{delim}{[}}}}{1}
+      {]}{{{\color{delim}{]}}}}{1},
+}
 
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{
-Guiding the visitor to appreciate all aspects of tourist attractions is paramount for building a memorable experience.
+\abstract{Guiding the visitor to appreciate all the attractions of a region is paramount for building a memorable experience.
 In the literature, we find many articles exploring the use of sophisticated technologies towards such a goal.
-The case of wilderness resources exhibits a peculiar aspect, since the presence of human artifacts and the signage, may, in fact, damage the experience, and introduce pollutants as well.
+The case of wilderness resources exhibits a peculiar aspect, since the presence of human artifacts and the signage may damage the experience, and introduce pollutants as well.
 In this paper, we consider the case of a natural trail targeted to niche tourism to understand the available options for signage aiming to provide directions, guide the exploration, and warn about dangers.
-The footprints of the various options are evaluated, and we describe the outcomes of an experimental setup implemented to verify our conclusions' applicability.
-}
+The footprints of the various options are evaluated, and we describe the outcomes of an experimental setup implemented to verify the applicability of our approach.}
 %\abstract{A single paragraph of about 200 words maximum.
@@
 We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations.
-The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.}
+The abstract should be an objective representation of the article, it must not contain results that are not presented and substantiated in the main text and should not exaggerate the main conclusions.}
 
@@
 
+%\newcites{Main}{References}
+%\newcites{Sitology}{Sitology}
+
+\tableofcontents
+
 %\setcounter{section}{-1} %% Remove this when starting to work on the template.
@@
 
-Although an official decision is still pending, there is a sound proposal of naming the time we are living as "anthropocene".
-Behind such a proposal is the recognition that humans have the potential to significantly alter the environment they live in at a global scale.
-
-The problem behind the above fact is that we humans are not able to understand the consequences of the deployment of such a potential, and not even to keep it under control.
-Even worse, such capabilities are often used aiming at short-term local effects, ignoring long-term global ones, with the risk of a long-term global deterioration.
-Such an attitude pervades all activities, starting from the most basic ones like the provision of food and homes, and is frequenly biased by economical profit.
-
-Leisure activities play a relevant role, since they bring profit to niche businesses.
-This creates a reinforcement effect which follows the negative dynamic seen above: a leisure activity that per se is neutral (e.g.
-skying) may have a relevant footprint due to induced effects (ski lifts) which reinforces the success of the activity (easier to practice) bringing more investments and impact.
-Unlike beneficial effects, which is usually well localized, the negative impact has a very wide spectrum encompassing environmental, social and finally economical domains, as discussed by Patthey in \cite{pat08a}.
-
-In an effort to manage such a situation, international agencies promote the "sustainability" keyword as a guideline to the conservation of our habitat.
+Although an official decision is still pending, there is a widely agreed proposal for the naming of the time that we are living as the Anthropocene.
+Behind this proposal is the recognition that humans have the potential to significantly alter the environment in which they live on a global scale.
+
+The problem with the above fact is that we humans are not able to understand long-term consequences of deploying such a potential, nor even to keep it under control.
+Even worse, such capacities are often used with the aim of short-term local effects, ignoring long-term global ones, with the risk of long-term global deterioration.
+This attitude pervades all activities, starting with the most basic ones such as the supply of food and homes, and is often influenced by economic profit.
+
+Leisure activities play an important role in bringing profit to niche businesses.
+This creates a reinforcing effect that follows the negative dynamics seen above: a leisure activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact.
+Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a}.
+
+To manage this situation, international agencies promote the keyword "sustainability" as a guideline for the conservation of our habitat.
 Jeffry Ramsey et al.
-\cite{ram15a} consider the concept behind the word is "vague and contested but not meaningless" and that it cannot be used without a concrete framework, especially in a normative context.
+\cite{ram15a} consider the concept behind this word as "vague and disputed but not meaningless" and that cannot be used without a concrete framework, especially in a normative context.
 To solve to problem of measuring sustainability, Tom Kuhlman et al.
-point out a inherent tension in the term between growth and stability.
-A tension that has been softed by a shift in meaning of the word, which moved from focusing on preservation for an indefinite future to a concept encompassing three "pillars", society, economy and environment, as determinants that need to be protected in the present in view of a future sustainable development.
-The authors put into relationship man-made and environmental assets, discriminating the two cases in which a loss in the latter may or may not be compensated by an increment of the former, introducinging the terms of {\em weak} and {\em strong} sustainability.
-
-The promotion of the human cultural heritage has its stand in such a complex scenario.
-As noted by Jorge Otero in \cite{ote22a}, local communities take advantage of heritage promotion: the preservation activity generates social growth, while tourism provides an extra income that may be re-invested in promotion.
-However, just like in the sky resorts case mentioned above, the risk of compromising the resource itself exists.
-Such an interplay rises a question of sustainability even for cultural preservation initiatives, two topics that are usually regarded as in synergy.
-As in the "World Heritage and Sustainable Development" document by Unesco \cite{lab17a} that calls for actions that "harness the potential of World Heritage properties and heritage in general, to contribute to sustainable development".
-
-This paper originates from such a painful contradiction trying to give an answer to the question: is there a way to unlock the potential of a cultural resource limiting the damage to a social, economic and environmental equilibrium?
-
-The question arises in the course of a project funded by the italian ministry of reasearch aiming at the application of cutting edge technologies to the investigation of karst caves that may have been inhabited in the past, from prehistory to today.
-Among the results expected from the project there is the definition of a strategy to disseminate research results, and allow non-specialists to visit and enjoy the findings, thus contributing to the economy of rural areas that are severely exposed to de-population.
-A sustainable strategy is even more crucial given that the geological enviroment of interest is notably fragile, as noted by Aleksander Antic\`c et al.
+\cite{kuh10a} point out an inherent tension in the term between growth and stability.
+This tension has been attenuated by a change in the meaning of the word, which has gone from focusing on preservation for an indefinite future to a concept that includes three "pillars", society, economy, and environment, as determinants that need to be protected in the present to promote sustainable development in the future.
+The authors relate artificial and environmental goods, discriminating the two cases where a loss in the latter can or cannot be compensated by an increase in the former, introducing the terms of {\em weak} and {\em strong} sustainability.
+
+The promotion of human cultural heritage is situated in such a complex scenario.
+As noted by Jorge Otero in \cite{ote22a}, local communities benefit from the promotion of heritage: conservation activity generates social growth, while tourism provides an extra income that can be reinvested in promotion.
+However, just as in the case of ski resorts, there is a risk of compromising the resource itself.
+This interaction raises a question of sustainability also for cultural conservation initiatives, two issues that are usually considered in synergy.
+As in the document "World Heritage and Sustainable Development" of UNESCO, which calls for actions that "harness the potential of World Heritage properties and heritage in general, to contribute to sustainable development".
+
+This research stems from such a painful contradiction in trying to answer the question: is there a way to unlock the potential of a cultural resource while at the same time limiting the damage to a social, economic, and environmental equilibrium?
+
+The question arises in the course of a project funded by the Italian Ministry of Research aimed at applying cutting-edge technologies to the investigation of karst caves that may have been inhabited in the past, from prehistory to today.
+Among the expected results of the project is the definition of a strategy to disseminate the results of the research and allow non-specialists to visit and enjoy the results, thus contributing to the economy of rural areas that are severely exposed to de-population.
+A sustainable strategy is even more crucial as the geological environment of interest is particularly fragile, as noted by Aleksander Antic\`c et al.
 in \cite{ant20a}.
 
-This paper provides a concept solution of a specific aspect.
-The solution is conceptual in that it tries to follow a guideline that takes into account as many side details as possible in a way that can be used in other milieu, but very practical in its unfolding.
-Given the tighly interconnected nature of the situation, we observe how a very specific aspect, the touristic signage, exhibits effects on the society, economy and environment.
-
-\section{Unlocking site potentials through signage}
-
-The signage used to guide and inform the visitor is of paramount importance and must be carefully designed in order to accomplish its functions while following the guidelines of a sustainable approach.
+This document provides a conceptual solution for a specific aspect.
+The solution is conceptual in that it tries to follow a guideline that takes into account as much detail as possible in a way that can be used in other environments but is very practical in its application.
+Given the closely interlinked nature of the situation, we observe how a very specific aspect, tourist signage, exhibits effects on society, the economy, and the environment.
+
+\section{Unlocking site potentials through signage \label{sec:signage}}
+
+The signage used to guide and inform the visitor is of paramount importance and must be carefully designed to accomplish its functions while following the guidelines of a sustainable approach.
+
+\begin{figure}
+\subfloat[CAI mark on a tree]{
+    \includegraphics[width = 0.66 \linewidth]{figure/CAI.png}
+    \label{fig:CAI}
+}
+\hfill
+\subfloat[A direction sign]{
+    \label{fig:pole-direction}
+    \includegraphics[width = 0.285 \linewidth]{figure/cartelli.png}
+}
+\caption{Signs frequently found on trails in Italy}
+\label{fig:traditional}
+\end{figure}
 
@@
 The ideal signage should include visible reference points and explicit actions, like "turn right after crossing the stream".
-A good example of such a signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark.
-The marks are placed in such a way that, standing nearby one of them, the hiker is able to spot the successive on the trail
-\item a sign should explain the reasons of interest of a site.
-This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit.
-In this sense, the CAI signal above is not of use.
-A simple board with a site name may be sufficient only when the site is sufficiently reknown.
+A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}).
+The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail
+\item a sign should explain the reasons for interest in a site.
+This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit (see figure \ref{fig:pole-direction}).
+In this sense, the CAI signal above is not useful.
+A simple board with a site name may be sufficient only when the site is sufficiently popular.
 Otherwise, a more structured message should be used to explain the reasons why the feature is regarded as relevant.
@@
 
-In our study we consider a range of solutions that help towards a sustainable support to a sustainable tourism, and specifically to the dynamic provision of information during a visit along a natural trail, not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}).
-We consider the presence of two stakeholders for the signage itaself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.
-
-%In our study we have considered a range of solutions that may help towards our goal of a sustainable support to a sustainable tourism, and specifically to the dynamic provision of information during a visit.
-To this end, the hosting organization (the {\em host} for short) implements a system of devices from which the visitor (or {\em user} ) extracts useful and enjoyable information.
-
-The purpose of such an installation is guiding the {\em user} in a tour that includes urban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research).
-The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.
-A traditional approach makes use of physical information boards with graphical or textual contents.
-
-%The purpose of such an installation is to guide the {\em user} in a tour that includes hurban streets, buildings, natural trails and caves (this latter being the topic of the project giving financial support to our research).
-The task of the host organization is to provide the user with all sorts of information that may guide him across the visit, making it as much profitable and enjoyable as possible.
-
-Our study starts from pointing out the issues related with such a solution:
+In our study, we consider a range of solutions that help towards sustainable support to sustainable tourism, specifically to the dynamic provision of information during a visit along a natural trail, not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}).
+We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.
+
+The purpose of such an installation is guiding the {\em user} on a tour that includes urban streets, buildings, natural trails, and caves (this latter being the topic of the project giving financial support to our research).
+The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as profitable and enjoyable as possible.
+A traditional approach makes use of physical information boards with graphical or textual content, as in figure \ref{fig:traditional}.
+
+Our study starts by pointing out the issues related to such a solution:
 
@@
 	\item installation logistics: depending on the location, the transportation and placement of a plaque may require a basement or other sorts of supports
-	\item environmental impact: in order to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site 
+	\item environmental impact: to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site 
 	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
@@
 
-Such considerations motivate an interest for an alternative way of communication.
-
-%A traditional approach makes use of physical information boards with graphical or textual contents. Our study starts from finding the issues related with such a solution:
-
-%\begin{itemize}
-%	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
-%	\item installation logistics: depending on the location, the placement of a plaque may require a basement or other sorts of supports
-%	\item environmental impact: in order to be effective, the plaque is prominent, and this may negatively affect the quality of the site 
-%	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
-%	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
-%	\item update limits: to update the content, the plaque must be replaced
-%	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues like those found during the installation
-%\end{itemize}
-
-%Such considerations motivate an interest for an alternative way of communication.
+Such considerations motivate an interest in an alternative way of communication.
 
@@
 
-We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention on the environment, but the impact of such intervention must be better than that related with traditional signs.
-To this end we include in our solution tools that are not part of the site but remain with the user.
-
-During the last decades, we witnessed the diffusion of smartphone devices, and we have no reasons to expect a change in such a trend.
-Smartphones empower individual communication capabilities, allowing to receive sound, visual and tactile interactions.
-The relevance of such capabilities for the improvement of a tourist experience have been widely investigated, etiher in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a}.
-to find ways to exploit such tools in our context.
-Also in this case we start from the limits of such technology, especially in the case of outdoor activities.
-
-One is related with the diffusion of such a technology, which, although widely available, is not equally familiar to everybody.
-This is related with {\em usability}, with a term borrowed from P.
+We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention in the environment, but the impact of such intervention must be better than that related to traditional signs.
+To this end, we include in our solution tools that are not part of the site but remain with the user.
+
+During the last decades, we witnessed the diffusion of smartphone devices, and we have no reason to expect a change in such a trend.
+Smartphones empower individual communication capabilities, allowing them to receive sound, visual, and tactile interactions.
+The relevance of such capabilities for the improvement of a tourist experience has been widely investigated, either in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a} to find ways to exploit such tools.
+Again, we start from the limits of such technology, especially in the case of outdoor activities.
+
+One is that smartphones although widely available, are not equally familiar to everybody.
+This is related to {\em usability}, with a term borrowed from P.
 Wan in \cite{wan22a}.
-The second is that several functions depends on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach an enabling infrastructure.
+The second is that several functions depend on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach the Internet.
 So the {\em applicability limits} of a solution need to be defined.
 
-% The second issue is more complex, especially when considering outdoor activities.
-In such cases the provision of a networking infrastructure may be impossibile without a severe environmental impact.
-Consider the installation of antennas to cover a wide area and power supply for the radios.
-
-Starting from the two aspects above we envision the directions towards a successful smartphone-based strategy, keeping in mind the overall sustainability of the approach.
-
-Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need of familiarizing with new applications.
+Starting from the two aspects above we envision the guidelines for a successful smartphone-based strategy, keeping in mind its sustainability.
+
+Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need to familiarize themselves with new applications.
 
@@
 In such cases, the provision of a networking infrastructure incurs a severe environmental impact.
-Consider the installation of antennas to cover a wide area, and the power supply for the radios.
+Consider the installation of antennas to cover a wide area and the power supply for the radios.
 
@@
 Liu in \cite{liu16a}, which details an infrastructure that guides the visitors inside an urban milieu.
-In that case the presence of a pervasive networking facilities is a cornerstone for the whole architecture, that deeply depends on Internet connectivity.
-
-Wan \cite{wan22a} evaluates the quality of signage, without making reference to a specific technology, but with many examples showing physical boards, using as a formal reference the Universal Design Principles \cite{udi97a}.
-
-The number of research papers explodes when we extend the range to articles that investigate the impact of smartphones in tourism.
-The topic {\em smart tourism} topic is very popular, and covered by several review papers that provide a framework to the vast literature.
-
-A popular research direction covers the social aspects related with the use of the smartphone.
+In that case, the presence of pervasive networking facilities is a cornerstone for the whole architecture, which deeply depends on Internet connectivity.
+
+Wan \cite{wan22a} evaluates the quality of signage, without referring to a specific technology, but with many examples showing physical boards, using as a formal reference the Universal Design Principles \cite{udi97a}.
+
+The number of research papers explodes when we extend the range to articles investigating smartphones' impact on tourism.
+The {\em smart tourism} topic is very popular, and covered by several review papers that provide a framework to the vast literature.
+
+A popular research direction covers the social aspects related to the use of the smartphone.
 W.
-Tan \cite{tan17a} covers all aspects of a touristic experience related with the smartphone, from the definition of travel destination to assistance during the visit.
-Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is considered in the Internet, despite the title indicates a nature-based destination, where notably the Internet is not always reachable.
-
-On the other hand, roaming in places not reached by the Internet, or {\em dead zones} using the words of Pearce and Gretzel in \cite{pea12a}, may evoke constrasting feelings, from rewarding to threatening.
-
-More recently, the smartphone has been considered not strictly related with communication.
-In 2021 A.Slavec et al.
-investigate the use of cameras \cite{sla21a} while on travel in locations with a relevant cultural heritage in order to sustain its preservation and engage the tourist using location based games, similar to Pokemon Go or geo-caching.
+Tan \cite{tan17a} covers all aspects of a touristic experience related to the smartphone, from the definition of travel destination to assistance during the visit.
+Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is considered on the Internet.
+Such an assumption is in contrast with the title, which indicates a nature-based destination, where notably the Internet is not always reachable.
+
+On the other hand, roaming in places not reached by the Internet, or {\em dead zones} using the words of Pearce and Gretzel in \cite{pea12a}, may evoke contrasting feelings, from rewarding to threatening.
+
+More recently, the smartphone has been considered not strictly related to communication.
+In 2021 A.
+Slavec et al.
+investigated the use of cameras \cite{sla21a} while on travel in locations with a relevant cultural heritage to sustain its preservation and engage the tourists using location-based games, similar to Pokemon Go or geo-caching.
 
@@
 Rodrigues et al.
-publish a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}.
-Although the title evokes a one-way contribution, i.e., the impact of digitalization is assumed to be positive on the quality and the sustainability of touristic offers, in the conclusions the authors reveal awareness for the necessity to address {\em "the preservation of tourism attractions/sites"} and call for for a {\em "a holistic approach ...
+published a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}.
+Although the title evokes a one-way contribution, i.e., the impact of digitalization is assumed to be positive on the quality and the sustainability of touristic offers, in the conclusions the authors reveal the awareness for the need to address {\em "the preservation of tourism attractions/sites"} and call for a {\em "a holistic approach ...
 to support a concept that still lacks conceptual and empirical clarification"}.
 
-In that direction we meet the phenomenon of {\em overtourism}, covered by a significant literature reviewed by Dodds and Butler \cite{dod19a}, which focusses on urban tourism and its social consequences.
-The impact of {\em overtourism} on resorts that trade on their natural resources is investigated in the case of the Hawaii Islands \cite{lin22a} or CostaRica \cite{mat10a} stressing the impact on the social fabric.
-
-The present paper wants to fill the gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined tourism support aspect, taking into account by design its sustainability.
-Once a range of relevant solutions has been identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies feasibility of a specific solution.
+In that direction, we meet the phenomenon of {\em overtourism}, covered by significant literature reviewed by Dodds and Butler \cite{dod19a}, which focuses on urban tourism and its social consequences.
+The impact of {\em overtourism} on resorts that trade on their natural resources is investigated in the case of the Hawaii Islands \cite{lin22a} or Costa Rica \cite{mat10a} stressing the impact on the social fabric.
+
+The present paper wants to fill the gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined aspect of tourism support, taking into account "by design" its sustainability.
+Once a range of relevant solutions has been identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies the feasibility of a specific solution.
 
@@
 Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve further investigation.
-We aim at isolating an issue, proposing a strategy for its solution and implementing a proof of concept for it.
-
-% Other papers examine the topic from the point of view of the user satisfaction, with an accent on the accessibility of the contents from impaired users.
-
-% The use of smartphone is support of a tourist experience is mostly regarded from the point of view of the tourist satisfaction.
-
-% This is the case of
-
-% Cannot depend on communication
-
-\subsection{A simple, low impact solution}
-
-%Starting from previous experiences documented in the bibliography we explore a few broad categories, to find weaknesses and strengths for each of them.
-
-We aim at designing a strategy for a sustainable and effective smartphone-based signage.
+We aim at isolating an issue, proposing a sustainable strategy for its solution, and implementing a proof of concept for it.
+
+\subsection{A simple, low-impact solution}
+
+We aim to design sustainable and effective smartphone-based signage.
 
 A straightforward approach concentrates on the wireless capabilities of such a device.
-Given the premise that the Internet is not reachable, the {\em host} stakeholder may provide a local network of low power radios covering the region of interest.
-Small servers connected to the network might provide specific Web services.
-The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single chip computer (SCC) has a coverage of tens of meters, and a volume in the order of the $cm^3$.
-They are able to provide WiFi Access Points as well as Web contents.
-The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image.
+Given the premise that the Internet is not reachable, the {\em host} stakeholder might provide a local network of low-power radios covering the region of interest.
+Small servers connected to the network would provide specific Web services.
+The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single-chip computer (SCC) has a coverage of tens of meters, and a volume in the order of the $cm^3$.
+They can provide WiFi Access Points as well as Web content.
+The ESP8266 has a capacity of 32KB, which is sufficient for explanatory text and a low-resolution image.
 Other SCCs, like the ESP32, are more powerful but exhibit a higher power consumption.
 
-% One strategy concentrates on the wireless capabilities of the smartphone, although avoiding the coverage of the whole region.
-We consider the implementation of several small servers operating in isolation that provide a few HTML pages.
-The approach requires a modest investment and a marginal environmental impact: a small device based on the ESP8266 single chip computer has a coverage fo tens of meters, and a volume in the order of the $cm^3$.
-They operate as WiFi Access Points, so the user has access to the data in the device by joining the AP and browsing the URL.
-The ESP8266 has a capacity of 32KB, which are sufficient for an explanatory text and a low resolution image.
-
 Power supply dependency is a serious limitation for such kind of solution.
-A radio transmitter is a rather power consuming device, in the range of Watts.
-Even if operation is intermittent, the operation of the transmitter cannot be guaranteed for long periods and the host organization should consider the installation of a power generator, which contributes to the economical cost and environmental footprint of the device.
-
-For this reason we do not consider a solution based on an ad-hoc network as a valid competitor against traditional board-based design for reasons related with poor sustainability.
-
-An alternative consists in the utilization of passive devices, like Near Field trasmitters.
-The transmitting device is flat with the size of a coin and a cost of less than one dollar per piece.
-The receiving smartphone must be very close to receive the content.
-The power needed to operate the radio is drained from the smartphone, so that batteries are not needed.
+A radio transmitter is a rather power-consuming device, in the range of Watts.
+Even if intermittent, the operation of a battery-operated networking device cannot be guaranteed for long periods and the host organization should consider power harvesting, which negatively contributes to the economic cost and environmental footprint of the device.
+
+For this reason, we do not consider a solution based on the deployment of a networking facility as a valid competitor against traditional board-based design.
+As explained, the reasons are related to poor sustainability.
+
+An alternative consists of the utilization of passive devices, like Near Field Communication (NFC) transmitters.
+The transmitting device is flat, the size of a coin, and costs less than one dollar per piece.
+To receive the content the smartphone must be very close to the NFC tag.
+The power needed to operate the radio is drained from the smartphone so that the transmitter does not depend on batteries.
 The NFC device capacity is in the range of the KBytes, nearly a page on this journal.
-Its manufacturing requires a smartphone with a specific app.
+Storing content in an NFC tag requires a smartphone with a specific app.
 
@@
 Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the {\em user's} device as a chink of text which can be treated as such.
-The smartphone can read it aloud to compensate user's inabilities or a translator to cope with linguistic issues without need of an Internet connection.
-
-Another solution in the family of passive devices is the QR-tag.
-Such technique does not require radio communication, but uses the smartphone camera to acquire a graphical code that is translated into text.
-The capacity of a QR-tag depends on the number of dots in the image, which in turn depends on the size of the code and the smartphone camera characteristics.
-We may consider that the capacity of a QR-code roughly equals that of an NFC tag.
-A QR code is larger than an NFC tag of comparable capacity, and manufacturing requires a printer.
-
-% TODO: \usepackage{graphicx} required
+The smartphone can read it aloud, to compensate user's inabilities, or translated, to cope with linguistic issues.
+Such capabilities do not need an Internet connection.
+
+Another solution in the family of passive devices is the QR tag (QR stands for Quick-Response).
+Such a technology does not require radio communication but uses the smartphone camera to acquire a graphical code that is translated into text.
+The capacity of a QR tag depends on the number of dots in the image, which in turn depends on the size of the code and the smartphone camera characteristics.
+We may consider that the capacity of a QR tag roughly equals that of an NFC tag.
+A QR tag is larger than an NFC tag of comparable capacity, and manufacturing requires a printer.
+
 \begin{figure}
@@
 	\includegraphics[width=0.9\linewidth]{figure/qr+NFC+coin}
-	\caption[Passive devices dimensions]{An NFC coin, a QR-code card used in the project, and a 1 Euro coin by comparison on a paper with a square of 1cm}
+	\caption[Passive devices dimensions]{An NFC coin, a QR tag used in the prototype, and a 1 Euro coin by comparison on a paper with a square of 1cm}
 	\label{fig:qrnfccoin}
@@
 
-%\begin{enumerate}
-%	\item Equitable Use
-%	\item Flexibility in Use
-%	\item Simple and Intuitive
-%	\item Perceptible Information
-%	\item Tolerance for Error
-%	\item Low Physical Effort
-%	\item Size and Space for Approach and Use
-%\end{enumerate}
-
-%The labels are self-descriptive, and we address the reader to the discussion in the referenced document for a more detailed explanation.
-%Checking passive devices solutions against the Universal Design principles, we assess their accessibility.
-
 \begin{itemize}
-	\item {\em Equitable use} is tighly related to the smartphone technology, which is itself considered a vehicle for equitability,
-	\item {\em Flexibility in use} is enabled by the computing capabilities of the device, which allows listening to the info instead or reading, translating the information in a different language, or storage for later use,
-	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single touch
-	\item{\em Perceptible information} is a critical requirement, which contrasts with the aim of keeping low the environmental intrusion.
-This point will be further discussed in the section devoted to the implementation
+	\item {\em Equitable use} is closely related to the smartphone technology, which is itself considered a vehicle for equitability,
+	\item {\em Flexibility in use} is enabled by the device capabilities, which allow listening instead of reading, translating the information in a different language, or storing it for later use,
+	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone
+	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation
 	\item{\em Tolerance for error}: there are no margins to use the device in a way that compromises user safety.
@@
 	\item{\em Low physical effort} holds, although the user needs to carry the smartphone
-	\item{Size and Space for approach and use} have to be carefully considered.
-In the case of the NFC the smartphone needs to be nearly in touch with the passive device, while the QR-code must be in line of sight and frameable without effort
+	\item{\em Size and Space for approach and use} need to be carefully considered.
+In the case of the NFC tag the smartphone needs to be nearly in touch with the passive device, while the QR code must be in the line of sight and frameable without effort
 \end{itemize}
 
-Such minimalistic solutions (see figure \ref{fig:qrnfccoin}) compares well with other solutions that make use of the user's smartphone.
-There is a trade-off concerning capacity, but in many circumstances a capacity of 2-300 words is sufficient to convey the description of the site, or provide directions for the the visit.
-If capacity limits are not an issue, a solution based on NFC or QR-codes exhibits a number of advantages:
+Such minimal solutions (see figure \ref{fig:qrnfccoin}) compare well with other more complex ones that make use of the user's smartphone.
+There is a trade-off concerning capacity, but in many circumstances, a capacity of 2-300 words is sufficient to convey the description of the site or provide directions for the visit.
+If capacity limits are not an issue, a solution based on NFC or QR tags exhibits several advantages:
+
 \begin{itemize} 
-\item does not require any power supply
-\item has a limited impact on the landscape
-\item does not entice theft
-\item has a negligible cost
-\item is durable
-\item produces a limited quantity of waste when disposed
-\item content can be stored for usage when the user reaches a zone covered by the Internet
+\item does not require any power supply,
+\item has a limited impact on the landscape,
+\item does not entice theft,
+\item has a negligible cost,
+\item is durable,
+\item produces a limited quantity of waste when disposed of,
+\item content can be stored for later usage; for instance, to visit a URL once the user reaches a zone covered by the Internet.
 \end{itemize}
@@
 \begin{itemize}
-	\item an NFC-tag is smaller than a QR-tag
-	\item writing an NFC-tag requires a smartphone, while the QR-tag needs a printer
-	\item reading an NFC-tag operates at near-to-contact distances, while QR-codes can be read from meters away
+	\item an NFC tag is smaller than a QR tag;
+	\item writing an NFC tag requires a smartphone, while the QR tag needs a printer;
+	\item reading an NFC tag works near-to-contact, while QR tags can be read from a distance;
 \end{itemize}
@@
 
-For a signage application the QR-tag is preferable because a noticeable size is needed. In addition, keeping the tag out of reach prevents vandalisms and misuse.
-
-In conclusion, we have reasons to select a QR-code based solution as a good candidate for a smartphone-based signage.
-We now consider how it it copes with the limits of a traditiona board-based approach (as listed in table \ref{tab:board}):
-
-%The adoption of passive devices compares favorably with traditional boards, considering the strengths and limits as listed above.
+For a signage application, a QR tag is preferable because a noticeable size is needed. In addition, keeping the tag out of reach prevents vandalism and misuse.
+
+In conclusion, we have reasons to select a QR-tag-based solution as a good candidate for smartphone-based signage.
+We now consider how it copes with the limits of a traditional board-based approach (as listed in table \ref{sec:signage}):
 
@@
 	\item logistics: QR-code board can be installed on any sort of pre-existent or natural support
-	\item impact: the board has a minimal interference with the landscape and may easely go un-noticed if not properly advertised
+	\item impact: the board has minimal interference with the landscape and may easily go unnoticed
 	\item accessibility for the visually impaired: the text can be read aloud
-	\item international: the text can be translated
-	\item update: the board can be easely replaced when the content becomes obsolete
-	\item disposal: the card releases a limited quantity of pollutants related with ink support (paper or plastic)
+	\item international: the text can be translated automatically
+	\item update: the board can be easily replaced when the content becomes obsolete
+	\item disposal: the card releases a limited quantity of pollutants related to ink support (paper or plastic)
 \end{itemize}
 
-%One of the issues the host must resolve is the availability of support for electronic devices. The provisioon of such support has a footprint that ha		
-%may be present, but, in the spirit of our investigation, power supply should be address renewable sources, like wind, sunlight and similar.
-%One option discriminates the use of an electronic device on the side of the user. 
-%In case we assume the user does not have a electronic device with her, the 
-%One option is to introduce active electronics: the touristic site administration provides a
-
-There are two relevant trade-offs that the host needs to resolve. One is related with the visibility of the tag. The trade-off is among visuali impact, and visibility.
-
-Il secondo?
+There are two relevant trade-offs that the host needs to resolve.
+One is related to the visibility of the tag.
+The trade-off is between visual impact and visibility.
+The other is related to monitoring tag utilization.
+This may be useful for all sorts of planning activities and is discussed in section \ref{sec:results}.
 
@@
 As anticipated in the introduction, our study wants to provide a proof of concept for a sustainable solution in a concrete setup.
-To comply with such an holistic approach, we need to start from the definition of the operation context.
-
-This section is devoted to the description of the economical and social context where we are going to operate, as well as the historical backgound representing the heritage resource we want to promote.
-Next we will discuss the technical detail of the solution.
-
-The geographical area of interest are the surroundings of Casoli, a small village in a mountaineous region in the north of Tuscany, in Italy.
-The area fall within the municipailty of Bagni di Lucca, in the province of Lucca.
-The archaeological team of our project formerly realized the municipal archaeological map in 2021 \cite{letizia2021}, thus assessing the reasons of interest for the heritage and the geomorphological features of the area.
-
-\subsection{The natural and cultural resources of Casoli and its vicinity}
-
-Bagni di Lucca is located on the north-eastern boundary of the Province of Lucca, in Val di Lima, and is part of the Media Valle del Serchio district.
-With its mountains it marks the historical border with the Modena and Pistoia area, it is very rich in potential for its naturalistic, archaeological and historical heritage, both expressed and as yet unexpressed.
-Thanks to the archaeological map the main sites of interest from prehistoric to contemporary times in this municipality are catalogued, photographed and georeferenced.
-
-From a historical-geographical point of view, this area is identified with the Lima stream and its tributaries, that impress to the area a particular geomorphology, very impervious despite the modest hilly elevations.
-
-The Lima stream has characterised the history of Bagni di Lucca since ever: some of the oldest evidence of human presence in the valley has
-been found along its ancient river terraces, such as caves and rock shelters frequented since the Palaeolithic age; the
-manufacturing industry (paper mill, flour mills and, in recent times, energy production) have exploited the since the Middle Age until the 1980s \cite{men76, bed05, ser21}.
-
-On the naturalistic side, the region of Bagni di Lucca encompasses an incredible concentration of biodiversity, counting no less than three sites of the Natura 2000 Network \cite{natura2000} European Economic Community (EEC) initiative.
-
-There are three SCIs-SACs (Sites of Community Interest and Special Areas of Conservation) located in the area north of the Lima stream, corresponding to the Apennine portion, covering 23\% of the municipal surface:
+To comply with such a holistic approach, we need to start with the definition of the operation framework.
+
+This section is devoted to the description of an economic and social context, as well as the historical background representing the heritage resource we want to promote.
+Next, we will discuss the technical details of the solution.
+
+The geographical area of interest is the surroundings of \emph{Casoli}, a small village in a mountainous region in the north of Tuscany, in Italy.
+The area falls within the municipality of \emph{Bagni di Lucca}, in the province of \emph{Lucca}.
+In 2021, the archaeological team of our project realized an archaeological map assessing the reasons of interest for the heritage and the geomorphological features of the area.
+
+\subsection{The natural and cultural resources of \emph{Casoli} and its vicinity}
+
+\emph{Bagni di Lucca} is located on the north-eastern boundary of the Province of \emph{Lucca}, in \emph{Val di Lima}, and is part of the \textit{Media Valle del Serchio} district.
+With its mountains it marks the historical border with the Modena and Pistoia area, it is very rich in potential for its naturalistic, archaeological, and historical heritage, both expressed and as yet unexpressed.
+Thanks to the archaeological map the main sites of interest from prehistoric to contemporary times in this municipality are cataloged, photographed, and georeferenced.
+
+From a historical-geographical point of view, this area is identified with the \emph{Lima} stream and its tributaries, which impress to the area a particular geomorphology, very impervious despite the modest hilly elevations.
+
+The \emph{Lima} stream has characterized the history of \emph{Bagni di Lucca} since ever: some of the oldest evidence of human presence in the valley has been found along its ancient river terraces, such as caves and rock shelters frequented since the Palaeolithic age; the manufacturing industry (paper mill, flour mills and, in recent times, energy production) have exploited since the Middle Age until the 1980s \cite{men76, bed05, ser21}.
+
+On the naturalistic side, the region of \emph{Bagni di Lucca} encompasses an incredible concentration of biodiversity, counting no less than three sites of the Natura 2000 Network \cite{nat00} European Economic Community (EEC) initiative.
+
+There are three SCIs-SACs (Sites of Community Interest and Special Areas of Conservation) located in the area north of the \emph{Lima} stream, corresponding to the Apennine portion, covering 23\% of the municipal surface:
 \begin {itemize}
-\item the limestone areas of Val di Lima and Balzo Nero; 
+\item the limestone areas of \emph{Val di Lima} and Balzo Nero; 
 \item Monte Prato Fiorito-Monte Coronato-Valle dello
@@
 
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=0.3\linewidth]{figure/toscana-casoli}
+%\end{figure}
+
 \begin{figure}
-	\centering
-	\includegraphics[width=0.3\linewidth]{figure/toscana-casoli}
-	\caption[The region of Casoli]{}
+\subfloat[Italy and Tuscany, bordered in red]{
+	\includegraphics[width = 0.3\linewidth]{figure/italia.png}
+	\label{fig:population}
+}
+\hfill
+\subfloat[Tuscany and the municipality of \emph{Bagni di Lucca} (in yellow)]{
+	\includegraphics[width = 0.3 \linewidth]{figure/toscana.png}
+	\label{fig:tourismflows}
+}
+\hfill
+\subfloat[Location of the \emph{Casoli}  village in the \emph{Bagni di Lucca} municipality, bordered in yellow.
+The small town of \emph{Bagni di Lucca} is visible located a few kilometers west of \emph{Casoli} ]{
+	\includegraphics[width = 0.3 \linewidth]{figure/bagni_di_lucca.png}
+	\label{fig:tourismflows}
+}
+\caption[The region of \emph{Casoli}]{The region of \emph{Casoli}}
 \label{fig:toscana-casoli}
@@
 
-The latter is also a SPA (Special Protection Area) and the Orrido di Botri State Reserve is located within it \cite{nat00}.
+The latter is also a SPA (Special Protection Area) and the \textit{Orrido di Botri} State Reserve is located within it \cite{nat00}.
 It is therefore a natural heritage with a fragile balance that needs to be preserved.
 
-In recent years, the main tourist attractions have concentrated on the Lima stream, with some associations and private entities promoting outdoor experiences, particularly fluvial sports, such as canyoning, rafting and sup.
-Others important tourist attraction are trekking and hiking, supported by a network of path tracked by local CAI (Club Alpino Italiano).
-
-In recent years the community opened two entirely new trails:
+In recent years, the main tourist attractions have concentrated on the \emph{Lima} stream, with some associations and private entities promoting outdoor experiences, particularly fluvial sports, such as canyoning, rafting, and sup.
+Other important tourist attractions are trekking and hiking, supported by a network of paths tracked by local CAI (\textit{Club Alpino Italiano}).
+
+In the last few years the community opened two entirely new trails:
 \begin{itemize}
-	\item the Alta Via dei Pastori (2019), a ring around Monte Prato Fiorito that takes up the ancient grazing route, and
-	\item t the Sentiero degli Avi (2020), a ring that from Montefegatesi
-	reaches Monte Coronato \cite{pin21}
+	\item the \textit{Alta Via dei Pastori} (2019), a ring around \textit{Monte Prato Fiorito} that takes up the ancient grazing route, and
+	\item the \textit{Sentiero degli Avi} (2020), a ring that from \textit{Montefegatesi}
+	reaches \textit{Monte Coronato} \cite{pin21}
 \end{itemize}
 
-Another recently developed project (2019) is the expansion of the Saint Bartholomew's Path, which runs through the territory of Pistoia, with a
-variant that from Popiglio continues in five stages in the municipality of Bagni di Lucca to Pieve di Controne, acting as the 'Lucca gateway' to the Path \cite{camsb}.
-
-Such initiatives led to a considerable revival of interest among Italian and international hikers for the area, especially for the Apennine side of the valley.
-
-We see the opportunity to improve the valorisation and enjoyment of this area by applying a slow tourism approach and involving the community, especially in the southern part of the municipality, which is still little frequented, more hilly and therefore less travelled by the trail network.
-We propose the creation of geo-itineraries characterised by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.
-
-\subsubsection{An historical perspective of an italian mountain site}
-
- The hill village of Casoli is located south of the Lima stream.
-The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age and used also as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.
+Another recently developed project (2019) is the expansion of the Saint Bartholomew's Path, which runs through the territory of \textit{Pistoia}, with a
+variant that from \textit{Popiglio} continues in five stages in the municipality of \emph{Bagni di Lucca} to \textit{Pieve di Controne}, acting as the '\emph{Lucca} gateway' to the Path \cite{camsb}.
+
+Such initiatives led to a considerable revival of interest among Italian and international hikers in the area, especially for the Apennine side of the valley.
+
+The way to improve the valorization and enjoyment of this area applies to a slow tourism approach.
+This approach can involve the community, especially in the southern part of the municipality, which is still less frequented, more hilly, and therefore less traveled by the network of trails.
+We propose the creation of geo-itineraries characterized by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.
+
+\subsubsection{An historical perspective of an Italian mountain site}
+
+ The village of \emph{Casoli} is located south of the \emph{Lima} stream on a hill named {\em "Tanette"}.
+In the local small lair, the name reveals the presence of karstic cavities, some inhabited between the Paleolithic and the Iron Age and used also as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.
 
@@
 prove that Ligurian populations occupied the region scattered and in small nuclei.
-The dedication of the Latin colony of Lucca in 180 B.C.
+The dedication of the Latin colony of \textit{Lucca} in 180 B.C.
 marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.
 
-We have little evidence of roman settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of Buca La Piella, investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside.
+We have little evidence of Roman settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of \textit{Buca La Piella}, investigated in 1975 by the Centre for Archaeological Studies of \emph{Lucca}; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside.
 In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found \cite{gia96, men81, cia03}.
 
-In Longobard and Carolingian times, Lucca's Val di Lima was one of the three administrative districts into which the mountains were divided and was called “fines Contronenses” \cite{qui02, cia06, cia11}.
-The only find from this period is the Grotta di Arzale, a rock shelter that opens up north-west of Casoli \cite{gia96}.
-
-The first attestation of the settlements of Casoli and Lacu (Lake) dates back to the 10th century \cite{bar41, gia96}.
-Most documents of this period show the fractioning and alienation of ecclesiastical heritage in favour of the city aristocracy, securing an accumulation of funds and power that was to form the basis of the subsequent domains \cite{qui02, gia96, for12, for15}.
-
-Written sources mention a castle in Casoli from 1180, but its foundation must be earlier.
-Between the 13th and 15th centuries, the fortification was at the centre, along with the other castles of the Val di Lima, of clashes between Lucca, Pisa and Florence, with alternating fortunes, as it was a strategic border area for the power of Lucca.
+In Longobard and Carolingian times, \emph{Lucca}'s \emph{Val di Lima} was one of the three administrative districts into which the mountains were divided and was called “fines Contronenses” \cite{qui02, cia06, cia11}.
+The only find from this period is the \textit{Grotta di Arzale}, a rock shelter that opens up northwest of \emph{Casoli} \cite{gia96}.
+
+The first attestation of the settlements of \emph{Casoli} and \emph{Lacu} (Lake) dates back to the 10th century \cite{gia96}.
+Most documents of this period show the fractioning and alienation of ecclesiastical heritage in favor of the city aristocracy, securing an accumulation of funds and power that was to form the basis of the subsequent domains \cite{qui02, gia96, for12, for15}.
+
+Written sources mention a castle in \emph{Casoli} from 1180, but its foundation must be earlier.
+Between the 13th and 15th centuries, the fortification was at the center, along with the other castles of the \emph{Val di Lima}, of clashes between \emph{Lucca}, \textit{Pisa} and Florence, with alternating fortunes, as it was a strategic border area for the power of \emph{Lucca}.
 Today, very little is preserved and the area has been partly inhabited \cite{gia96, for12, rom16}.
 
-The same document from 1180 mentions the church of San Donato, located in the main square of the village of Casoli.
+The same document from 1180 mentions the church of \textit{San Donato}, located in the main square of the village of \emph{Casoli}.
 The current structure dates back to the 12th-13th centuries, with subsequent renovations.
-Initially dedicated only to San Donato, it acquired the double dedication after the abandonment of the church of Sant'Andrea de Lacu.
-The latter is located on a plateau to the east of Lake Casoli and is attested from 1260 \cite{ber18, gia96, cap17}, but already in 15th century we learn of its state of abandonment \cite{gia96, con12}.
-This was probably due to the depopulation of the lake area and the simultaneous strengthening of the town of Casoli, which was fortified and better protected during this unstable phase.
-
-Today, the romanesque church of Sant'Andrea de Lacu is in a state of decay, with a rectangular plan, ending on the east side with a semi-circular apse.
-Inside, there is a reused element of the previous building, testifying to an older origin, probably contemporary with the ancient settlement of Lacu (Lake) mentioned in sources from the 10th century and no longer visible today.
+Initially dedicated only to \textit{San Donato}, it acquired a double dedication after the abandonment of the church of \textit{Sant'Andrea de Lacu}.
+The latter is located on a plateau to the east of Lake \emph{Casoli} and is attested from 1260 \cite{ber18, gia96, cap17}, but already in the 15th century, we learn of its state of abandonment \cite{gia96, con12}.
+This was probably due to the depopulation of the lake area and the simultaneous strengthening of the town of \emph{Casoli}, which was fortified and better protected during this unstable phase.
+
+Today, the Romanesque church of \textit{Sant'Andrea de Lacu} is in a state of decay, with a rectangular plan, ending on the east side with a semi-circular apse.
+Inside, there is a reused element of the previous building, testifying to an older origin, probably contemporary with the ancient settlement of \textit{Lacu} (Lake) mentioned in sources from the 10th century and no longer visible today.
 
 In the early modern age, villages of the area experienced a progressive architectural renewal, which still largely characterizes the settlements today.
-Within Casoli, a series of residential buildings with imposing dated portals are preserved, some with courtyards, and mansions that denote a discrete deployment of resources by wealthy social classes.
-Also dating from the modern era is the Madonna di Castello Oratory, characterised by an entrance eclosure and located along the cobbled road that traces the ancient route between the village and the summit area of the medieval castle \cite{gia96}.
-
-Along the road that led toward Lucchio crossing the area of Casoli Lake there are the chapel of Madonna di Col di Piano and two “metati” referable to the contemporary age, used for drying chestnuts.
+Within \emph{Casoli}, a series of residential buildings with imposing dated portals are preserved, some with courtyards, and mansions that denote a discrete deployment of resources by wealthy social classes.
+Also dating from the modern era is the \textit{Madonna di Castello} Oratory, characterized by an entrance enclosure and located along the cobbled road that traces the ancient route between the village and the summit area of the medieval castle \cite{gia96}.
+
+Along the road that led toward \textit{Lucchio} crossing the area of \emph{Casoli} Lake, there is the chapel of \textit{Madonna di Col di Piano} and two “\textit{metati}”, buildings destined for chestnuts drying, referable to the contemporary age.
 Referred to as the "bread tree," chestnut fruits and the flour derived from them were a staple of the mountain diet until the mid-1900s \cite{buc92, puc10}.
 
-\subsubsection{The socio-economical context of Casoli and its vicinity}
-
-......
+\subsection{An integrated view of tourism in the \emph{Casoli} region}
+
+%\subsubsection{Study of the tourist context for governance approaches}
+
+Within the paper's aim framework and from a touristic point of view, the relevance of the touristic context of the territory is crucial in understanding the socio-cultural system of potential tourist destinations.
+In light of this and for a holistic study approach to social features of tourism, the host and traveler relationship is directly related to local development and local government systems \cite{amo21}.
+
+In this study, the holistic perception of local stakeholders and community empowerment (with its increasing potential) in tourism at the local level acquires a particular value.
+For this specific purpose, this paragraph is well related to the tourism dynamics of the Municipality of \emph{Bagni di Lucca}, the administrative land hosting \emph{Casoli}.
+
+\subsubsection{The tourism context of / The tourist dynamics in \emph{Bagni di Lucca}, the \emph{Val di Lima} and \emph{Casoli}}
+
+In Italy, a municipality is the smallest geographical-administrative unit offering accessible data about tourism, which are the fundamentals of our research.
+The municipality of \emph{Bagni di Lucca} boasts historical popularity for its landscapes told by famous poets (such as Giosuè Carducci, Giovanni Pascoli, Eugenio Montale, George Gordon Byron, Percy Bysshe Shelley) and celebrities like Paolina Bonaparte.
+The typical Liberty style architecture characterizes the villas like the \textit{Real Casinò} featuring elegant gardens representing the power of the Republic of \emph{Lucca} in the early XIX century and the \emph{Belle Epoque}.
+Such historical prominence made the town a well-known destination abroad, thanks to the massive emigration and the presence of the Thermal Bath structure \footnote{Terme di Bagni di Lucca: \url{https://www.termebagnidilucca.it/}}, nearby must-see attractions of the time.
+
+\emph{Bagni di Lucca} owes part of its charm to the river \emph{Lima}, which runs along the city before flowing into the \textit{Serchio} River.
+The \emph{Lima} Valley (\emph {Val di Lima} in the local idiom) is a touristic basin set in the \textit{Serchio} Valley, dotted with medieval and Roman villages with historical remains such as underground caves.
+The \emph{Val di Lima} area is well-known for its environmental beauties including rupestrian and lake ecosystems, and offers a wide range of tourist attractions related to sports, from water sports, such as canyoning and rafting, to trekking in water and land, climbing, mountain biking, and horseback riding.
+Overall, the \emph{Val di Lima} has the capabilities to promote a very identifying destination brand straddling the town of Lucca and the mountain area of the \textit{Garfagnana}.
+
+The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like the \emph{Orrido di Botri} with the \emph{Canyon Park\footnote{\url{ https://www.canyonpark.it/}}}, and other service companies and associations for experiential tourism.
+
+The demographic trend enlightens a relevant aspect of \emph{Bagni di Lucca} society, and the employment statistics of the population mirror the social, cultural, and economic aspects of its touristic ecosystem.
+
+\begin{figure}
+\subfloat[Population trend (2001-2022)]{
+	\includegraphics[width = 0.5\linewidth]{figure/population.png}
+	\label{fig:population}
+}
+\hfill
+\subfloat[Tourism flows and population (2017-2022)]{
+	\includegraphics[width = 0.5 \linewidth]{figure/tourism.png}
+	\label{fig:tourismflows}
+}
+\caption{Demography and tourist flows in the municipality of \emph{Bagni di Lucca}}
+\end{figure}
+
+%\begin{figure}
+%    \centering
+%    \includegraphics[width=0.75\linewidth]{figure/Population.png}
+%    \caption{\emph{Bagni di Lucca}’s population trend, 2001-2022}
+%    \label{fig:population}
+%\end{figure}
+
+\emph{Bagni di Lucca} stretches on 164.70 $km^2$ area with a population density \footnote{Index referring to the number of people living in a territory per square kilometer.
+It is the rate between the annual number of residents and the surface area.
+In our case and for 2022, the density is 5593 people over 164.70 $km^2$.} of 3395 inhabitants per square kilometer and a job occupancy rate of 33.53\%.
+Regarding human growth at the territorial level, it is relevant to consider the \emph{Bagni di Lucca} population trend during the 2001-2022 period.
+The curve shown by figure \ref{fig:population} registers a constant and slight decrease of one thousand people in a twenty-year time frame (2001: 6556 people; 2022: 5593 people).
+
+During the latest decades, numerous touristic operators have started their touristic activities by taking advantage of the climatic and morphological characteristics of the territory.
+From a quantitative perspective, it is relevant to understand the tourist flows during a 5-year time frame; the chart in figure \ref{fig:tourismflows} highlights a decreasing trend from 2017 to 2021 by considering the pandemic breakdown impacts on tourism.
+
+%\begin{figure}
+%    \centering
+%    \includegraphics[width=0.75\linewidth]{figure/TourismFlows.png}
+%    \caption{\emph{Bagni di Lucca's} tourism flows, 2017-2022}
+%    \label{fig:tourismflows}
+%\end{figure}
+
+The Tourism Density Index \footnote{The Tourism Density Index shows the tourism concentration in the higher touristic season.
+It is the rate between the number of tourists and the surface area.} is 106.41 tourists per $km^2$ in 2022.
+The arrivals exhibit a drop from 2019 (from  4339 in 2017 to 2363 in 2021), and touristic presences have suffered a slight decrease during the same period, as shown in figure \ref{fig:tourismflows}, while a further increase has been registered in 2022 (5.876 arrivals).
+
+The touristic area of \emph{Val di Lima} features a tourist area belonging to \emph{Bagni di Lucca} Municipality, the second touristic area after \textit{Barga}, in the \textit{Media Valle del Serchio Area}.
+
+%\begin{table}
+%    \footnotesize
+%    \setstretch{1,4}
+%    \centering
+%    \begin{tabular}{|c|c|c|c|c|} \hline 
+%         Region & Touristic Density & Average stay &  Accommodation density & Coexistence Index\\
+%           & (visitors/{$km^2$}) & (days) & (accommodations/{$km^2$}) & (100 $\times$ foreign/italian)\\ \hline
+%         Bagni di Lucca & 106,41 &  2,98 &  10,14 & 103,74\%\\ \hline
+%         Lucca & 1866,96 & 3,36 & 19,30 & 69,87\%\\ \hline
+%    \end{tabular}
+%    \caption{Tourist numbers in \emph{Bagni di Lucca} in 2022\tablefootnote{Each  index gives a different type of touristic information about \emph{Bagni di Lucca} Municipality: the touristic density represents the number of tourists per $km^2$; the Average stay index is the average number of nights spent in town by tourists; the Accommodation Density Index is related to the number of touristic presences on the number of beds occupied by tourists all over the year; the coexistence index shows the relevance of foreign tourists per 100 Italian tourists.}}
+%    \label{tab:tourism}
+%\end{table}
+
+\begin{table}
+    \footnotesize
+    \setstretch{1,2}
+    \centering
+    \begin{tabular}{l||l|l} \hline 
+        {\bf Region} & {\bf Bagni di Lucca} & {\bf Lucca} \\ \hline
+        Resident density (residents/{$km^2$}) & 33.95 & 215.71 \\
+        Touristic density (visitors/{$km^2$}) & 106.41 & 1866.96 \\
+        Average stay  (days) &  2.98 &  3.36 \\
+        Accommodation density (accommodations/{$km^2$}) & 10.14 & 19.30 \\
+        Coexistence Index (100 $\times$ Foreign/Italian) & 103.74\% &  69.87\% \\ 
+         \hline
+    \end{tabular}
+    \caption{Tourist numbers in \emph{Bagni di Lucca} in 2022\tablefootnote{Each index gives a different type of touristic information about \emph{Bagni di Lucca} Municipality: the touristic density represents the number of tourists per $km^2$; the Average stay index is the average number of nights spent in town by tourists; the Accommodation Density Index is related to the number of touristic presences on the number of beds occupied by tourists all over the year; the coexistence index shows the relevance of foreign tourists per 100 Italian tourists.}}
+    \label{tab:tourism}
+\end{table}
+
+Table \ref{tab:tourism} provides statistical data concerning tourist flows and accommodation in \emph{Bagni di Lucca}.
+Notably, the tourist density of 106.41 tourists per $km^2$ is higher than the population density (33.95 inhabitants per $km^2$).
+
+The number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay a number of nights more than a weekend and less then a week.
+On the other hands, this data means that people like staying in this touristic comprehensory and they may have founded attractions, services and activities they need.
+Consequently, a well-organized turistic system attracts potentially touristic presences widespread on the municipality territory.
+
+The coexistence index describes the distribution of tourist nationalities: 103.74 foreign tourists per 100 Italians in \emph{Bagni di Lucca} indicates a significant share of foreign tourists.
+
+For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study with the purpose to highlight the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}.
+As concerns tourist indexes shown in Chart 3, the touristic density counts 1866, 96 people per square kilometer (1773 kmq), with an average stay of 3,36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2,98 nights).
+The provincial area of \emph{Lucca} shows an accommodation density of 19,30 on 10, 14 of \emph{Bagni di Lucca}, while the coexistence index of \emph{Lucca} (with 69,87 foreign tourists on 100 Italian tourists) is lower than \emph{Bagni di Lucca}’s coexistence index (with 103,74 foreign tourists in 100 Italian tourists).
+Definitely, it can be claimed that \emph{Bagni di Lucca} is well positioned in the whole touristic context of \emph{Lucca} since it represents one of the less populated municipalities in the territory with interesting tourist indexes.
+In fact, \emph{Bagni di Lucca} boasts its historical thermal tourism tradition together with a well-equipped environment for sport and adventure tourism, on a surface of 164,70 kmq, that is one of the biggest municipality areas in the province of \emph{Lucca}.
+
+Summing up, the quantitative data give evidence of a vibrant and attractive tourist reality constantly attracting tourists during the last decade.
+Thanks to such tourist flows \emph{Bagni di Lucca} is considered one of the main tourist areas of the \emph{Val di Lima} and the Serchio Valley.
+Several promotional websites advertise \emph{Bagni di Lucca} attractions\footnote{Among local websites, these two destination websites are: \url{www.turismobagnidilucca.com} and \url{www.valdilima.org}}: the sitography closing this paper contains a long but incomplete list that confirms such a statement, and Table \ref{tab:experience} exposes the promotional touristic system offering varied experiences in the \emph{Val di Lima} area.
+The variety of outdoor touristic services dates back to the ‘90s when some associations began organizing mainly trekking excursions and rafting experiences.
+
+From the point of view of our research, the context of the \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well touristic services).
+
+After a detailed analysis of the touristic and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities  \cite{loz09}.
+
+%Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of Lucca and the Garfagnana mountain area.
+The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like Orrido di Botri, with The Canyon Park \footnote{\url{ https://www.canyonpark.it/}}, and other service companies and associations for experiential tourism.
+
+\begin{table}
+    \centering
+    \begin{tabular}{|p{70mm}||p{45mm}|} \hline 
+         {\bf Touristic promotion }&  {\bf Kind of service}\\
+         \hline \hline 
+         Aguarajà River Experience \newline \url{https://aguaraja.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
+         \hline 
+         Rafting H2O \newline \url{https://www.raftingh2o.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
+         \hline 
+         RocKonda \newline \url{https://www.rockonda.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ \hline 
+         A.S.D. Garfagnana Rafting \newline \url{https://www.garfagnanarafting.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ 
+        \hline 
+         Val di Lima Off-Road \newline \url{https://www.valdilimaoffroad.com} & Quad, enduro\\ \hline 
+         Canyon Park \newline \url{https://www.canyonpark.it/canyon-sup}& Canyoning, aqua-trekking, stand-up padding, mental care and yoga, eco laboratory\\
+         \hline 
+         Pro Loco Bagni di Lucca \newline  \url{https://www.turismobagnidilucca.com/} \newline \url{https://www.valdilima.org/}& Sportfishing, mental care, trekking, survival, paragliding, horse riding, diving\\
+         \hline 
+         eBike Adventure Tour \newline \url{http://www.ebikeadventuretour.it/}& Ebike, Mountain bike\\ \hline 
+         Val di Lima eBike \newline  \url{https://valdilimaebike.it/} & Mountain e-bike, Quad, Enduro, Paintball\\
+         \hline 
+        Terme Bagni di Lucca \newline \url{https://www.termebagnidilucca.it/} &Spa, beauty and wellness experience\\
+        \hline
+    \end{tabular}
+    \caption{\textbf{Tourism experiences’ offer in \emph{Val di Lima} area \tablefootnote{The list might not include all private stakeholders of the territory, since some touristic operators might have an Internet presence and others might operate as public-private players.
+The companies and associations in the list have been individually contacted but some did not respond to our survey on touristic information.}}}
+    \label{tab:experience}
+\end{table}
+
+The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality.
+It is 7,22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters of altitude.
+
+\emph{Casoli} is a fascinating village settled on the green and rolling hill landscapes crossed by the river \emph{Lima}, in the \emph{Media Valle del Serchio}.
+Its territory features multiple kinds of touristic sights: from soft and hard outdoor experiential attractions (specifically the Canyon park activities like canyoning, rafting, and trekking experiences)) to Romanesque churches’ remains and huts.
+Qualitative studies oriented to tourist attractions, and widely, to the complexity of the \emph{semi-structured} tourism show that \emph{Casoli} exhibits unique features despite its tourism variety.
+
+There aren’t touristic information at the administrative level of \emph{Casoli} , so that the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT.
+
+Given the context and the local vocation, we conclude that \emph{Casoli} is a small destination for tourists practicing heterogeneous touristic circuits in individual and combined ways.
+According to qualitative and quantitative analysis \cite{loz09} on the touristic and social context of \emph{Bagni di Lucca}, as the municipal area encompassing \emph{Casoli} village, the mentioned area boasts a good touristic appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities.
+Local stakeholders operating in the tourism system offer a variety of services all year round.
+Whenever possible the offer is enriched with original touristic approaches paying attention to sustainability.
+As part of the \emph{Bagni di Lucca}’s ecosystem with its touristic vocation, \emph{Casoli} is defined as a well-known destination.
+Given that \emph{Casoli} is a target of semi-structured tourism, and to preserve its peculiar social fabric, sustainability of development actions is of paramount importance.
+The latter aspect suggests paying attention to environmental conservation, cultural heritage conservation, intergenerational continuity, and economic equity \cite{kuh10}.
+
+\subsubsection{Guidelines for a sustainable development}
+
+In \cite{dre22} Dianne Dredge describes the role of private and public entities in making tourism sustainable, which we summarize in the following actions:
+
+%In this perspective, policy-makers play a key role in the planning of sustainable measures and in the monitoring of impacts; in particular, in the context of tourism, sustainability is a topic discussed in its many aspects, including resident ecosystem resilience approaches and regeneration actions by policy-makers and/or private tour operators \cite{dre22}.
+
+%In light of the above analysis, the following actions from public and private entities might foster sustainable development:
+
+\begin{itemize}
+
+\item Public actions: 
+\begin{itemize}
+\item to foster an inclusive stakeholder approach for the tourism ecosystem 
+\item to promote a community-involved tourism vision 
+\item to monitor and analyze tourism flows 
+\item to plan a \emph{slow tourism} development strategy to achieve environmental, cultural, and socioeconomic sustainability objectives 
+\item to develop digitization tools and strategies for culture and tourism fruition
+\end{itemize}
+
+\item Professional actions: 
+\begin{itemize}
+\item to renovate experiential tourism with a sustainable approach 
+\item to develop potential touristic areas with unexpressed tourism appeal 
+\item to diversify tourism offer based on tourist provenance and service preferences 
+\item to monitor tourism flows and tourist behaviors 
+\item to foster a public-private collaboration approach to use public funds for tourism and cultural projects toward sustainable objectives 
+\end{itemize}
+
+\end{itemize}
+
+The above-mentioned tourism and culture measures should be included in a wider and holistic planning vision for \emph{Bagni di Lucca} as a heterogeneous cultural and sustainable destination.
+The combination of private and public interest as a long-term strategy for a comprehensive tourism approach needs a bottom-up vision involving residents and local operators in tourism and tourism-related sectors \cite{lem20, lem22}.
+
+In this way, community-involved tourism can be a promising solution for sparsely populated areas with significant tourist attractions, such as the village of Casoli.
+This type of tourism involves active participation and entrepreneurship from the local community to promote self-employment, community management, and stakeholder decision-making processes \cite{nag15}.
+
+In order to create a successful strategic plan, it is crucial to have a thorough understanding of the destination's morphology, environment, history, culture, society, and economy, with sustainability being the key value.
+Policymakers should recognize such needs and act in this way, both politically and socially as described by Beatrix et al.
+in \cite{bea10}.
+As concerns the empiric case of \emph{Casoli} village, within the wider context of \emph{Bagni di Lucca}, tourism is a key driver for re-population and re-qualification strategies, mainly where historical and archaeological sight can be promoted with natural attractions.
+
+In such a context digitalization plays a primary role in innovation actions, allowing effective monitoring of tourism dynamics, improving the touristic experience with geo-localization tools, and providing the tools to design and operate tourism initiatives \cite{lem22}.
+
+\subsubsection{Our activity}
+
+We include three stakeholder communities in the evaluation of the response to the initiative.
+We do not speak about {\em success} since the initiative is deliberately soft, and an outcome is expected in years, and the definition of the evaluation criteria rises new questions.
+So we are content to see if the stakeholders regarded the initiative as an intrusion, as helpful, or simply neutral.
+
+We considered four stakeholder communities:
+
+\begin{itemize}
+	\item {\em the residents} the local community that currently inhabits the site
+	\item {\em entrepreneurs} who currently have a business on the site
+	\item {\em administrators} that are in charge of managing the site resources and that, at due time, will respond to the two stakeholders above, and finally
+	\item {\em the users}, those that come to \emph{Casoli} to visit the Cave of La Piella and that we find the QR-codes on their way
+\end{itemize}
+
+The category of the users escapes an evaluation since the QR tag does not record reading operations.
+We added to the text message in the QR-tag an external link, that we monitor for the number of hits on pages dedicated to the specific tag: however, this feedback largely underestimates the number of times the tag is read.
+
+Considering that each stakeholder category has specific concerns that need to be addressed, our focus has been on understanding the unique characteristics of each category, including their needs and expectations for tourism and environment development.
+
+As concerns residents, we had the opportunity to meet only two people living in \emph{Casoli}  and providing us a few information which was not useful for our study aims; local entrepreneurs we met informed us about the trekking tourism flows passing through \emph{Casoli}  and the Piella Cave path, which usually starts from the Canyon Park experience.
+
+The administrators of \emph{Bagni di Lucca} Municipality were highly involved in addressing the study's concerns and provided valuable information regarding archaeological and cultural aspects that align with our study goals, namely to enhance \emph{Casoli}  area as a sustainable area for hi-tech oriented tourists.
+
+In light of the above assumptions, a reflection on community-involved tourism perspective is needed.
+As shown by this empirical study, a small village like \emph{Casoli}  can require a valorization strategy taking advantage of the wider local tourism context.
+
+According to a holistic development vision, when a tourist destination meets its community issues it means that local stakeholders must balance a tourist industry vision with sustanability goals at a comprehensive perspective on environment, socio-cultural and economic long-term benefits, as described by George B.
+Et Alii, ()
+
+Indeed,  a multi-stakeholder approach could represent a valuable tool for “minor tourism” and well-established tourist destinations.
+
+Literature on this scientific focus, as described by Richards G and Derek H. () gives evidence on local stakeholders' role as ‘designers’ of their living territory. 
+This is particularly important in inland and marginal tourist areas that require a sustainable approach to development, rather than relying on mass tourism.
+
+A systemic vision is necessary to manage these areas effectively.
+With these concerns, inner area governance needs a sustainable-led community and policy approach considering socio-cultural local aims for tourism system implementation and environment development.
+In these terms, local awareness is fundamental for private and public stakeholders enabling sustainable decision-making processes.
+
+On the pragmatic hand, the scientific aim of our territorial exploration is to promote cultural and tourist points of interest with high tourist value. 
+In this regard, we have analyzed the potentiality of involving local administrations and professional tourism operators to create new tourist itineraries by valorizing existing hiking trails.
+Firstly, the Municipal Structural Plan should include an archaeological survey; secondly, local tourism stakeholders will benefit from a destination management action to promote environmental and socio-cultural sustainability.
+
+Definitely, community-involved tourism may represent the balanced intersection between community and development through policymakers' actions and decision makings (Jamal T. and Dredge, 2014). 
+In this view, the socio-political framework of tourism, involving governance measures, has to include environmental requalifying and destination planning (Sharabani et alii, 2020; Richards and Hall, 2003).
+
+Policy measures for sustainable and resilient tourism should focus on culture-led and community empowerment activities, such as preserving the history and memory of the place, promoting slow tourism and cultural innovation, as well as implementing digitalization activities.
 
@@
 
-%The study of the Casoli area made it possible to identify the historical-archaeological sites considered of greatest interest for the purposes of a geotouristic route.
-Once these sites had been located and the possible paths to reach them analysed, also by consulting the online sharing of itineraries by hikers and climbers, we carried out field surveys between April and June 2023 to check the status of the trails and sites for their enjoyment.
-(questo penso di metterlo dopo)
-
-As explained in the previous section, the area around Casoli has a long history witnessed by architectural and geological features: it is therefore a good candidate for our case study.
-In the rest of this section we define the features of interest, and report about their accessibility.
-
-%The site at the centre of the project's interest is Buca La Piella, a cave located along the Solco del Monte and off the conventional circuits, which is why it is less frequented and consequently less compromised for the scientific analyses that will be carried out there, compared to the caves located along the Lima stream, near the state highway.
-The first part of the route coincides for about 500 m with CAI path 386, which together with path 388 forms the Casoli-Lucchio ring around Monte Memoriante.
-Access to the path, which is poorly marked with a faded red and white simple signpost, is along the asphalt road that leads from Ponte Maggio to Casoli and starts from the slope with a short stone ramp.
-After a short climb, one walks at altitudealong the slope and then enters the forest, skirting the final part of the Solco del Monte, known as Rio di Castello.
+As explained in the previous section, the area around \emph{Casoli} is a good candidate for our case study because of its long history, which is witnessed by architectural and geological features.
+In the following, we define the relevant features and report on their accessibility.
 
@@
 It is located in a fascinating environment outside conventional circuits, which is why it is less frequented and consequently less compromised.
-It is reachable leaving a marked path to follow the bed of the Lima river.
-The way is dotted with minor karsic caves, and the {\em Buca La Piella} is reachable climbing up on a steep slope in the wood on the right of the Lima river.
-
-%A faint track through the grass diverts from the path to climb up the side of the slope to the rock face where one of the many karstic formations that characterise the area opens up.
-This is a small cave, called the 'Antro dell'Ugola' in the Tuscan speleological register.
-It consists of two openings whose entrance has a conformation reminiscent of a uvula.
-There are no proven finds, but the space inside and the position could indicate a frequentation that will be verified with the comparison analyses that we will carry out.
+It is reachable leaving a marked path to follow the bed of a tributary of the \emph{Lima} stream.
+The trail is dotted with minor karst caves, and the {\em Buca La Piella} is reachable by climbing up on a steep slope in the wood on the left of the watercourse.
 
@@
 It is reachable following a faint and intermittent track that departs from a marked path.
-It is craracterize by a suspended geological formation reminding an uvula, hence the name.
-
-Both caves are going to be analyzed in the course of the Underlandscape project \cite{Underlandscape}, and findings can further enhance the interest for both sites.
-
-\textbf{Qui aggiungerei notizie sugli altri punti in cui sono stati collocati i QR-code, correlandoli alla descrizione storica
-}
-
-%We return to the main path, this one looking over the Rio to proceed in the direction of Lucchio, while we abandon it to penetrate between the high rock walls carved over time by the Rio; mobile network coverage is absent, while the GPS has poor coverage and in many places is unreliable.
-When walking along the stream bed, great care is required, due to slippery pebbles and sections with wet foliage and moss.
-After about 300 m, follow a track that climbs up the slope to the right.
-When you reach a small plateau, you will find a rock shelter, which is not marked in the Tuscan cave cadastre and has not yet been investigated.
-Buca La Piella is located about 120 m further north.
-To reach it, climb up to the rock face by crossing a rather steep wooded slope.
-A great deal of caution is required, and for future use it would be a good idea to at least provide a safety rope.
-When you reach the wall, there is a short tunnel that leads into the cave.
-A very wide and high cavity opens up towards the east.
-The vault has a hole in the centre, and there are stalactites and stalagmites.
-The ground descends gently towards the opening, beyond which the slope becomes steep, making it necessary in the future to secure it with nets or fences.
+It is characterized by a suspended geological formation reminding an uvula, hence the name.
+
+Both caves are going to be analyzed in the course of the Underlandscape project \cite{Underlandscape}, and findings can further enhance the interest in both sites.
+
+Along the dirt road from the village in the direction of Lake \emph{Casoli} are some of the most interesting sites.
+The medieval remains of the {\em Church of Sant'Andrea de Lacu}, now in a state of serious disrepair, without its roof and with static problems, but characterized by its Romanesque forms and 10th-century decorations.
+
+The \emph{metati} (structures for drying chestnuts), together with the hundred-year-old chestnut trees, testify to the exploitation of forest resources since the Middle Ages.
+
+A little further along the CAI path is the {\em Oratory of the Madonna di Col di Piano}. Recently renovated, with its canopy it has been a shelter for wayfarers since modern times.
+
+Inside the town of \emph{Casoli}, narrow streets branch off to the main square, where the medieval {\em Church of Saints Donato e Andrea} with its bell tower can be admired.
+From here, an ancient street climbs to the top of the hill.
+On the way up there is a small oratory called {\em Madonna di Castello}, still consecrated but in poor conditions.
+
+On the summit are the remains of the {\em Medieval Castle} with its walls.
+In need of consolidation of the walls, it cannot be visited at the moment as it is on private property; it is only possible to observe part of its structures from the path.
+
+Two other sites that are particularly curious but cannot be visited because located on private properties are the so-called  {\em Celtic Calendar}, an artificially excavated rock interpreted as a kind of sundial, and an epigraph reused in a house interpreted as Lombard.
 
@@
 
-To test our signage proposal, we placed on the field several cards carrying QR codes. Se the list in Table~\ref{tab:qrtags}.
-
-% TODO: \usepackage{graphicx} required
+The concept behind the implementation was that of guiding the visitor through a self-organized experience using a signage technique that improves the traditional boards, as explained in section \ref{sec:minimize}.
+To verify on the field the practical aspects and the implementation process we realized several QR tags that were later secured to an existing support.
+See the list in Table~\ref{tab:qrtags}.
+
+The proposed signage cannot replace CAI signage (see figure \ref{fig:CAI}), which is recognized throughout the country and is very effective.
+However, the new signage can effectively complement it to provide more detailed information at multiple levels, from short text to in-depth information on the project site.
+In addition, being able to read the written directions along the path can be of great help to inexperienced hikers, particularly on unclear and not-quite-visible trails such as the one we suggested to reach Buca La Piella.
+
+Tags are printed on Synaps™, a non-biodegradable polyester synthetic film by Agfa-Gevaert NV.
+Agfa documents the production process giving guarantees of sustainability \citeWeb{agfa}.
+According to Agfa, Synaps is more sustainable than laminated paper.
+From previous experiences, it is also more durable.
+
+The tags are tied to pre-existing structures, like trees or poles, to keep the environmental footprint as lightweight as possible. The tag is tied with a thin biodegradable string.
+
+A QR tag is designed to implement three levels of reading (LoR), with an increasing level of technology involved:
+\begin{itemize}
+    \item visual: the information is printed on the tag. The user does not need any technology to read the content;
+    \item QR tag text: this is the text encoded in the tag. The user needs a smartphone-like device and an appropriate app to read, possibly aloud, the content;
+    \item URL: these are Web URLs encoded in the tag text. The user needs an Internet connection to visit the URL.
+\end{itemize}
+
+We placed QR tags both along the trail and at sites of interest; the former are placed at detours and, at more complex points, at close range so that from one you can see the next, as per the CAI standard \cite{cai10}.
+They can therefore have the function of a simple signpost to visually indicate the continuity of the trail.
+The second LoR gives access to the written indications recorded within it, which are more accurate and precise than those printed on the tag.
+The third LoR provides further capabilities, as detailed below, but is accessible only whenever the location is covered by the Internet.
+
+There are many ways to distribute the content among the three LoRs, dictated by tag purpose and location.
+For our experiment, we used the same organization for all tags, which privileges the second LoR, namely the text encoded in the tag (see figure \ref{fig:qrcodelegend}):
+
+\begin{itemize}
+    \item visual LoR (see figure \ref{fig:qrcodelegend}): the tag is approximately the size of a game card, 80 $cm^2$.
+$40\%$ of the frame (32 $cm^2$) provides mechanic resistance to the holes needed to secure the tag.
+$60\%$ of the remaining (48 $cm^2$) contains the QR-code, $20\%$ for a heading containing the name of the location, one line instruction for use, and the project logo, and a footer ($12\%$) for further instructions and project credits;
+
+    \item QR text LoR (see figure \ref{fig:qrcodecontent}): it describes the site or gives directions to reach the destination of the trail.
+The text includes historical, archaeological, and naturalistic information, together with the URL for the next LoR.
+In our prototype, the maximum length of the text amounts to 700 characters.
+A Huawei FIG-LX1 (2017) decodes the tag from a distance of more than 50cm using an Android application available on the Play Store.
+The same phone can read aloud the content (in Italian) without Internet connectivity;
+
+    \item URL LoR: such information is useful only in a few locations since the area is not uniformly covered by broadband networks: for instance, during a recon to the {\em Piella} cave, none of the smartphones of the participants received sufficient broadband network signal to visit the linked page.
+However, the application records the URL so that the user might visit it when entering a covered area.
+Each page contains site-specific information, an interactive map with the location of all the QR tags (see figure \ref{fig:linkmap}), and a form for user feedback (see figure \ref{fig:linkpoll}).
+The map is hosted by the UMap Web service (\url{https://umap.openstreetmap.fr/it/}) \citeWeb{umap}, and displays the location and content of all QR-tags (as shown in figure \ref{fig:qrcodecontent}).
+The server hosting the site is configured to collect aggregated statistics about visitors' activity.
+\end{itemize}
+
 \begin{figure}
-	\centering
-	\includegraphics[width=\linewidth]{figure/mappa}
-	\caption[Map with QR-tag positions]{Map with QR-tag positions}
-	\label{fig:mappa}
+\subfloat[QR-tag parts]{
+	\includegraphics[width = 0.55 \linewidth]{figure/QR-code-legend.png}
+	\label{fig:qrcodelegend}
+}
+\hfill
+\subfloat[QR-tag content on a 5.5' screen]{
+	\includegraphics[width = 0.35 \linewidth]{figure/qr-code-content}
+	\label{fig:qrcodecontent}
+}
+\caption{QR tag parts and content}
 \end{figure}
 
+\begin{figure}
+\subfloat[Linked page: the map]{
+	\includegraphics[width = 0.60 \linewidth]{figure/webpage-map}
+	\label{fig:linkmap}
+}
+\subfloat[Linked page: survey form]{
+	\includegraphics[width = 0.30 \linewidth]{figure/webpage-poll.png}
+	\label{fig:linkpoll}
+}
+\caption{Content of the Web page linked to the URL in a QR tag}
+\label{fig:webpage}
+\end{figure}
+
 \begin{table}
-	\begin{tabular}{|>{\raggedright\arraybackslash}p{3cm}|l|l|l|l|}
+    \begin{tabular} {|>{\raggedright\arraybackslash}p{3cm}|l|l|l|l|}
         \hline 
-\textbf{Name} & \textbf{Long} & \textbf{Lat} & \textbf{URL Key} & \textbf{Length} \\
-\hline
-Buca La Piella & 10.68361 & 44.03667 & t5ysrm & 476 \\
-\hline
-Calendario celtico & 10.66985 & 44.03986 & my0kp8 & 484 \\
-\hline
-Castello & 10.67062 & 44.04004 & 4l4r6y & 597 \\
-\hline
-Chiesa dei SS. Donato e Andrea & 10.66947 & 44.03928 & lwtyx6 & 632 \\
-\hline
-Iscrizione longobarda & 10.67244 & 44.03906 & 60m75s & 369 \\
-\hline
-Lago di Casoli & 10.67761 & 44.03468 & xqjpbk & 369 \\
-\hline
-Madonna di Castello & 10.66994 & 44.03960 & qlci89 & 395 \\
-\hline
-Madonna di Col di Piano & 10.67760 & 44.03173 & 3w44wr & 414 \\
-\hline
-Metato & 10.67764 & 44.03591 & sbgnl0 & 660 \\
-\hline
-Sant'Andrea de Lacu & 10.67592 & 44.03530 & fxq83v & 657 \\
-\hline
-Antro dell’Ugola & 10.68296 & 44.03871 & mprs0w & 226 \\
-\hline
-Deviazione per Antro dell’Ugola & 10.68343 & 44.03971 & e4n2js & 88 \\
-\hline
-Deviazione per Buca La Piella & 10.68427 & 44.03573 & hve4pj & 139 \\
-\hline
-Deviazione per Sant’Andrea de Lacu & 10.67663 & 44.03507 & 13uav3 & 200 \\
-\hline
-Ingresso Buca La Piella & 10.68358 & 44.03592 & 5ahvp8 & 190 \\
-\hline
-Riparo sottoroccia lungo percorso Piella & 10.68397 & 44.03704 & kwr1wx & 286 \\
-\hline
+        \textbf{Name} & \textbf{Long} & \textbf{Lat} & \textbf{URL Key} & \textbf{Length} \\ \hline
+        Buca La Piella & 10.68361 & 44.03667 & t5ysrm & 476 \\ \hline
+        Calendario celtico & 10.66985 & 44.03986 & my0kp8 & 484 \\ \hline
+        Castello & 10.67062 & 44.04004 & 4l4r6y & 597 \\ \hline
+        Chiesa dei SS. Donato e Andrea & 10.66947 & 44.03928 & lwtyx6 & 632 \\ \hline
+        Iscrizione longobarda & 10.67244 & 44.03906 & 60m75s & 369 \\ \hline
+        Lago di Casoli & 10.67761 & 44.03468 & xqjpbk & 369 \\ \hline
+        Madonna di Castello & 10.66994 & 44.03960 & qlci89 & 395 \\ \hline
+        Madonna di Col di Piano & 10.67760 & 44.03173 & 3w44wr & 414 \\ \hline
+        Metato & 10.67764 & 44.03591 & sbgnl0 & 660 \\ \hline
+        Sant'Andrea de Lacu & 10.67592 & 44.03530 & fxq83v & 657 \\ \hline
+        Antro dell’Ugola & 10.68296 & 44.03871 & mprs0w & 226 \\ \hline
+        Deviazione per Antro dell’Ugola & 10.68343 & 44.03971 & e4n2js & 88 \\ \hline
+        Deviazione per Buca La Piella & 10.68427 & 44.03573 & hve4pj & 139 \\ \hline
+        Deviazione per Sant’Andrea de Lacu & 10.67663 & 44.03507 & 13uav3 & 200 \\ \hline
+        Ingresso Buca La Piella & 10.68358 & 44.03592 & 5ahvp8 & 190 \\ \hline
+        Riparo sottoroccia lungo percorso Piella & 10.68401 & 44.03573 & kwr1wx & 286 \\    \hline
     \end{tabular}
-\caption{The QR-tags placed around Casoli.
-The URL key is a random code that appears in the web page linked to the QR code.
-The Length refers to the description in the QR-code and is in characters.
+\caption{The QR-tags placed around \emph{Casoli} .
+The URL key is a random code that appears on the web page linked to the QR code.
+The Length refers to the description in the QR tag and is in characters.
 \label{tab:qrtags}}
@@
 
-A card sample is in Figure \ref{fig:toscana-casoli}.
-On the top there is the logo of the funding project, a descriptive title ({\em Castle} in english) and concise directions for use {\em Scan fo further information}).
-Under the QR-code is the notice that other QR-codes are present in the vicinity, and a reference to the funding project.
-
-The text in the QR-code (in figure \ref{fig:qrcodecontent}) contains the main text with 614 characters (100 words) and a URL 86 characters long.
-The URL is useful only if the Internet is reachable, and brings to a new page containing a map with the location of all the tags (Figure \ref{fig:linkmap}) and a short survey form.
-Note that, while the QR-tag content is always available, the URL is reachable in locations with broadband coverage.
-In our case, during a recon to the {\em Piella} cave none of the smartphones of the participants received sufficient broadband network signal.
+The production process of a series of tags has also been investigated to streamline the task and use only basic Information Technology skills.
+
+The master document is a \emph{GeoJSON} file describing the area and the tagged features: a GPS tracking application can be used to record an initial version of it during a recon.
+
+The finalized master document contains a single \emph{GeoJSON} \emph{FeatureCollection} object containing one \emph{Feature} object for each tag.
+Each \emph{Feature} contains a \emph{Geometry} object of type \emph{Point}, and a \emph{properties} object with fields containing information needed to create the tag: the \emph{uid}, a random string of six lowercase letters used to produce the associated URL, the \emph{title} printed on the tag, and the text encoded in the QR code (see table \ref{code:geojson}) where part of the {\em description} field is omitted.
+The \emph{properties} are also shown on the UMap map.
+
+Editing the basic map to fill-in the QR-code content and the other fields does not require technical skills since the tools integrated into the UMap user interface already fit the purpose (see figure \ref{fig:umap-editing}).
+The resulting geojson object for a tagged feature is shown in figure \ref{fig:geojson-sample}.
+
+The master document is rendered by UMap on the page linked to the tags (as in figure \ref{fig:linkmap}).
 
 \begin{figure}
-\subfloat[QR-tag content]{
-	\includegraphics[width = 0.3 \linewidth]{figure/qr-code-content}
-	\label{fig:qrcodecontent}
-}
-\subfloat[Linked page: the map]{
-	\includegraphics[width = 0.3 \linewidth]{figure/sito-a}
-	\label{fig:linkmap}
-}
-\subfloat[Linked page: survey form]{
-	\includegraphics[width = 0.3 \linewidth]{figure/sito-b}
-	\label{fig:linksurvey}
-}
+\subfloat[Editing a feature associated with a QR-tag in UMap]{
+	\includegraphics[width = 0.35 \linewidth]{figure/umap-editing.png}
+	\label{fig:umap-editing}
+}
+\hfill
+\subfloat[GeoJSON \emph{Object} representing the same the feature]{
+	\includegraphics[width = 0.60 \linewidth]{figure/geojson-sample.png}
+	\label{fig:geojson-sample}
+}
+\caption{Editing the GeoJSON with uMap editor. The resulting file is automatically converted into printable QR-tags}
+\label{fig:editing}
 \end{figure}
 
-We placed QR-tags both along the trail and at sites of interest; the former are placed at detours and, at more complex points, at close range so that from one you can see the next, as per the CAI standard \cite{cai10}.
-
-They can therefore have either the function of a simple signpost to indicate the continuity of the trail, or be scanned to read the written indications recorded within it, which are more accurate and precise.
-
-%QR codes placed at sites of interest provide historical, archaeological and naturalistic information through short text descriptions.
-They also give the possibility, where phone coverage allows, to deepen the project and discover more details about the proposed itinerary through a link to the dedicated site.
-
-This approach ensures that the tourist is guided through the visitor experience and, at the same time, that the impact of signage on the environment is reduced compared with traditional boards, as explained in section \ref{sec:minimize}.
-
-%, considering that there is no need for panels and vertical signage involving the placement of poles.
-In addition to being more impactful, classic vertical signage can more easily suffer degradation and needs more maintenance, especially the information paneling on the sites; whereas in our proposal the possible maintenance of the QR code would constitute very short time and rather low cost.
-
-Clearly, our methodology does not aim to replace CAI signage, which is recognized throughout the country and is very effective, especially in terms of horizontal typology, but at the very least it is repurposed to complement it in order to provide more detailed information at multiple levels of interest, from short text to in-depth information on the project site.
-In addition, being able to read the written directions along the path can be of great help to even the most inexperienced hikers, particularly on unclear and not-quite-visible trails such as the one we suggested to reach Buca La Piella.
-
-% ====
-
-% Concentrate on outdoor activities
-
-% Starting from previous experiences documented in the bibliography we explore a few broad categories, to find weaknesses and strengths for each of them.
-
-% One of the issues the host must resolve is the availability of support for electronic devices. The provisioon of such support has a footprint that ha
-
-% may be present, but, in the spirit of our investigation, power supply should be address renewable sources, like wind, sunlight and similar.
-
-% One option discriminates the use of an electronic device on the side of the user.
-
-% In case we assume the user does not have a electronic device with her, the 
-% One option is to introduce active electronics: the touristic site administration provides a
-
-%Materials and Methods should be described with sufficient details to allow others to replicate and build on published results.
-Please note that publication of your manuscript implicates that you must make all materials, data, computer code, and protocols associated with the publication available to readers.
-Please disclose at the submission stage any restrictions on the availability of materials or information.
-New methods and protocols should be described in detail while well-established methods can be briefly described and appropriately cited.
-
-%Research manuscripts reporting large datasets that are deposited in a publicly avail-able database should specify where the data have been deposited and provide the relevant accession numbers.
-If the accession numbers have not yet been obtained at the time of submission, please state that they will be provided during review.
-They must be provided prior to publication.
-
-%Interventionary studies involving animals or humans, and other studies require ethical approval must list the authority that provided approval and the corresponding ethical approval code.
-%\begin{quote}
-%This is an example of a quote.
-%\end{quote}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
+A \emph{Bash} script running on a Linux system makes straightforward the conversion of the GeoJSON file into a printable array of QR tags.
+The code, a total of 90 source lines written in Bash and Python taking advantage of the ogr2ogr and qrencode commands, is on GitHub \citeWeb{github-qr}.
+
+\section{Results and discussion \label{sec:results}}
+
+Our point is to illustrate a methodology to improve the touristic offer of a site that is not specific to a use case, but applicable to a broad range of frameworks.
+The methodology must be sustainable in the broader sense: social, environmental, and financial.
+
+We applied the methodology to a small-scale example: a signage system addressing the guidance of visitors in a rural area with a cultural heritage to be protected and valorized at the same time.
+
+Firstly we evaluated the aspects of interest of the site, and we found natural formations and human artifacts tracing back to the Paleolithic Era.
+Other relevant aspects are that the site is not included in mass tourism circuits but is not far from them.
+The area is exposed to depopulation due to the shortage of productive activities.
+
+To preserve the native social and financial framework we do not aim at mass tourism.
+Instead, we target a motivated, non-casual visitor, and devise potential reasons for its visit.
+We found many of them: a not easily accessible cave inhabited since pre-history, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
+
+We investigated the task of guiding the visitor and telling the history of the place with inexpensive and non-intrusive techniques.
+The proposed solution consists of a dozen QR tags placed at key locations.
+The total cost of the tags amounts to a few Euros for the printing service, although finding a shop with an appropriate printing device may be difficult.
+The impact on the environment is extremely limited, as shown in figure \ref{fig:TagPlacement}.
+
+\begin{figure}
+\subfloat[A tag secured to a tree near a shelter]{
+	\includegraphics[width = 0.33 \linewidth]{figure/Riparo1-070623-cut.jpg}
+	\label{fig:TagOnTree}
+}
+\hfill
+\subfloat[Google Earth view of the surroundings of the tag. The map is north-oriented, and shows  the village of \emph{Casoli}  in the left-high corner]{
+	\includegraphics[width = 0.62 \linewidth]{figure/GoogleEarth.png}
+	\label{fig:GoogleEarth}
+}
+\caption{Tag placement at different scales}
+\label{fig:TagPlacement}
+\end{figure}
+
+The preparation of the signage requires a solid knowledge of the site. This demanded consulting historical documents, assessing the touristic vocation of the area, and two on-site surveys.
+
+On the technical side, the proposed signage provides a benchmark difficult to improve in terms of environmental impact and cost, but its {\em performance} is difficult to evaluate.
+
+One performance index might be the number of times a given tag is scanned.
+Indeed, we recorded several hits to the URLs in the QR tags, but their number and the way they are collected are not appropriate to form a statistic of their utilization.
+One reason is the fact that the use of the tag is not always associated with a hit, and vice-versa.
+
+The reactions we collected by interviewing local stakeholders are positive, which gives us the impression that the initiative is respectful and consistent with the social and economic framework.
+The measurement of the efficacy in financial terms is beyond the scope of this paper, but it has to be said that the aim of this work is to provide assistance and improve the experience of actual visitors, not to multiply their number.
+
+Although we are satisfied that the solution complies with the sustainability principles that inspired this work, we identified two issues that indicate directions for future research:
+
+\begin{itemize}
+    \item The balance between visual impact and effectiveness is critical.
+The current dimension of the tag is such that the visual impact is limited, but this negatively affects its efficacy: it is easy to miss a tag containing possibly relevant information;
+    \item Monitoring visitor's activity, like the frequency of visits of a given tag, or the sequencing of visited tags, which may be of interest for managing the site, depends on user motivation.
+\end{itemize}
+
+Trading off cost and simplicity for effectiveness and measurability there are solutions, based, for instance, on Bluetooth communication or dedicated applications, that may improve such aspects.
+
+\section{Conclusions}
+
+Like any other productive activity, tourism is exposed to being unsustainable, damaging the very resources that allow it.
+To avoid heavy drawbacks the design of the activity has to consider the complex framework where it will operate, and the word used to indicate this is \emph{holistic}, and it is extremely difficult to put into practice.
+A methodology made of examples and guidelines simplifies the task.
+
+This paper is a step in this direction.
+We address a common problem, that of providing the visitor with directions in an area not reached by mass tourism, with the aim of acquiring the financial resources for its survival while preserving its distinguishing traits.
+
+We show how a balanced use of technology may help in the task, with a process that is necessarily multidisciplinary, joining skills coming from the diverse domains: sociology, economics, humanities and technical.
+The paper demonstrates how their interplay brings ad-hoc results targeting sustainability and efficacy.
+
+The resulting implementation is original in the literature, and minimalistic in its deployment. The detailed description enables its reuse as a starting point in future initiatives.
+
+The discussion regarding the deployed experiment indicates directions for future investigation on the technical side.
+For instance, the availability of non-intrusive long-range devices that send a message whenever sensing a presence might improve signals discoverability.
+The Bluetooth Low Evergy technology goes in that direction, but hidden tags with BLE and energy harvesting would have broken the economy constraints.
+
+From a different point of view, the lack of discoverability suggests a recreational scenario inspired by Geocaching® and Pokemon-GO®.
+
+In conclusion, more sustainable solutions to old problems exist and are reachable melting technology and humanities.
+Policy-makers are crucial in the development and monitoring of sustainable measures.
+This is especially important in the tourism industry where sustainability is a multifaceted topic that includes resident ecosystem resilience and regeneration efforts by both policy-makers and private tour operators.
 
@@
 %
-%\subsection{Figures, Tables and Schemes}
+%\subsection{Figures, Tables, and Schemes}
 %
@@
 (\textbf{b}) Description of what is contained in the second panel.
-Figures should be placed in the main text near to the first time they are cited.
+Figures should be placed in the main text near the first time they are cited.
 A caption on a single line should be centered.\label{fig1}}
@@
 
+%\nocite{*}
 \bibliography{biblio/paper.bib}
+
+\nociteWeb{*}
+\bibliographystyleWeb{abbrv}
+\bibliographyWeb{biblio/sitography.bib}
 

### 2023-11-04 paper.tex
Commit riportato del 31 ottobre - Spedito in prima revisione

@@
 % Full title of the paper (Capitalized)
-\Title{A sustainable approach to tourist signage on (historical) trails}
-
+\Title{A sustainable approach to tourist signage on heritage trails\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}}
 % MDPI internal command: Title for citation in the left column
-\TitleCitation{A sustainable approach to tourist signage on historical trails}
-
+\TitleCitation{A sustainable approach to tourist signage on heritage trails}
 % Author Orchid ID: enter ID or remove command
-\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
-%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name
+\newcommand{\orcidauthorA}{0009-0003-1136-570X} % Add \orcidA{} behind the author's name
+\newcommand{\orcidauthorB}{0000-0002-9734-2044} % Add \orcidA{} behind the author's name
+\newcommand{\orcidauthorC}{0000-0002-5542-7891} % Add \orcidB{} behind the author's name
 
 % Authors, for the paper (add full first names)
-\Author{Firstname Lastname $^{1,\dagger,\ddagger}$\orcidA{}, Firstname Lastname $^{2,\ddagger}$ and Firstname Lastname $^{2,}$*}
+\Author{ Maria Grazia Deri $^{2,}$*\orcidC{}, Letizia Chiti $^{3}$\orcidA{}, Augusto Ciuffoletti $^{1}$\orcidB{} }
 
@@
 % MDPI internal command: Authors, for metadata in PDF
-\AuthorNames{Firstname Lastname, Firstname Lastname and Firstname Lastname}
+\AuthorNames{Maria Grazia Deri, Letizia Chiti, and Augusto Ciuffoletti}
 
 % MDPI internal command: Authors, for citation in the left column
-\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
+\AuthorCitation{ Deri, M.G.; Chiti, L.; Ciuffoletti, A.}
 % If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.
@@
 \address{%
-$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
-$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}
+$^{1}$ \quad University of Pisa (retired); augusto.ciuffoletti@unipi.it\\
+$^{2}$ \quad University of Pisa; mariagrazia.deri@sp.unipi.it\\
+$^{3}$ \quad University of Pisa; l.chiti2@studenti.unipi.it}
 
 % Contact information of the corresponding author
-\corres{Correspondence: e-mail@e-mail.com; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}
+\corres{Correspondence: augusto.ciuffoletti@unipi.it}
 
 % Current address and/or shared authorship
-\firstnote{Current address: Affiliation 3.} 
-\secondnote{These authors contributed equally to this work.}
+%\firstnote{Current address: Affiliation 3.} 
+%\secondnote{These authors contributed equally to this work.}
 % The commands \thirdnote{} till \eighthnote{} are available for further notes
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{Guiding the visitor to appreciate all the attractions of a region is paramount for building a memorable experience.
+\abstract{Guiding the visitor to appreciate the sights during an excursion is crucial to building a memorable experience.
 In the literature, we find many articles exploring the use of sophisticated technologies towards such a goal.
-The case of wilderness resources exhibits a peculiar aspect, since the presence of human artifacts and the signage may damage the experience, and introduce pollutants as well.
-In this paper, we consider the case of a natural trail targeted to niche tourism to understand the available options for signage aiming to provide directions, guide the exploration, and warn about dangers.
-The footprints of the various options are evaluated, and we describe the outcomes of an experimental setup implemented to verify the applicability of our approach.}
+The case of tourism proposals for inland areas exhibits a noteworthy aspect, since the presence of human artifacts and the signage may damage the experience, and introduce pollutants as well.
+This paper analyzes the signage for a natural trail putting into practice a holistic methodology that takes into account all aspects of the problem, starting from the touristic vocation of the area and the social fabric.
+The analysis indicates slow, community-involved tourism as the preferable target, with extreme requirements of sustainability in the social, economic, and environmental domains for the signage solution.
+Appropriate governance guidelines are given to implement the proposal.
+Footprint evaluations address the QR-code technology, which is used to design a real-scale solution finally deployed on the field.}
 %\abstract{A single paragraph of about 200 words maximum.
@@
 % Keywords
-\keyword{keyword 1; keyword 2; keyword 3 (List three to ten pertinent keywords specific to the article; yet reasonably common within the subject discipline.)}
+\keyword{Inland areas; heritage trail; digital sustainability; slow tourism; destination governance; community-involved tourism; cultural heritage; holistic approach; QR-code; trail signage}
 
@@
 
-%\newcites{Main}{References}
-%\newcites{Sitology}{Sitology}
-
-\tableofcontents
+%\tableofcontents
 
@@
 Although an official decision is still pending, there is a widely agreed proposal for the naming of the time that we are living as the Anthropocene.
-Behind this proposal is the recognition that humans have the potential to significantly alter the environment in which they live on a global scale.
-
-The problem with the above fact is that we humans are not able to understand long-term consequences of deploying such a potential, nor even to keep it under control.
+Behind this proposal is the recognition that humans have the potential to significantly alter on a global scale the environment in which they live.
+
+The problem with the above fact is that we humans are not able to understand the long-term consequences of deploying such a potential, nor even to keep it under control.
 Even worse, such capacities are often used with the aim of short-term local effects, ignoring long-term global ones, with the risk of long-term global deterioration.
@@
 \cite{ram15a} consider the concept behind this word as "vague and disputed but not meaningless" and that cannot be used without a concrete framework, especially in a normative context.
-To solve to problem of measuring sustainability, Tom Kuhlman et al.
-\cite{kuh10a} point out an inherent tension in the term between growth and stability.
+To solve the problem of measuring sustainability, Tom Kuhlman et al.
+\cite{kuh10a} point out an inherent conceptual tension between growth and stability.
 This tension has been attenuated by a change in the meaning of the word, which has gone from focusing on preservation for an indefinite future to a concept that includes three "pillars", society, economy, and environment, as determinants that need to be protected in the present to promote sustainable development in the future.
@@
 
-The promotion of human cultural heritage is situated in such a complex scenario.
+The promotion of cultural heritage is situated in such a complex scenario.
 As noted by Jorge Otero in \cite{ote22a}, local communities benefit from the promotion of heritage: conservation activity generates social growth, while tourism provides an extra income that can be reinvested in promotion.
-However, just as in the case of ski resorts, there is a risk of compromising the resource itself.
-This interaction raises a question of sustainability also for cultural conservation initiatives, two issues that are usually considered in synergy.
-As in the document "World Heritage and Sustainable Development" of UNESCO, which calls for actions that "harness the potential of World Heritage properties and heritage in general, to contribute to sustainable development".
-
-This research stems from such a painful contradiction in trying to answer the question: is there a way to unlock the potential of a cultural resource while at the same time limiting the damage to a social, economic, and environmental equilibrium?
-
-The question arises in the course of a project funded by the Italian Ministry of Research aimed at applying cutting-edge technologies to the investigation of karst caves that may have been inhabited in the past, from prehistory to today.
-Among the expected results of the project is the definition of a strategy to disseminate the results of the research and allow non-specialists to visit and enjoy the results, thus contributing to the economy of rural areas that are severely exposed to de-population.
+However, just as in the case of ski resorts, there is a risk of compromising the heritage itself.
+This interaction raises a question of sustainability also for heritage valorization initiatives, two issues that, instead,  are often considered in a virtual synergy.
+As stated in the document titled 'World Heritage and Sustainable Development' by UNESCO \cite{lab17a}, calling for actions that "harness the potential of World Heritage properties and heritage in general, to contribute to sustainable development".
+
+This research stems from such a tough contradiction in trying to answer the question: is there a way to unlock the potential of cultural heritage while at the same time limiting the damage to a social, economic, and environmental equilibrium?
+
+The question arises in the course of the Underlandscape project \citeWeb{underlandscape} funded by the Italian Ministry of Research aimed at applying cutting-edge technologies to the investigation of karst caves that may have been inhabited, from prehistory to today.
+Among the expected results of the project is the definition of a strategy to disseminate the research outcomes allowing non-specialists to visit and enjoy the results, thus contributing to the economy of inland areas that are severely exposed to de-population.
 A sustainable strategy is even more crucial as the geological environment of interest is particularly fragile, as noted by Aleksander Antic\`c et al.
@@
 
-This document provides a conceptual solution for a specific aspect.
-The solution is conceptual in that it tries to follow a guideline that takes into account as much detail as possible in a way that can be used in other environments but is very practical in its application.
-Given the closely interlinked nature of the situation, we observe how a very specific aspect, tourist signage, exhibits effects on society, the economy, and the environment.
+This document provides a conceptual solution for a specific aspect: heritage trail signage.
+The solution is conceptual in that it tries to follow a guideline that takes into account many aspects and provides a practical way that is adaptable to other environments.
 
@@
 
-The functions of signage are twofold
+The functions of signage are two:
 \begin {itemize}
 \item an effective signage must guide the tourist across the resort.
-A map represents a starting point to this end, but not all visitors feel comfortable reading a map.
-The ideal signage should include visible reference points and explicit actions, like "turn right after crossing the stream".
+A map represents a starting point for this purpose, but not all visitors feel comfortable reading a map.
+The ideal signage should provide visible reference points and suggest actions, like "turn right after crossing the stream".
 A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}).
-The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail
+The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail;
 \item a sign should explain the reasons for interest in a site.
 This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit (see figure \ref{fig:pole-direction}).
-In this sense, the CAI signal above is not useful.
-A simple board with a site name may be sufficient only when the site is sufficiently popular.
-Otherwise, a more structured message should be used to explain the reasons why the feature is regarded as relevant.
+In this sense, the CAI signal above is not sufficient.
+A simple board with a site name fits only when the site is sufficiently popular.
+Otherwise, a more structured message should be used to explain its relevance.
 \end{itemize}
 
-In our study, we consider a range of solutions that help towards sustainable support to sustainable tourism, specifically to the dynamic provision of information during a visit along a natural trail, not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}).
+In our study, we consider a range of solutions that support sustainable tourism, specifically the dynamic provision of information during a visit along a natural trail not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}).
 We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.
@@
 The task of the {\em host} organization is to provide the {\em user} with all sorts of information that may guide him across the visit, making it as profitable and enjoyable as possible.
+
 A traditional approach makes use of physical information boards with graphical or textual content, as in figure \ref{fig:traditional}.
-
 Our study starts by pointing out the issues related to such a solution:
@@
 \begin{itemize}
-	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance
-	\item installation logistics: depending on the location, the transportation and placement of a plaque may require a basement or other sorts of supports
-	\item environmental impact: to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site 
-	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons
-	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well
-	\item update limits: to update the content, the plaque must be replaced
-	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues similar to those found during the installation
+	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance;
+	\item installation logistics: depending on the location, the transportation and placement of a plaque may require a basement or other sorts of supports;
+	\item environmental impact: to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site; 
+	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons;
+	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well;
+	\item update limits: to update the content, the plaque must be replaced;
+	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues similar to those found during the installation.
 \end{itemize}
 
-Such considerations motivate an interest in an alternative way of communication.
+Such points motivate an investigation to find alternative ways.
 
@@
 
-We characterize the problem as an instance of {\em weak sustainability}: we do not preclude intervention in the environment, but the impact of such intervention must be better than that related to traditional signs.
+We characterize our approach as an instance of {\em weak sustainability}: we do not preclude intervention in the environment, but the impact of such intervention must be better than that related to traditional signs.
 To this end, we include in our solution tools that are not part of the site but remain with the user.
@@
 During the last decades, we witnessed the diffusion of smartphone devices, and we have no reason to expect a change in such a trend.
-Smartphones empower individual communication capabilities, allowing them to receive sound, visual, and tactile interactions.
+Smartphones enhance the communication capabilities of individuals by allowing them to receive sound, visual, and tactile interactions.
 The relevance of such capabilities for the improvement of a tourist experience has been widely investigated, either in urban environments \cite{liu16a} or in a rural milieu \cite{kum20a} to find ways to exploit such tools.
-Again, we start from the limits of such technology, especially in the case of outdoor activities.
+As we did with physical information boards, we highlight the limitations of this technology, particularly when it comes to outdoor activities.
 
@@
 
-Regarding {\em usability}, the basic recommendation is to keep the operation as simple as possible, within the experience of the majority of users, without requiring the need to familiarize themselves with new applications.
+Regarding \textit{usability}, the basic recommendation is to keep the operation within the experience of the majority of users, without requiring them to familiarize themselves with new applications.
 
@@
 
-On the other hand, roaming in places not reached by the Internet, or {\em dead zones} using the words of Pearce and Gretzel in \cite{pea12a}, may evoke contrasting feelings, from rewarding to threatening.
+On the other hand, roaming in places not reached by the Internet, or {\em dead zones} using the term introduced by Pearce and Gretzel in \cite{pea12a}, may evoke contrasting feelings, from rewarding to threatening.
 
@@
 
-The present paper wants to fill the gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined aspect of tourism support, taking into account "by design" its sustainability.
-Once a range of relevant solutions has been identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies the feasibility of a specific solution.
-
-We stress that this article does not aim now at quantifying user satisfaction or the economic revenue associated with the specific solution.
-Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve further investigation.
+The present paper wants to fill the gap highlighted by Rodriguez, providing a conceptual yet pragmatic approach to a well-defined aspect of tourism support, taking into account its sustainability "by design".
+Once a range of relevant solutions is identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies the feasibility of a specific solution.
+
+We emphasize that this article is not meant to quantify user satisfaction or the economic revenue associated with the specific solution.
+Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve a thorough investigation.
 We aim at isolating an issue, proposing a sustainable strategy for its solution, and implementing a proof of concept for it.
@@
 
-We aim to design sustainable and effective smartphone-based signage.
-
-A straightforward approach concentrates on the wireless capabilities of such a device.
-Given the premise that the Internet is not reachable, the {\em host} stakeholder might provide a local network of low-power radios covering the region of interest.
-Small servers connected to the network would provide specific Web services.
+%We aim to design sustainable and effective smartphone-based signage.
+
+The wireless capabilities of a smartphone are the focus of a straightforward approach.
+Given the premise that the Internet is assumed to be unreachable, the {\em host} stakeholder might provide a local network of low-power radios covering the region of interest.
+Small servers connected to the network would host specific Web services.
 The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single-chip computer (SCC) has a coverage of tens of meters, and a volume in the order of the $cm^3$.
-They can provide WiFi Access Points as well as Web content.
+It can provide a WiFi Access Point as well as Web content.
 The ESP8266 has a capacity of 32KB, which is sufficient for explanatory text and a low-resolution image.
@@
 
-Power supply dependency is a serious limitation for such kind of solution.
+Such a solution is severely limited by the power supply dependency.
 A radio transmitter is a rather power-consuming device, in the range of Watts.
@@
 The NFC device capacity is in the range of the KBytes, nearly a page on this journal.
-Storing content in an NFC tag requires a smartphone with a specific app.
-
-The NFC technology is currently very diffused, and reading an NFC tag as text requires the installation of an appropriate application.
-Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the {\em user's} device as a chink of text which can be treated as such.
-The smartphone can read it aloud, to compensate user's inabilities, or translated, to cope with linguistic issues.
+Engraving content onto an NFC tag is a straightforward task that requires a smartphone and an ad-hoc app.
+
+The NFC technology is currently very diffused, although reading an NFC tag as text requires the installation of an appropriate application.
+Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the {\em user's} device as a chink of text.
+The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues.
 Such capabilities do not need an Internet connection.
@@
 	\item {\em Flexibility in use} is enabled by the device capabilities, which allow listening instead of reading, translating the information in a different language, or storing it for later use,
-	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone
-	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation
+	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone,
+	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation,
 	\item{\em Tolerance for error}: there are no margins to use the device in a way that compromises user safety.
-The deliberate or accidental release of the passive device into the environment determines a minor pollution
-	\item{\em Low physical effort} holds, although the user needs to carry the smartphone
+The deliberate or accidental release of the passive device into the environment determines a minor pollution,
+	\item{\em Low physical effort} holds, although the user needs to carry a smartphone,
 	\item{\em Size and Space for approach and use} need to be carefully considered.
-In the case of the NFC tag the smartphone needs to be nearly in touch with the passive device, while the QR code must be in the line of sight and frameable without effort
+In the case of the NFC tag the smartphone needs to be nearly in touch with the passive device, while the QR code must be in the line of sight and frameable without effort.
 \end{itemize}
@@
 
-As anticipated in the introduction, our study wants to provide a proof of concept for a sustainable solution in a concrete setup.
-To comply with such a holistic approach, we need to start with the definition of the operation framework.
-
-This section is devoted to the description of an economic and social context, as well as the historical background representing the heritage resource we want to promote.
-Next, we will discuss the technical details of the solution.
-
-The geographical area of interest is the surroundings of \emph{Casoli}, a small village in a mountainous region in the north of Tuscany, in Italy.
+Our study aims to demonstrate a sustainable solution in a concrete setting through a proof of concept.
+In order to follow a holistic approach, we must start with the definition of the operation framework.
+
+This section is dedicated to describing the economic and social context, as well as the historical background that characterizes the heritage resource we want to promote.
+
+The geographical area of interest is the surroundings of \emph{Casoli}, a small village in a mountainous region in the north of Tuscany, Italy.
 The area falls within the municipality of \emph{Bagni di Lucca}, in the province of \emph{Lucca}.
-In 2021, the archaeological team of our project realized an archaeological map assessing the reasons of interest for the heritage and the geomorphological features of the area.
+In 2021, the archaeological team of our project realized a thematic map assessing the reasons of interest for the heritage and the geomorphological features of the area.
 
@@
 \emph{Bagni di Lucca} is located on the north-eastern boundary of the Province of \emph{Lucca}, in \emph{Val di Lima}, and is part of the \textit{Media Valle del Serchio} district.
-With its mountains it marks the historical border with the Modena and Pistoia area, it is very rich in potential for its naturalistic, archaeological, and historical heritage, both expressed and as yet unexpressed.
+With its mountains it marks the historical border with the \textit{Modena} and \textit{Pistoia} area, it is very rich in potential for its naturalistic, archaeological, and historical heritage, both expressed and as yet unexpressed.
 Thanks to the archaeological map the main sites of interest from prehistoric to contemporary times in this municipality are cataloged, photographed, and georeferenced.
 
-From a historical-geographical point of view, this area is identified with the \emph{Lima} stream and its tributaries, which impress to the area a particular geomorphology, very impervious despite the modest hilly elevations.
-
-The \emph{Lima} stream has characterized the history of \emph{Bagni di Lucca} since ever: some of the oldest evidence of human presence in the valley has been found along its ancient river terraces, such as caves and rock shelters frequented since the Palaeolithic age; the manufacturing industry (paper mill, flour mills and, in recent times, energy production) have exploited since the Middle Age until the 1980s \cite{men76, bed05, ser21}.
-
-On the naturalistic side, the region of \emph{Bagni di Lucca} encompasses an incredible concentration of biodiversity, counting no less than three sites of the Natura 2000 Network \cite{nat00} European Economic Community (EEC) initiative.
+From a historical-geographical point of view, this area is identified with the \emph{Lima} stream and its tributaries, which impress the area with a peculiar geomorphology impervious despite the modest hilly elevations.
+
+The \emph{Lima} stream has characterized the history of \emph{Bagni di Lucca} since ever: the valley's ancient river terraces contain some of the earliest evidence of human presence, such as caves and rock shelters frequented since the Palaeolithic age; the manufacturing industry (paper mill, flour mills and, in recent times, energy production) have exploited since the Middle Age until the 1980s \cite{men76, bed05, ser21}.
+
+On the naturalistic side, the region of \emph{Bagni di Lucca} encompasses an incredible concentration of biodiversity, counting no less than three sites of the Natura 2000 Network \citeWeb{nat00} European Economic Community (EEC) initiative.
 
@@
 \begin {itemize}
-\item the limestone areas of \emph{Val di Lima} and Balzo Nero; 
-\item Monte Prato Fiorito-Monte Coronato-Valle dello
-Scesta; 
-\item Orrido di Botri.
+\item the limestone areas of \emph{Val di Lima} and \emph{Balzo Nero}; 
+\item \emph{Monte Prato Fiorito}-\emph{Monte Coronato}-\emph{Valle dello
+Scesta}; 
+\item \emph{Orrido di Botri}.
 \end{itemize}
@@
 	\includegraphics[width = 0.3\linewidth]{figure/italia.png}
-	\label{fig:population}
 }
@@
 	\includegraphics[width = 0.3 \linewidth]{figure/toscana.png}
-	\label{fig:tourismflows}
 }
@@
 	\includegraphics[width = 0.3 \linewidth]{figure/bagni_di_lucca.png}
-	\label{fig:tourismflows}
 }
@@
 
-The latter is also a SPA (Special Protection Area) and the \textit{Orrido di Botri} State Reserve is located within it \cite{nat00}.
+The latter is also a SPA (Special Protection Area) and the \textit{Orrido di Botri} State Reserve is located within it \citeWeb{nat00}.
 It is therefore a natural heritage with a fragile balance that needs to be preserved.
@@
 
-In the last few years the community opened two entirely new trails:
+In the last few years, the community opened two entirely new trails:
 \begin{itemize}
@@
 Another recently developed project (2019) is the expansion of the Saint Bartholomew's Path, which runs through the territory of \textit{Pistoia}, with a
-variant that from \textit{Popiglio} continues in five stages in the municipality of \emph{Bagni di Lucca} to \textit{Pieve di Controne}, acting as the '\emph{Lucca} gateway' to the Path \cite{camsb}.
+variant that from \textit{Popiglio} continues in five stages in the municipality of \emph{Bagni di Lucca} to \textit{Pieve di Controne}, acting as the '\emph{Lucca} gateway' to the Path \citeWeb{camsb}.
 
@@
 This approach can involve the community, especially in the southern part of the municipality, which is still less frequented, more hilly, and therefore less traveled by the network of trails.
-We propose the creation of geo-itineraries characterized by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.
+We envision the creation of geo-itineraries characterized by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.
 
@@
 
-Ceramic fragments dated between the 3rd century B.C.
-and the 1st-2nd century A.D.
-prove that Ligurian populations occupied the region scattered and in small nuclei.
-The dedication of the Latin colony of \textit{Lucca} in 180 B.C.
-marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.
+Ceramic fragments dated between the 3rd century BC and the 1st-2nd century AD prove that Ligurian populations occupied the region scattered and in small nuclei.
+The dedication of the Latin colony of \textit{Lucca} in 180 BC marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.
 
@@
 
-In Longobard and Carolingian times, \emph{Lucca}'s \emph{Val di Lima} was one of the three administrative districts into which the mountains were divided and was called “fines Contronenses” \cite{qui02, cia06, cia11}.
+In Longobard and Carolingian times, \emph{Lucca}'s \emph{Val di Lima} was one of the three administrative districts into which the mountains were divided and was called \emph{fines Contronenses} \cite{qui02, cia06, cia11}.
 The only find from this period is the \textit{Grotta di Arzale}, a rock shelter that opens up northwest of \emph{Casoli} \cite{gia96}.
 
-The first attestation of the settlements of \emph{Casoli} and \emph{Lacu} (Lake) dates back to the 10th century \cite{gia96}.
+The first attestation of the settlements of \emph{Casoli} and \emph{Lacu} dates back to the 10th century \cite{gia96}.
 Most documents of this period show the fractioning and alienation of ecclesiastical heritage in favor of the city aristocracy, securing an accumulation of funds and power that was to form the basis of the subsequent domains \cite{qui02, gia96, for12, for15}.
@@
 Written sources mention a castle in \emph{Casoli} from 1180, but its foundation must be earlier.
-Between the 13th and 15th centuries, the fortification was at the center, along with the other castles of the \emph{Val di Lima}, of clashes between \emph{Lucca}, \textit{Pisa} and Florence, with alternating fortunes, as it was a strategic border area for the power of \emph{Lucca}.
-Today, very little is preserved and the area has been partly inhabited \cite{gia96, for12, rom16}.
+Between the 13th and 15th centuries, the fortification was at the center, along with the other castles of the \emph{Val di Lima}, of clashes between \emph{Lucca}, \textit{Pisa} and \emph{Firenze}, with alternating fortunes, as it was a strategic border area for the power of \emph{Lucca}.
+Today, very little is preserved and the area is partly inhabited \cite{gia96, for12, rom16}.
 
@@
 Initially dedicated only to \textit{San Donato}, it acquired a double dedication after the abandonment of the church of \textit{Sant'Andrea de Lacu}.
-The latter is located on a plateau to the east of Lake \emph{Casoli} and is attested from 1260 \cite{ber18, gia96, cap17}, but already in the 15th century, we learn of its state of abandonment \cite{gia96, con12}.
-This was probably due to the depopulation of the lake area and the simultaneous strengthening of the town of \emph{Casoli}, which was fortified and better protected during this unstable phase.
+The latter is located on a plateau to the east of \emph{Casoli} lake and is attested from 1260 \cite{ber18, gia96, cap17}, but we are aware that it was abandoned in the 15th century.
+\cite{gia96, con12}.
+This was probably due to the depopulation of the lake area and the simultaneous strengthening of the town of \emph{Casoli}, which was fortified and better protected during this unstable period.
 
 Today, the Romanesque church of \textit{Sant'Andrea de Lacu} is in a state of decay, with a rectangular plan, ending on the east side with a semi-circular apse.
-Inside, there is a reused element of the previous building, testifying to an older origin, probably contemporary with the ancient settlement of \textit{Lacu} (Lake) mentioned in sources from the 10th century and no longer visible today.
+Inside, there is a reused element of the previous building, testifying to an older origin, probably contemporary with the ancient settlement of \textit{Lacu}, meaning "lake" and mentioned in sources from the 10th century and no longer visible today.
 
@@
 Within \emph{Casoli}, a series of residential buildings with imposing dated portals are preserved, some with courtyards, and mansions that denote a discrete deployment of resources by wealthy social classes.
-Also dating from the modern era is the \textit{Madonna di Castello} Oratory, characterized by an entrance enclosure and located along the cobbled road that traces the ancient route between the village and the summit area of the medieval castle \cite{gia96}.
-
-Along the road that led toward \textit{Lucchio} crossing the area of \emph{Casoli} Lake, there is the chapel of \textit{Madonna di Col di Piano} and two “\textit{metati}”, buildings destined for chestnuts drying, referable to the contemporary age.
+Also dating from the modern era is the \textit{Madonna di Castello} oratory, characterized by an entrance enclosure and located along the cobbled road that traces the ancient route between the village and the summit area of the medieval castle \cite{gia96}.
+
+Along the road that led toward \textit{Lucchio} crossing the area of \emph{Casoli} lake, there is the chapel of \textit{Madonna di Col di Piano} and two “\textit{metati}”, buildings destined for chestnuts drying, referable to the contemporary age.
 Referred to as the "bread tree," chestnut fruits and the flour derived from them were a staple of the mountain diet until the mid-1900s \cite{buc92, puc10}.
@@
 
-Within the paper's aim framework and from a touristic point of view, the relevance of the touristic context of the territory is crucial in understanding the socio-cultural system of potential tourist destinations.
+The relevance of the touristic context of the territory is crucial in understanding the socio-cultural system of potential tourist destinations.
 In light of this and for a holistic study approach to social features of tourism, the host and traveler relationship is directly related to local development and local government systems \cite{amo21}.
-
-In this study, the holistic perception of local stakeholders and community empowerment (with its increasing potential) in tourism at the local level acquires a particular value.
-For this specific purpose, this paragraph is well related to the tourism dynamics of the Municipality of \emph{Bagni di Lucca}, the administrative land hosting \emph{Casoli}.
-
-\subsubsection{The tourism context of / The tourist dynamics in \emph{Bagni di Lucca}, the \emph{Val di Lima} and \emph{Casoli}}
+In this study, the perception and community empowerment in local tourism are crucial for future destination planning.
+On the socio-economic hand, it is fundamental to understand touristic data and indexes which are specifically related to \textit{Bagni di Lucca's} tourism dynamics, as the administrative land hosting \textit{Casoli}.
+In this perspective, the following paragraph concerns {Bagni di Lucca's} demography and touristic data analysis that will be depicted for local sustainable governance measures.
+
+ %As a result, we have expanded our horizons to include the Municipality of \textit{Bagni di Lucca}, which is the administrative land that hosts \textit{Casoli}.
+
+% Anche la prima frase andrebbe un pò chiarita, e non soon sicuro che quello che ho scritto rispecchi quello che vuoi dire :-). Qui sotto il tuo testo originale
+%For this specific purpose, this paragraph is well related to the tourism dynamics of the Municipality of \emph{Bagni di Lucca}, the administrative land hosting \emph{Casoli}.
+--> sono due concetti completamente diversi......
+
+\subsubsection{Tourism dynamics in \emph{Bagni di Lucca}, the \emph{Val di Lima} and\emph{Casoli}}
 
@@
 
-The demographic trend enlightens a relevant aspect of \emph{Bagni di Lucca} society, and the employment statistics of the population mirror the social, cultural, and economic aspects of its touristic ecosystem.
+In reason of a whole insight of the mentioned touristic area, a socio-cultural approach is needed, by considering also tourism flows.
+In this perspective, it must be considered that there isn’t touristic information at the administrative level of \emph{Casoli}, so the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT (National Statistical Institute).
+ The demographic trend (Figure 4, a) enlightens a relevant aspect of \emph{Bagni di Lucca} society and economic aspects of its touristic ecosystem (Figure 4, b).
 
@@
 Notably, the tourist density of 106.41 tourists per $km^2$ is higher than the population density (33.95 inhabitants per $km^2$).
-
-The number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay a number of nights more than a weekend and less then a week.
-On the other hands, this data means that people like staying in this touristic comprehensory and they may have founded attractions, services and activities they need.
-Consequently, a well-organized turistic system attracts potentially touristic presences widespread on the municipality territory.
+This information should be considered in local planning to enhance the number of residents.
+
+As shown in \ref{tab:tourism}, the number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay a number of nights more than a weekend and less than a week.
+On the other hand, this data means that people like staying in this tourist comprehensory and they may have found attractions, services and activities they need.
+Consequently, a well-organized touristic system attracts potentially touristic presences widespread on the municipality territory.
 
@@
 
-For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study with the purpose to highlight the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}.
-As concerns tourist indexes shown in Chart 3, the touristic density counts 1866, 96 people per square kilometer (1773 kmq), with an average stay of 3,36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2,98 nights).
-The provincial area of \emph{Lucca} shows an accommodation density of 19,30 on 10, 14 of \emph{Bagni di Lucca}, while the coexistence index of \emph{Lucca} (with 69,87 foreign tourists on 100 Italian tourists) is lower than \emph{Bagni di Lucca}’s coexistence index (with 103,74 foreign tourists in 100 Italian tourists).
+For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study with the purpose of highlighting the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}.
+As concerns tourist indexes shown in Chart 3, the touristic density counts 1866.96 people per square kilometer (1773 $km^2$), with an average stay of 3.36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2.98 nights).
+The provincial area of \emph{Lucca} shows an accommodation density of 19.30 on 10.14 of \emph{Bagni di Lucca}, while the coexistence index of \emph{Lucca} (with 69.87 foreign tourists on 100 Italian tourists) is lower than \emph{Bagni di Lucca}’s coexistence index (with 103.74 foreign tourists in 100 Italian tourists).
 Definitely, it can be claimed that \emph{Bagni di Lucca} is well positioned in the whole touristic context of \emph{Lucca} since it represents one of the less populated municipalities in the territory with interesting tourist indexes.
-In fact, \emph{Bagni di Lucca} boasts its historical thermal tourism tradition together with a well-equipped environment for sport and adventure tourism, on a surface of 164,70 kmq, that is one of the biggest municipality areas in the province of \emph{Lucca}.
+In fact, \emph{Bagni di Lucca} boasts its historical thermal tourism tradition together with a well-equipped environment for sport and adventure tourism, on a surface of 164.70 $km^2$, that is one of the biggest municipality areas in the province of \emph{Lucca}.
 
@@
 Thanks to such tourist flows \emph{Bagni di Lucca} is considered one of the main tourist areas of the \emph{Val di Lima} and the Serchio Valley.
-Several promotional websites advertise \emph{Bagni di Lucca} attractions\footnote{Among local websites, these two destination websites are: \url{www.turismobagnidilucca.com} and \url{www.valdilima.org}}: the sitography closing this paper contains a long but incomplete list that confirms such a statement, and Table \ref{tab:experience} exposes the promotional touristic system offering varied experiences in the \emph{Val di Lima} area.
+Several promotional websites advertise \emph{Bagni di Lucca} attractions\footnote{Among local websites, these two destination websites are: \url{www.turismobagnidilucca.com} and \url{www.valdilima.org}}: the sitography closing this paper contains a long yet incomplete list that confirms such a statement, and Table \ref{tab:experience} exposes the promotional touristic system offering varied experiences in the \emph{Val di Lima} area.
 The variety of outdoor touristic services dates back to the ‘90s when some associations began organizing mainly trekking excursions and rafting experiences.
@@
 
-%Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of Lucca and the Garfagnana mountain area.
-The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like Orrido di Botri, with The Canyon Park \footnote{\url{ https://www.canyonpark.it/}}, and other service companies and associations for experiential tourism.
+Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of Lucca and the \textit{Garfagnana} mountain area.
+
+%The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like Orrido di Botri, with The Canyon Park \footnote{\url{ https://www.canyonpark.it/}}, and other service companies and associations for experiential tourism.
 
@@
 Its territory features multiple kinds of touristic sights: from soft and hard outdoor experiential attractions (specifically the Canyon park activities like canyoning, rafting, and trekking experiences)) to Romanesque churches’ remains and huts.
-Qualitative studies oriented to tourist attractions, and widely, to the complexity of the \emph{semi-structured} tourism show that \emph{Casoli} exhibits unique features despite its tourism variety.
-
-There aren’t touristic information at the administrative level of \emph{Casoli} , so that the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT.
+Definitely, despite its variety of tourist attractions, \emph{Casoli} exhibits unique features in its comprehensive tourism.
+
+%There aren’t touristic information at the administrative level of \emph{Casoli} , so that the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT.
 
@@
 As part of the \emph{Bagni di Lucca}’s ecosystem with its touristic vocation, \emph{Casoli} is defined as a well-known destination.
-Given that \emph{Casoli} is a target of semi-structured tourism, and to preserve its peculiar social fabric, sustainability of development actions is of paramount importance.
+Considering that \emph{Casoli} is a destination for slow tourism, ensuring the sustainability of development actions is crucial to maintaining its unique social fabric..
 The latter aspect suggests paying attention to environmental conservation, cultural heritage conservation, intergenerational continuity, and economic equity \cite{kuh10}.
@@
 
-In \cite{dre22} Dianne Dredge describes the role of private and public entities in making tourism sustainable, which we summarize in the following actions:
+As described by \cite{dre22} Dianne Dredge and in light of the above analysis, private and public entities can contribute to sustainable tourism  by developing the following actions:
 
 %In this perspective, policy-makers play a key role in the planning of sustainable measures and in the monitoring of impacts; in particular, in the context of tourism, sustainability is a topic discussed in its many aspects, including resident ecosystem resilience approaches and regeneration actions by policy-makers and/or private tour operators \cite{dre22}.
-
-%In light of the above analysis, the following actions from public and private entities might foster sustainable development:
 
@@
 \item to foster an inclusive stakeholder approach for the tourism ecosystem 
-\item to promote a community-involved tourism vision 
+\item to promote a \textit{community-involved tourism} vision 
 \item to monitor and analyze tourism flows 
@@
 \item to develop potential touristic areas with unexpressed tourism appeal 
-\item to diversify tourism offer based on tourist provenance and service preferences 
+\item to diversify tourism offers based on tourist provenance and service preferences 
 \item to monitor tourism flows and tourist behaviors 
@@
 
-In this way, community-involved tourism can be a promising solution for sparsely populated areas with significant tourist attractions, such as the village of Casoli.
+In this way, \textit{community-involved tourism} can be a promising solution for sparsely populated areas with significant tourist attractions, such as the village of \emph{Casoli}.
 This type of tourism involves active participation and entrepreneurship from the local community to promote self-employment, community management, and stakeholder decision-making processes \cite{nag15}.
@@
 We include three stakeholder communities in the evaluation of the response to the initiative.
-We do not speak about {\em success} since the initiative is deliberately soft, and an outcome is expected in years, and the definition of the evaluation criteria rises new questions.
-So we are content to see if the stakeholders regarded the initiative as an intrusion, as helpful, or simply neutral.
+We do not speak about {\em success} since the initiative is deliberately soft, an outcome expected in years, and the definition of the evaluation criteria raises new questions.
+We are content with determining if the stakeholders viewed the initiative as intrusive, helpful, or simply neutral.
 
@@
 	\item {\em the residents} the local community that currently inhabits the site
-	\item {\em entrepreneurs} who currently have a business on the site
-	\item {\em administrators} that are in charge of managing the site resources and that, at due time, will respond to the two stakeholders above, and finally
+	\item {\em the entrepreneurs} who currently have a business on the site
+	\item {\em the administrators} that are in charge of managing the site resources and that, at due time, will respond to the two stakeholders above, and finally
 	\item {\em the users}, those that come to \emph{Casoli} to visit the Cave of La Piella and that we find the QR-codes on their way
@@
 
-The category of the users escapes an evaluation since the QR tag does not record reading operations.
-We added to the text message in the QR-tag an external link, that we monitor for the number of hits on pages dedicated to the specific tag: however, this feedback largely underestimates the number of times the tag is read.
+The QR tag does not record reading operations, preventing an evaluation of the user category.
+A URL was added to the text message in the QR tag to monitor the number of hits.: however, this feedback largely underestimates the number of times the tag is read.
 
@@
 
-As concerns residents, we had the opportunity to meet only two people living in \emph{Casoli}  and providing us a few information which was not useful for our study aims; local entrepreneurs we met informed us about the trekking tourism flows passing through \emph{Casoli}  and the Piella Cave path, which usually starts from the Canyon Park experience.
+Regarding residents, we had the chance to meet a few people living in \emph{Casoli}, which did not provide useful information; local entrepreneurs we met informed us about the trekking tourism flows passing through \emph{Casoli}  and the Piella Cave path, which usually starts from the Canyon Park experience.
 
@@
 
-In light of the above assumptions, a reflection on community-involved tourism perspective is needed.
+In light of the above assumptions, a reflection on \textit{community-involved tourism} perspective is needed.
 As shown by this empirical study, a small village like \emph{Casoli}  can require a valorization strategy taking advantage of the wider local tourism context.
 
-According to a holistic development vision, when a tourist destination meets its community issues it means that local stakeholders must balance a tourist industry vision with sustanability goals at a comprehensive perspective on environment, socio-cultural and economic long-term benefits, as described by George B.
-Et Alii, ()
+According to a holistic development vision, when a tourist destination meets its community issues it means that local stakeholders must balance a tourist industry vision with sustainability goals at a comprehensive perspective on the environment, socio-cultural and economic long-term benefits, as described by George B.
+Et Alii \cite {geo07}.
 
@@
 
-Literature on this scientific focus, as described by Richards G and Derek H. () gives evidence on local stakeholders' role as ‘designers’ of their living territory. 
+Literature on this scientific focus, as described by Richards G. and Derek H. \cite {ric03} gives evidence of local stakeholders' role as ‘designers’ of their living territory.
 This is particularly important in inland and marginal tourist areas that require a sustainable approach to development, rather than relying on mass tourism.
@@
 
-Definitely, community-involved tourism may represent the balanced intersection between community and development through policymakers' actions and decision makings (Jamal T. and Dredge, 2014). 
-In this view, the socio-political framework of tourism, involving governance measures, has to include environmental requalifying and destination planning (Sharabani et alii, 2020; Richards and Hall, 2003).
+Definitely, \textit{community-involved tourism} may represent the balanced intersection between community and development through policymakers' actions and decision-making \cite {jam14}.
+In this view, the socio-political framework of tourism, involving governance measures, has to include environmental requalifying and destination planning \cite {ric03}.
 
@@
 
-As explained in the previous section, the area around \emph{Casoli} is a good candidate for our case study because of its long history, which is witnessed by architectural and geological features.
+As explained in the previous section, the area around \textit{Casoli} is a suitable case study location due to its long history, which is reflected in its architectural and geological features.
 In the following, we define the relevant features and report on their accessibility.
@@
 
-Both caves are going to be analyzed in the course of the Underlandscape project \cite{Underlandscape}, and findings can further enhance the interest in both sites.
-
-Along the dirt road from the village in the direction of Lake \emph{Casoli} are some of the most interesting sites.
-The medieval remains of the {\em Church of Sant'Andrea de Lacu}, now in a state of serious disrepair, without its roof and with static problems, but characterized by its Romanesque forms and 10th-century decorations.
+Both caves are going to be analyzed in the course of the Underlandscape project \citeWeb{underlandscape, underlandscapeweb}, and findings can further enhance the interest in both sites.
+
+Along the dirt road from the village in the direction of the lake of \emph{Casoli} are some of the most interesting sites.
+The medieval remains of the church of {\em Sant'Andrea de Lacu}, now in a state of serious disrepair, without its roof and with static problems, but characterized by its Romanesque forms and 10th-century decorations.
 
@@
 
-A little further along the CAI path is the {\em Oratory of the Madonna di Col di Piano}. Recently renovated, with its canopy it has been a shelter for wayfarers since modern times.
-
-Inside the town of \emph{Casoli}, narrow streets branch off to the main square, where the medieval {\em Church of Saints Donato e Andrea} with its bell tower can be admired.
+A little further along the CAI path is the oratory named {\em Madonna di Col di Piano}. Recently renovated, with its canopy it has been a shelter for wayfarers since modern times.
+
+Inside the town of \emph{Casoli}, narrow streets branch off to the main square, where the medieval church dedicated to the Saints \emph{Donato} and \emph{Andrea} with its bell tower can be admired.
 From here, an ancient street climbs to the top of the hill.
@@
 
-On the summit are the remains of the {\em Medieval Castle} with its walls.
+On the summit are the remains of the medieval castle with its walls.
 In need of consolidation of the walls, it cannot be visited at the moment as it is on private property; it is only possible to observe part of its structures from the path.
@@
 
-The concept behind the implementation was that of guiding the visitor through a self-organized experience using a signage technique that improves the traditional boards, as explained in section \ref{sec:minimize}.
-To verify on the field the practical aspects and the implementation process we realized several QR tags that were later secured to an existing support.
-See the list in Table~\ref{tab:qrtags}.
-
-The proposed signage cannot replace CAI signage (see figure \ref{fig:CAI}), which is recognized throughout the country and is very effective.
-However, the new signage can effectively complement it to provide more detailed information at multiple levels, from short text to in-depth information on the project site.
-In addition, being able to read the written directions along the path can be of great help to inexperienced hikers, particularly on unclear and not-quite-visible trails such as the one we suggested to reach Buca La Piella.
+The concept behind the implementation of the QR-based signage is to guide the visitor through a self-organized experience.
+To verify on the field the practical aspects and deployment we realized and placed several QR tags.
+The list is in Table~\ref{tab:qrtags}.
+
+The proposed signage cannot replace CAI one (see figure \ref{fig:CAI}), which is recognized throughout the country and is very effective.
+The QR-based signage complements it and provides more detailed information at multiple levels.
+For one, the availability of written directions along the trail leading to the \textit{Buca La Piella}, which is hard to follow and not covered by CAI signage, is of great help to inexperienced hikers.
 
@@
 
-The tags are tied to pre-existing structures, like trees or poles, to keep the environmental footprint as lightweight as possible. The tag is tied with a thin biodegradable string.
-
-A QR tag is designed to implement three levels of reading (LoR), with an increasing level of technology involved:
+To minimize the environmental footprint, the tags are attached to pre-existing structures, like trees or poles. The tag is tied with a thin biodegradable string, as in figure \ref{fig:TagOnTree}.
+
+The QR tag is designed to implement three levels of reading (LoR), with an increasing level of technology involved:
 \begin{itemize}
     \item visual: the information is printed on the tag. The user does not need any technology to read the content;
-    \item QR tag text: this is the text encoded in the tag. The user needs a smartphone-like device and an appropriate app to read, possibly aloud, the content;
+    \item QR tag text: this is the text encoded in the tag. The user needs a smartphone-like device and an appropriate app to read aloud or translate the content;
     \item URL: these are Web URLs encoded in the tag text. The user needs an Internet connection to visit the URL.
@@
 
-We placed QR tags both along the trail and at sites of interest; the former are placed at detours and, at more complex points, at close range so that from one you can see the next, as per the CAI standard \cite{cai10}.
+We placed QR tags both along the trail and at sites of interest; the former are at detours or at close range so that from one you can see the next, as per the CAI standard \cite{cai10}.
 They can therefore have the function of a simple signpost to visually indicate the continuity of the trail.
@@
     \item URL LoR: such information is useful only in a few locations since the area is not uniformly covered by broadband networks: for instance, during a recon to the {\em Piella} cave, none of the smartphones of the participants received sufficient broadband network signal to visit the linked page.
-However, the application records the URL so that the user might visit it when entering a covered area.
-Each page contains site-specific information, an interactive map with the location of all the QR tags (see figure \ref{fig:linkmap}), and a form for user feedback (see figure \ref{fig:linkpoll}).
+However, the application records the URL so that the user can visit it when entering a covered area.
+Each Web page contains site-specific information, an interactive map with the location of all the QR tags (see figure \ref{fig:linkmap}), and a form for user feedback (see figure \ref{fig:linkpoll}).
 The map is hosted by the UMap Web service (\url{https://umap.openstreetmap.fr/it/}) \citeWeb{umap}, and displays the location and content of all QR-tags (as shown in figure \ref{fig:qrcodecontent}).
@@
 The finalized master document contains a single \emph{GeoJSON} \emph{FeatureCollection} object containing one \emph{Feature} object for each tag.
-Each \emph{Feature} contains a \emph{Geometry} object of type \emph{Point}, and a \emph{properties} object with fields containing information needed to create the tag: the \emph{uid}, a random string of six lowercase letters used to produce the associated URL, the \emph{title} printed on the tag, and the text encoded in the QR code (see table \ref{code:geojson}) where part of the {\em description} field is omitted.
+Each \emph{Feature} contains a \emph{Geometry} object of type \emph{Point}, and a \emph{properties} object with fields containing information needed to create the tag: its unique identifier \emph{uid}, a random string of six lowercase letters used to produce the associated URL, the \emph{title} printed on the tag, and the text to be encoded in the QR code (see figure \ref{fig:geojson-sample}).
 The \emph{properties} are also shown on the UMap map.
 
-Editing the basic map to fill-in the QR-code content and the other fields does not require technical skills since the tools integrated into the UMap user interface already fit the purpose (see figure \ref{fig:umap-editing}).
-The resulting geojson object for a tagged feature is shown in figure \ref{fig:geojson-sample}.
+Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the UMap user interface already fit the purpose (see figure \ref{fig:umap-editing}).
+The resulting \emph{GeoJSON} object describing a tagged feature is shown in figure \ref{fig:geojson-sample}.
 
@@
 \hfill
-\subfloat[GeoJSON \emph{Object} representing the same the feature]{
+\subfloat[\emph{GeoJSON} \emph{Object} representing the same the feature]{
 	\includegraphics[width = 0.60 \linewidth]{figure/geojson-sample.png}
@@
 }
-\caption{Editing the GeoJSON with uMap editor. The resulting file is automatically converted into printable QR-tags}
+\caption{Editing the \emph{GeoJSON} with uMap editor. The resulting file is automatically converted into printable QR-tags}
 \label{fig:editing}
@@
 
-A \emph{Bash} script running on a Linux system makes straightforward the conversion of the GeoJSON file into a printable array of QR tags.
-The code, a total of 90 source lines written in Bash and Python taking advantage of the ogr2ogr and qrencode commands, is on GitHub \citeWeb{github-qr}.
+A \emph{Bash} script running on a Linux system makes straightforward the conversion of the \emph{GeoJSON} file into a printable array of QR tags.
+The code, a total of 90 source lines written in Bash and Python makes use of the \texttt{ogr2ogr} and \texttt{qrencode} commands, and is on GitHub \citeWeb{github-qr}.
 
@@
 
-Our point is to illustrate a methodology to improve the touristic offer of a site that is not specific to a use case, but applicable to a broad range of frameworks.
-The methodology must be sustainable in the broader sense: social, environmental, and financial.
-
-We applied the methodology to a small-scale example: a signage system addressing the guidance of visitors in a rural area with a cultural heritage to be protected and valorized at the same time.
-
-Firstly we evaluated the aspects of interest of the site, and we found natural formations and human artifacts tracing back to the Paleolithic Era.
-Other relevant aspects are that the site is not included in mass tourism circuits but is not far from them.
-The area is exposed to depopulation due to the shortage of productive activities.
-
-To preserve the native social and financial framework we do not aim at mass tourism.
-Instead, we target a motivated, non-casual visitor, and devise potential reasons for its visit.
-We found many of them: a not easily accessible cave inhabited since pre-history, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
-
-We investigated the task of guiding the visitor and telling the history of the place with inexpensive and non-intrusive techniques.
+We aim to define a design approach that will enhance the tourist experience of an inland area.
+The approach is conceptual and therefore applicable to a wide range of use cases.
+First, the requirements for sustainability in design are defined, taking into account the three pillars of sustainability: social, environmental, and economic.
+
+To put our approach into practice we used a small-scale use case: a signage system addressing the guidance of visitors in an inland area.
+The area has a cultural and environmental heritage to be protected and valorized at the same time.
+
+Our design started by defining a profile for the area according to the \emph{three pillars}.
+The environmental aspect is characterized by natural formations and human artifacts tracing back to the Paleolithic Era.
+As for tourism-related economics, currently the dominant component in the area, we noticed that the location selected for our use case exhibits unexpressed potentialities although the area surrounding the location has a recognized touristic vocation.
+From the social point of view, we observed that the area is exposed to depopulation due to the shortage of productive activities.
+
+One conclusion after the above analysis is that, in order to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable.
+Regarding the potential reasons for a visit, we found a not easily accessible cave inhabited since pre-history, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
+
+We investigated a technique for guiding the visitor and telling the history of the place. The fragile economy limits financial investment, while the natural heritage requires non-intrusive techniques.
+
 The proposed solution consists of a dozen QR tags placed at key locations.
@@
 
-The preparation of the signage requires a solid knowledge of the site. This demanded consulting historical documents, assessing the touristic vocation of the area, and two on-site surveys.
-
-On the technical side, the proposed signage provides a benchmark difficult to improve in terms of environmental impact and cost, but its {\em performance} is difficult to evaluate.
-
-One performance index might be the number of times a given tag is scanned.
-Indeed, we recorded several hits to the URLs in the QR tags, but their number and the way they are collected are not appropriate to form a statistic of their utilization.
-One reason is the fact that the use of the tag is not always associated with a hit, and vice-versa.
+The preparation of the signage required a solid knowledge of the area that was gained by carrying out archaeological surveys and consulting historical documents and cartography.
+The evaluation of touristic vocation was based on the available statistical data and person-to-person informal interviews.
+Two on-site surveys provided further practical indications and the opportunity to place the QR tags.
+
+On the technical side, the proposed signage is a benchmark difficult to improve in terms of environmental impact and cost.
+
+%It fits cultural heritage assets that are not currently accessible, either because located on private property or showing unsafe structural conditions.
+The preservation of cultural heritage and its readiness for safe fruition may become difficult to attain in areas, like the hilly and mountainous terrain we studied, featuring a vast heritage scattered over a hardly accessible area.
+In these cases the QR tags improve the visitors experience by providing, beside textual data, also Web URLs for videos, aerial shots, or photos.
+Such multimedia resources immerse the visitor even more in the history of the area, and drive an experiential understanding.
+This copes with the duty of protection and conservation of abandoned sites, since positive enjoyment and widespread knowledge of the site develop, for an inland area undergoing depopulation, a virtuous loop between tourism and economy engine triggering investments in the cultural heritage policies.
+
+In areas like our use case, where the cultural heritage extends across scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging.
+In these cases, QR tags could improve the visitors' experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos.
+Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot.
+The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.
+
+Its {\em performance} is currently difficult to evaluate.
+The recording of the number of times a given tag is scanned, in theory, a significant performance index, is unreliable for technical reasons.
+In fact, the QR tag does not have the capability to record or transmit read events, and, on the side of the smartphone, the user may be unable to visit the URL in the tag or deny such operation for security reasons.
+In addition, the URL can be visited by means different from the QR reading.
+For this reason, we do not discuss such an index.
 
 The reactions we collected by interviewing local stakeholders are positive, which gives us the impression that the initiative is respectful and consistent with the social and economic framework.
-The measurement of the efficacy in financial terms is beyond the scope of this paper, but it has to be said that the aim of this work is to provide assistance and improve the experience of actual visitors, not to multiply their number.
+The measurement of the efficacy in financial terms is currently beyond the scope of this paper and will be assessable only in the long term depending on the collaboration of the involved stakeholders.
 
@@
 The current dimension of the tag is such that the visual impact is limited, but this negatively affects its efficacy: it is easy to miss a tag containing possibly relevant information;
-    \item Monitoring visitor's activity, like the frequency of visits of a given tag, or the sequencing of visited tags, which may be of interest for managing the site, depends on user motivation.
+    \item Monitoring visitor's activity, like the frequency of visits of a given tag, or the sequencing of visited tags, which may be of interest for managing the site, is currently ineffective.
 \end{itemize}
 
-Trading off cost and simplicity for effectiveness and measurability there are solutions, based, for instance, on Bluetooth communication or dedicated applications, that may improve such aspects.
+Trading off cost and simplicity for effectiveness and measurability there are solutions, based, for instance, on Bluetooth communication or dedicated applications, that might solve such issues.
 
@@
 
-Like any other productive activity, tourism is exposed to being unsustainable, damaging the very resources that allow it.
-To avoid heavy drawbacks the design of the activity has to consider the complex framework where it will operate, and the word used to indicate this is \emph{holistic}, and it is extremely difficult to put into practice.
-A methodology made of examples and guidelines simplifies the task.
+Tourism, like other productive activities, is at risk of being unsustainable and damaging the very resources that support it.
+To avoid heavy drawbacks the design of a tourism initiative has to consider the complex framework where it will operate according to an \emph{holistic} approach.
+A methodology made of examples and guidelines helps cope with such a challenging task.
 
 This paper is a step in this direction.
-We address a common problem, that of providing the visitor with directions in an area not reached by mass tourism, with the aim of acquiring the financial resources for its survival while preserving its distinguishing traits.
-
-We show how a balanced use of technology may help in the task, with a process that is necessarily multidisciplinary, joining skills coming from the diverse domains: sociology, economics, humanities and technical.
-The paper demonstrates how their interplay brings ad-hoc results targeting sustainability and efficacy.
+We address an ordinary problem, that of providing the visitor with directions, with the aim of promoting slow tourism and preserving distinguishing traits of an inland area.
+
+We show how a balanced use of technology may help in the task, with a process that is necessarily multidisciplinary, joining skills coming from the diverse domains: sociology, economics, humanities, and engineering.
+The paper demonstrates how their interplay returns ad-hoc tools targeting sustainability and efficacy.
 
@@
 
-The discussion regarding the deployed experiment indicates directions for future investigation on the technical side.
-For instance, the availability of non-intrusive long-range devices that send a message whenever sensing a presence might improve signals discoverability.
-The Bluetooth Low Evergy technology goes in that direction, but hidden tags with BLE and energy harvesting would have broken the economy constraints.
+On the technical side, the discussion regarding the deployed experiment indicates directions for future investigation.
+The definition of a potential application will foster research aiming at passive or very low-energy devices that are able to effectively advertise their presence and collect statistics.
 
@@
 Policy-makers are crucial in the development and monitoring of sustainable measures.
-This is especially important in the tourism industry where sustainability is a multifaceted topic that includes resident ecosystem resilience and regeneration efforts by both policy-makers and private tour operators.
+This is especially important in the tourism industry where sustainability is a multifactor topic that includes resident ecosystem resilience and regeneration efforts by both policy-makers and private tour operators.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{For research articles with several authors, a short paragraph specifying their individual contributions must be provided.
-The following statements should be used ``Conceptualization, X.X.
-and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y.
-and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing---original draft preparation, X.X.; writing---review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y.
-All authors have read and agreed to the published version of the manuscript.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation.
+\authorcontributions 
+{Conceptualization, C.A., D.M.G.
+and C.L.; methodology, C.A., D.M.G.
+and C.L.; software, C.A.; validation, C.A., D.M.G.
+and C.L.; investigation, C.A., D.M.G.
+and C.L.; data curation, C.A.
+and D.M.G.; writing---original draft preparation, C.A.; writing---review and editing, C.A., D.M.G.
+and C.L.; supervision, C.A.; historical-archaeological framework, C.L.; touristic data analysis, community-involved tourism and guidelines for sustainable tourism, D.M.G.
+All authors have read and agreed to the published version of the manuscript.}
+
+%Please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation.
 Authorship must be limited to those who have contributed substantially to the work~reported.}
 
-\funding{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''.
-Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}
-
-\institutionalreview{In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study.
+\funding{This research was funded by the Italian "Ministero dell’Istruzione, dell’Università e della Ricerca" under project PRIN 2020 "Underlandscape" (2020428LS8)}
+
+%\institutionalreview{In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study.
 You might choose to exclude this statement if the study did not require ethical approval.
@@
 
-\informedconsent{Any research article describing a study involving humans should contain this statement.
+%\informedconsent{Any research article describing a study involving humans should contain this statement.
 Please add ``Informed consent was obtained from all subjects involved in the study.'' OR ``Patient consent was waived due to REASON (please provide a detailed justification).'' OR ``Not applicable'' for studies not involving humans.
@@
 
-Written informed consent for publication must be obtained from participating patients who can be identified (including by the patients themselves).
+%Written informed consent for publication must be obtained from participating patients who can be identified (including by the patients themselves).
 Please state ``Written informed consent has been obtained from the patient(s) to publish this paper'' if applicable.}
 
-\dataavailability{We encourage all authors of articles published in MDPI journals to share their research data.
-In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study.
-Where no new data were created, or where data is unavailable due to privacy or ethical re-strictions, a statement is still required.
-Suggested Data Availability Statements are available in section “MDPI Research Data Policies” at \url{https://www.mdpi.com/ethics}.}
-
-\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections.
-This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}
-
-\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflict of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results.
-Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section.
-If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the~results''.}
+\dataavailability{Statistics referenced in the paper are publicly available through the referenced web portals. The software mentioned in the paper is available on GitHub \citeWeb{github-qr}}
+
+\acknowledgments{The valuable experience Monica Baldassarri has been fundamental in the unfolding of our research. Enrica Salvatori helped in getting in touch with local administrators.}
+
+\conflictsofinterest{The authors declare no conflict of interest.}
 
@@
 %% Optional
-\sampleavailability{Samples of the compounds ... are available from the authors.}
+%\sampleavailability{Samples of the compounds ... are available from the authors.}
 
@@
 \begin{tabular}{@{}ll}
-MDPI & Multidisciplinary Digital Publishing Institute\\
-DOAJ & Directory of open access journals\\
-TLA & Three letter acronym\\
-LD & Linear dichroism
+URL & Uniform Resource Locator\\
+QR & Quick Response\\
+GeoJSON & Geographical JavaScript Object Notation\\
+CAI & Club Alpino Italiano\\
+UNESCO & United Nations Educational, Scientific and Cultural Organization\\
+SCC & Single Chip Computer\\
+NFC & Near Field Communication\\
+EEC & European Economic Community\\
+SCI & Sites of Community Interest\\
+SAC & Special Areas of Conservation\\
+SPA & Special Protection Area\\
+ISTAT & Istituto Nazionale di Statistica
 \end{tabular}

### 2023-11-04 figure/datiMG/script.sh

@@ -1,0 +1,22 @@
+#!/bin/bash -e
+gnuplot <<- EOF
+  set xlabel "Year"
+  set ylabel "Number of people"
+  set yrange [5000:7000]
+  set terminal png font "Times-New-Roman,20"
+  set output "population.png"
+  plot "population.txt" using 1:2 title "Population" lw 3 lt 3 with lines
+EOF
+
+gnuplot <<- EOF
+  set xlabel "Year"
+  set ylabel "Number of people"
+  set yrange [0:25000]
+  set style fill solid 0.5
+  set boxwidth 0.3
+  set terminal png font "Times-New-Roman,20"
+  set output "tourism.png"
+  plot 'tourism.txt' using (\$1-0.17):2 title "Arrivals" lw 3 lt 4 with boxes,\
+  'tourism.txt' using (\$1+0.17):3 title "Presence" lw 3 lt 3 with boxes,\
+  'tourism.txt' using 1:4 title "Residents" lw 3 lt 1 with lines
+EOF

### 2023-11-04 paper.tex
Commit riportato dal 3 novembre - Revisore 2 - Augusto

@@
 \definecolor{delim}{RGB}{20,105,176}
-\colorlet{numb}{magenta!60!black}
 
 \lstdefinelanguage{json}{
-    basicstyle=\normalfont\ttfamily,
+    basicstyle=\normalfont\ttfamily\footnotesize,
     numbers=left,
@@
     frame=lines,
-    backgroundcolor=\color{background},
-    literate=
-     *{0}{{{\color{numb}0}}}{1}
-      {1}{{{\color{numb}1}}}{1}
-      {2}{{{\color{numb}2}}}{1}
-      {3}{{{\color{numb}3}}}{1}
-      {4}{{{\color{numb}4}}}{1}
-      {5}{{{\color{numb}5}}}{1}
-      {6}{{{\color{numb}6}}}{1}
-      {7}{{{\color{numb}7}}}{1}
-      {8}{{{\color{numb}8}}}{1}
-      {9}{{{\color{numb}9}}}{1}
-      {:}{{{\color{punct}{:}}}}{1}
-      {,}{{{\color{punct}{,}}}}{1}
-      {\{}{{{\color{delim}{\{}}}}{1}
-      {\}}{{{\color{delim}{\}}}}}{1}
-      {[}{{{\color{delim}{[}}}}{1}
-      {]}{{{\color{delim}{]}}}}{1},
+    backgroundcolor=\color{background}
 }
@@
 % Authors, for the paper (add full first names)
-\Author{ Maria Grazia Deri $^{2,}$*\orcidC{}, Letizia Chiti $^{3}$\orcidA{}, Augusto Ciuffoletti $^{1}$\orcidB{} }
+\Author{ Maria Grazia Deri $^{2,}$\orcidC{}, Letizia Chiti $^{3}$\orcidA{}, Augusto Ciuffoletti $^{1}$*\orcidB{} }
 
@@
 This creates a reinforcing effect that follows the negative dynamics seen above: a leisure activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact.
-Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a}.
+Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a} relating to outdoor winter sports.
 
@@
 
+{\huge Anticipare qui uso dei QR code}
+
 \section{Unlocking site potentials through signage \label{sec:signage}}
@@
 The signage used to guide and inform the visitor is of paramount importance and must be carefully designed to accomplish its functions while following the guidelines of a sustainable approach.
+
+The functions of signage are two:
+\begin {itemize}
+\item an effective signage must guide the tourist across the resort.
+A map represents a starting point for this purpose, but not all visitors feel comfortable reading a map.
+The ideal signage should provide visible reference points and suggest actions, like "turn right after crossing the stream".
+A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}).
+The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail;
+\item a sign should explain the reasons for interest in a site.
+This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit (see figure \ref{fig:pole-direction}).
+In this sense, the CAI signal above is not sufficient.
+A simple board with a site name fits only when the site is sufficiently popular.
+Otherwise, a more structured message should be used to explain its relevance.
+\end{itemize}
 
@@
 
-The functions of signage are two:
-\begin {itemize}
-\item an effective signage must guide the tourist across the resort.
-A map represents a starting point for this purpose, but not all visitors feel comfortable reading a map.
-The ideal signage should provide visible reference points and suggest actions, like "turn right after crossing the stream".
-A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}).
-The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail;
-\item a sign should explain the reasons for interest in a site.
-This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit (see figure \ref{fig:pole-direction}).
-In this sense, the CAI signal above is not sufficient.
-A simple board with a site name fits only when the site is sufficiently popular.
-Otherwise, a more structured message should be used to explain its relevance.
-\end{itemize}
-
-In our study, we consider a range of solutions that support sustainable tourism, specifically the dynamic provision of information during a visit along a natural trail not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}).
+In our study, we consider a range of solutions that support sustainable tourism, specifically the dynamic provision of information during a visit along a natural trail not covered by Internet connectivity.
 We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.
@@
 	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons;
-	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well;
+	\item accessibility for foreign visitors: to limit the size of the plaque, the number of translations must be limited as well;
 	\item update limits: to update the content, the plaque must be replaced;
@@
 
-A popular research direction covers the social aspects related to the use of the smartphone.
+A popular research direction covers the social aspects related to the use of the smartphones.
 W.
@@
 Once a range of relevant solutions is identified, we proceed to the empirical part: a {\em proof of concept} implementation that verifies the feasibility of a specific solution.
+
+Regarding the specific technology we refer to, QR codes have been invented in 1994 for industrial applications.
+The current version, described in \cite{isoqr}, is dated February 2015.
+In 2013 an early paper by Martinez et al.
+\cite{mar13a} suggests its utilization to implement a virtual tour in the {\em Las Quilamas Natural Park} in Spain.
+At that time the availability of QR-code readers was not as diffused as today, and therefore the work may be considered among the seminal ones.
+The authors proposed to use the QR code as a link to a panel in PDF format viewed on the user's portable device.
+A similar work in a urban environment is presented in \cite{fin13a} and \cite{tat15a}, the latter illustrating the QR system serving the monuments in the Serbian town of Niš.
+
+Since then the use of the QR code technology for heritage promotion has been suggested in other papers.
+In \cite{sol18a} the authors describe its application in a museum: after defining the points of interest, tags have been placed and linked to related documents.
+The authors conducted a poll about user perception on two segments of users: Italians and Polish.
+The QR tags were not unanimously considered useful: in particular half of the Italian users considered them as not useful.
+It is interesting to note that the motivation of the other segment of users, less skeptical about usefulness, shared a solid educational motivation to access QR-linked content (70\% in the Polish group against 50\% in the Italian one).
+This fact sheds light on the necessity to provide content the user needs: curiosity is not a sufficient motivation.
+
+Recent publications about the use of QR codes for heritage valorization do not show significant novelties.
+In 2023, the authors of \cite{pan23a} introduce the installation of six QR codes in a museum and analyze the reactions of 42 visitors.
+The tags are created using the Taplink \url{https://taplink.at/} web service and are linked to content on the Web.
+
+There are examples of urban outdoor applications of QR tags, like
+
+All the found publications share a common attitude, which can be summarized in the motto "analog portal to digital world", as in the title of \cite{bai12a}.
+None of them evaluates the possibility of storing text in the tag itself, thus eliminating the requirement of an Internet connection.
+
+The present paper promotes the use of QR codes with sustainability and as the primary or only way of providing information to the visitor.
+In the literature, QR codes are considered ancillary to other forms of communication, like physical installations or the Web.
+Therefore, in our case the informational content of the tag, which is commonly ignored, becomes relevant.
+It is a new and original way of using QR codes for the delivery of information.
 
@@
 	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation,
-	\item{\em Tolerance for error}: there are no margins to use the device in a way that compromises user safety.
-The deliberate or accidental release of the passive device into the environment determines a minor pollution,
+	\item{\em Tolerance for error}: there is no chance of accidentally causing a hazardous situation using the QR tag.
+The unintended release of the passive device into the environment determines a minor pollution,
 	\item{\em Low physical effort} holds, although the user needs to carry a smartphone,
@@
 
-There are two relevant trade-offs that the host needs to resolve.
-One is related to the visibility of the tag.
-The trade-off is between visual impact and visibility.
-The other is related to monitoring tag utilization.
-This may be useful for all sorts of planning activities and is discussed in section \ref{sec:results}.
+The host needs to resolve two issues.
+One problem is the balance between visual impact and visibility.
+Monitoring tag utilization, which is useful for all types of planning activities, is the other topic.
+They are discussed in section \ref{sec:results}.
 
@@
 
-In recent years, the main tourist attractions have concentrated on the \emph{Lima} stream, with some associations and private entities promoting outdoor experiences, particularly fluvial sports, such as canyoning, rafting, and sup.
+In recent years, the main tourist attractions have concentrated on the \emph{Lima} stream, with some associations and private entities promoting outdoor experiences, particularly fluvial sports, such as canyoning, rafting, and paddleboarding.
 Other important tourist attractions are trekking and hiking, supported by a network of paths tracked by local CAI (\textit{Club Alpino Italiano}).
@@
 
-\subsubsection{Tourism dynamics in \emph{Bagni di Lucca}, the \emph{Val di Lima} and\emph{Casoli}}
+\subsubsection{Tourism dynamics in \emph{Bagni di Lucca}, the \emph{Val di Lima} and \emph{Casoli}}
 
@@
 As part of the \emph{Bagni di Lucca}’s ecosystem with its touristic vocation, \emph{Casoli} is defined as a well-known destination.
-Considering that \emph{Casoli} is a destination for slow tourism, ensuring the sustainability of development actions is crucial to maintaining its unique social fabric..
+Considering that \emph{Casoli} is a destination for slow tourism, ensuring the sustainability of development actions is crucial to maintaining its unique social fabric.
 The latter aspect suggests paying attention to environmental conservation, cultural heritage conservation, intergenerational continuity, and economic equity \cite{kuh10}.
@@
 \subfloat[QR-tag parts]{
-	\includegraphics[width = 0.55 \linewidth]{figure/QR-code-legend.png}
+	\includegraphics[width = 0.60 \linewidth]{figure/QR-code-legend.jpeg}
 	\label{fig:qrcodelegend}
@@
 \subfloat[QR-tag content on a 5.5' screen]{
-	\includegraphics[width = 0.35 \linewidth]{figure/qr-code-content}
+	\includegraphics[width = 0.30 \linewidth]{figure/qr-code-content.jpg}
 	\label{fig:qrcodecontent}
@@
 
+%\begin{figure}
+%\subfloat[Linked page: the map]{
+%	\includegraphics[width = 0.60 \linewidth]{figure/webpage-map}
+%	\label{fig:linkmap}
+%}
+%\subfloat[Editing a feature associated with a QR-tag in UMap]{
+%	\includegraphics[width = 0.35 \linewidth]{figure/umap-editing.png}
+%	\label{fig:umap-editing}
+%}
+%\subfloat[Linked page: survey form]{
+%	\includegraphics[width = 0.30 \linewidth]{figure/webpage-poll.png}
+%	\label{fig:linkpoll}
+%}
+%\caption{Content of the Web page linked to the URL in a QR tag}
+%\label{fig:webpage}
+%\end{figure}
+
 \begin{figure}
-\subfloat[Linked page: the map]{
-	\includegraphics[width = 0.60 \linewidth]{figure/webpage-map}
-	\label{fig:linkmap}
-}
-\subfloat[Linked page: survey form]{
-	\includegraphics[width = 0.30 \linewidth]{figure/webpage-poll.png}
-	\label{fig:linkpoll}
-}
-\caption{Content of the Web page linked to the URL in a QR tag}
-\label{fig:webpage}
+\includegraphics[width =\linewidth]{figure/combo_gen.jpeg}
+\caption{Generation of the QR tag and UMap visualization from UMap content.
+The uMap editor (bottom left) provides a simplified access to the GeoJSON description.
+The same interface allows to specify the template controlling the Web visualization of the map (right side).
+An ad-hoc application referenced in the paper allows to transform the uMap file downloaded from the server into a batch of printable QR tags (top left).
+The tag is highlighted in red in the picture on the Web display}
 \end{figure}
@@
         \hline 
-        \textbf{Name} & \textbf{Long} & \textbf{Lat} & \textbf{URL Key} & \textbf{Length} \\ \hline
+        \textbf{Name} & \textbf{Long} & \textbf{Lat} & \textbf{FID} & \textbf{Length} \\ \hline
         Buca La Piella & 10.68361 & 44.03667 & t5ysrm & 476 \\ \hline
@@
 \caption{The QR-tags placed around \emph{Casoli} .
-The URL key is a random code that appears on the web page linked to the QR code.
+The FID (Feature IDentifier) is a random code that appears on the web page linked to the QR code.
 The Length refers to the description in the QR tag and is in characters.
@@
 The finalized master document contains a single \emph{GeoJSON} \emph{FeatureCollection} object containing one \emph{Feature} object for each tag.
-Each \emph{Feature} contains a \emph{Geometry} object of type \emph{Point}, and a \emph{properties} object with fields containing information needed to create the tag: its unique identifier \emph{uid}, a random string of six lowercase letters used to produce the associated URL, the \emph{title} printed on the tag, and the text to be encoded in the QR code (see figure \ref{fig:geojson-sample}).
+Each \emph{Feature} contains a \emph{Geometry} object of type \emph{Point}, and a \emph{properties} object with fields containing information needed to create the tag: its unique identifier \emph{uid}, a random string of six lowercase letters used to produce the associated URL, the \emph{title} printed on the tag, and the text to be encoded in the QR code (see figure \ref{fig:geojson-sample})
+
+\begin{table}
+\begin{lstlisting}[language=json]
+{
+  "type": "Feature",
+  "geometry": { 
+    "type": "Point", 
+    "coordinates": [10.67592,44.035298]
+  },
+  "properties": {
+    "fid": "fxq83v",
+    "name": "Sant'Andrea de Lacu",
+    "description": "Secondo le ... ad uso abitativo.",
+    "imageTag": "1m0Vad1Om2ihSVezPUkMKfuCUlg3ptC6z",
+    "urlPrefix": "https://sites.google.com/ ..."
+  }
+}
+\end{lstlisting}
+\caption{\emph{GeoJSON} \emph{Object} of a feature associated with a QR-code. Ellipsis (...) are used to shorten long strings  \label{fig:geojson-sample}}
+\end{table}
+
 The \emph{properties} are also shown on the UMap map.
@@
 
-\begin{figure}
-\subfloat[Editing a feature associated with a QR-tag in UMap]{
-	\includegraphics[width = 0.35 \linewidth]{figure/umap-editing.png}
-	\label{fig:umap-editing}
-}
-\hfill
-\subfloat[\emph{GeoJSON} \emph{Object} representing the same the feature]{
-	\includegraphics[width = 0.60 \linewidth]{figure/geojson-sample.png}
-	\label{fig:geojson-sample}
-}
-\caption{Editing the \emph{GeoJSON} with uMap editor. The resulting file is automatically converted into printable QR-tags}
-\label{fig:editing}
-\end{figure}
+%\begin{figure}
+%\subfloat[Editing a feature associated with a QR-tag in UMap]{
+%	\includegraphics[width = 0.35 \linewidth]{figure/umap-editing.png}
+%	\label{fig:umap-editing}
+%}
+%\hfill
+%\subfloat[\emph{GeoJSON} \emph{Object} representing the %same the feature]{
+%	\includegraphics[width = 0.60 \linewidth]{figure/geojson-sample.png}
+%	\label{fig:geojson-sample}
+%}
+%\caption{Editing the \emph{GeoJSON} with uMap editor. The resulting file is automatically converted into printable QR-tags}
+%\label{fig:editing}
+%\end{figure}
 
@@
 
-The proposed solution consists of a dozen QR tags placed at key locations.
-The total cost of the tags amounts to a few Euros for the printing service, although finding a shop with an appropriate printing device may be difficult.
+The proposed solution consists of placing at selected locations the QR tags listed in table \ref{tab:qrtags}.
+The total cost of the tags amounts to a few Euros for the printing service, although finding a shop with an appropriate device may be difficult.
 The impact on the environment is extremely limited, as shown in figure \ref{fig:TagPlacement}.
@@
 
-Trading off cost and simplicity for effectiveness and measurability there are solutions, based, for instance, on Bluetooth communication or dedicated applications, that might solve such issues.
+Trading off cost and simplicity for effectiveness and measurability there are alternative solutions.
+
+One technology that fits the requirements of our task is Bluetooth Low Energy (BLE): the autonomy of a battery-operated BLE device is compatible with the periodic maintenance of other forms of signage.
+A Bluetooth device can send to the visitor's smartphone a beacon to indicate the presence of a passive sign, like a not-so-visible QR tag, or directly deliver information.
+However this requires the installation of the app on the visitor, and the design of the IT product including the device and the application, which may cost far more than the physical devices.
+
+A different application-based solution privileging passive devices is based on a writable NFC tag close to the one delivering the content.
+To update the number of visitors that have read the informational NFC, the user application reads and increments the count recorded on the writable NFC.
+Here too the primary drawback compared with our QR-tag based solution is the need for an ad-hoc application.
+In addition the event of data pollution is not excluded.
+
+The demonstration of the existence of such needs biases the technological evolution: for instance, if counting NFC reads were a frequent requirement, tags with such a feature would be produced at low costs, and the increment operation might become a default in NFC reading apps.
+The research on low-energy and low-cost devices is also that of demonstrating that the market for such devices is expanding.
 
@@
 
-The resulting implementation is original in the literature, and minimalistic in its deployment. The detailed description enables its reuse as a starting point in future initiatives.
+The resulting implementation is minimalistic in its deployment. The detailed description enables its reuse as a starting point in future initiatives.
 

### 2023-11-04 paper.tex
Sincronizzato con - Ancora modifiche augusto

@@
 Even worse, such capacities are often used with the aim of short-term local effects, ignoring long-term global ones, with the risk of long-term global deterioration.
-This attitude pervades all activities, starting with the most basic ones such as the supply of food and homes, and is often influenced by economic profit.
-
-Leisure activities play an important role in bringing profit to niche businesses.
+This attitude pervades all activities, starting with the most basic ones as the supply of food and homes, and is often influenced by economic profit.
+
+Leisure activities play are relevant in bringing profit to niche businesses.
 This creates a reinforcing effect that follows the negative dynamics seen above: a leisure activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact.
@@
 
-The promotion of cultural heritage is situated in such a complex scenario.
+The promotion of cultural heritage, which the Agenda 21 for Culture identifies as a fourth pillar for sustainability \cite{fourthpillar}, is situated in such a complex scenario.
 As noted by Jorge Otero in \cite{ote22a}, local communities benefit from the promotion of heritage: conservation activity generates social growth, while tourism provides an extra income that can be reinvested in promotion.
@@
 Its {\em performance} is currently difficult to evaluate.
-The recording of the number of times a given tag is scanned, in theory, a significant performance index, is unreliable for technical reasons.
-In fact, the QR tag does not have the capability to record or transmit read events, and, on the side of the smartphone, the user may be unable to visit the URL in the tag or deny such operation for security reasons.
-In addition, the URL can be visited by means different from the QR reading.
-For this reason, we do not discuss such an index.
+The recording of the number of times a given tag is scanned, in theory, a significant performance index, is unreliable for technical reasons, one being that the QR tag cannot record or transmit read events.
+Another is that the user may be unable or unwilling to visit the URL in the tag.
+Finally, the URL can be visited by means different from the QR reading.
+For such reasons, despite we acknowledging its importance for the stakeholders, we do not discuss such an index.
 

### 2023-11-07 paper.tex
txs auto checkin

@@
 
-This document provides a conceptual solution for a specific aspect: heritage trail signage.
-The solution is conceptual in that it tries to follow a guideline that takes into account many aspects and provides a practical way that is adaptable to other environments.
-
-{\huge Anticipare qui uso dei QR code}
+This document addresses the specific aspect of heritage trail signage, and describes a solution based on QR code tags.
+The design gathers suggestions from several disciplines to take into account the many aspects that contribute to sustainability.
+The result is a practical signage technology that is adaptable to other environments sharing defined traits.
 
@@
  \item a sign should explain the reasons for interest in a site.
-This includes featuring the site where the sign is installed, but also other nearby points of interest that invite the visitor to continue the visit (see figure \ref{fig:pole-direction}).
+This generic characterization includes all sorts of semantic characterization of the site, and may have educational or motivational intents.
+This includes describing historical facts related with the site, as well as advertising nearby points of interest (see figure \ref{fig:pole-direction}).
 In this sense, the CAI signal above is not sufficient.
 A simple board with a site name fits only when the site is sufficiently popular.
-Otherwise, a more structured message should be used to explain its relevance.
+Otherwise, a more structured message should be used to pursue an educational intent.
 \end{itemize}
@@
 
+This section outlines the historical framework that explains the selection of QR tag locations and their associated content.
+It is based on previous archaeological and historical investigations.
+The sites mentioned in the narration can be found among the QR codes listed in table \ref{tab:qrtags}.
+
 The village of \emph{Casoli} is located south of the \emph{Lima} stream on a hill named {\em "Tanette"}.
@@
 In Italy, a municipality is the smallest geographical-administrative unit offering accessible data about tourism, which are the fundamentals of our research.
-The municipality of \emph{Bagni di Lucca} boasts historical popularity for its landscapes told by famous poets (such as Giosuè Carducci, Giovanni Pascoli, Eugenio Montale, George Gordon Byron, Percy Bysshe Shelley) and celebrities like Paolina Bonaparte.
+
+%The municipality of \emph{Bagni di Lucca} boasts historical popularity for its landscapes told by famous poets (such as Giosuè Carducci, Giovanni Pascoli, Eugenio Montale, George Gordon Byron, Percy Bysshe Shelley) and celebrities like Paolina Bonaparte.
 The typical Liberty style architecture characterizes the villas like the \textit{Real Casinò} featuring elegant gardens representing the power of the Republic of \emph{Lucca} in the early XIX century and the \emph{Belle Epoque}.
@@
     \end{tabular}
-    \caption{Tourist numbers in \emph{Bagni di Lucca} in 2022\tablefootnote{Each index gives a different type of touristic information about \emph{Bagni di Lucca} Municipality: the touristic density represents the number of tourists per $km^2$; the Average stay index is the average number of nights spent in town by tourists; the Accommodation Density Index is related to the number of touristic presences on the number of beds occupied by tourists all over the year; the coexistence index shows the relevance of foreign tourists per 100 Italian tourists.}}
+    \caption{Tourist indexes in \emph{Bagni di Lucca} in 2022\tablefootnote{Each index gives a different type of touristic information about \emph{Bagni di Lucca} Municipality: the touristic density represents the number of tourists per $km^2$; the Average stay index is the average number of nights spent in town by tourists; the Accommodation Density Index is related to the number of touristic presences on the number of beds occupied by tourists all over the year; the coexistence index shows the relevance of foreign tourists per 100 Italian tourists.}}
     \label{tab:tourism}
@@
 
-Summing up, the quantitative data give evidence of a vibrant and attractive tourist reality constantly attracting tourists during the last decade.
+Several promotional websites advertise Bagni di Lucca attractions such as these two destination websites: www.turismobagnidilucca.com and www.valdilima.
+org; the sitography closing this paper contains a long yet incomplete list that confirms such a statement.
+
+From the point of view of our research, the context of Bagni di Lucca helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well as touristic services).
+Indeed, Val di Lima can promote a very identifying destination brand straddling the town of Lucca and the Garfagnana mountain area.
+The Municipality of Bagni di Lucca and the tourist area of the Val di Lima are the context for the small and fascinating village of Casoli, which is one of the 31 villages belonging to the Municipality.
+It is 7.22 km far from Bagni di Lucca, at an altitude of 500 meters.
+
+Casoli village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as: canyoning, rafting, and trekking experiences in the Canyon park.
+Additionally, it features Romanesque churches and huts attracting visitors looking for various experiences such as thermal, environmental, outdoor, sport and cultural activities.
+Throughout the year, local stakeholders offer a variety of services, often incorporating sustainable tourism practices.
+After a detailed analysis of the tourist and social context of Bagni di Lucca, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{x} dating back to the ‘90s, when some associations began organizing mainly trekking excursions and rafting experiences (\citeWeb{agu,r2o,ronda,garfraft,valdilima,canyonpark,experience,ebikea,ebikel,bdlt2,bdl}).
+
+As part of Bagni di Lucca's ecosystem, with its focus on tourism, Casoli is a destination for slow travel.
+It is important to ensure that any development actions taken are sustainable, in order to preserve its unique social fabric.
+This means paying attention to conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{x}.
+
+%Summing up, the quantitative data give evidence of a vibrant and attractive tourist reality constantly attracting tourists during the last decade.
 Thanks to such tourist flows \emph{Bagni di Lucca} is considered one of the main tourist areas of the \emph{Val di Lima} and the Serchio Valley.
@@
 The variety of outdoor touristic services dates back to the ‘90s when some associations began organizing mainly trekking excursions and rafting experiences.
-
-From the point of view of our research, the context of the \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well touristic services).
-
-After a detailed analysis of the touristic and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities  \cite{loz09}.
-
-Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of Lucca and the \textit{Garfagnana} mountain area.
-
-%The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like Orrido di Botri, with The Canyon Park \footnote{\url{ https://www.canyonpark.it/}}, and other service companies and associations for experiential tourism.
-
-\begin{table}
-    \centering
-    \begin{tabular}{|p{70mm}||p{45mm}|} \hline 
-         {\bf Touristic promotion }&  {\bf Kind of service}\\
-         \hline \hline 
-         Aguarajà River Experience \newline \url{https://aguaraja.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
-         \hline 
-         Rafting H2O \newline \url{https://www.raftingh2o.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
-         \hline 
-         RocKonda \newline \url{https://www.rockonda.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ \hline 
-         A.S.D. Garfagnana Rafting \newline \url{https://www.garfagnanarafting.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ 
-        \hline 
-         Val di Lima Off-Road \newline \url{https://www.valdilimaoffroad.com} & Quad, enduro\\ \hline 
-         Canyon Park \newline \url{https://www.canyonpark.it/canyon-sup}& Canyoning, aqua-trekking, stand-up padding, mental care and yoga, eco laboratory\\
-         \hline 
-         Pro Loco Bagni di Lucca \newline  \url{https://www.turismobagnidilucca.com/} \newline \url{https://www.valdilima.org/}& Sportfishing, mental care, trekking, survival, paragliding, horse riding, diving\\
-         \hline 
-         eBike Adventure Tour \newline \url{http://www.ebikeadventuretour.it/}& Ebike, Mountain bike\\ \hline 
-         Val di Lima eBike \newline  \url{https://valdilimaebike.it/} & Mountain e-bike, Quad, Enduro, Paintball\\
-         \hline 
-        Terme Bagni di Lucca \newline \url{https://www.termebagnidilucca.it/} &Spa, beauty and wellness experience\\
-        \hline
-    \end{tabular}
-    \caption{\textbf{Tourism experiences’ offer in \emph{Val di Lima} area \tablefootnote{The list might not include all private stakeholders of the territory, since some touristic operators might have an Internet presence and others might operate as public-private players.
+%
+%From the point of view of our research, the context of the \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well touristic services).
+%
+%After a detailed analysis of the touristic and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities  \cite{loz09}.
+%
+%Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of Lucca and the \textit{Garfagnana} mountain area. 
+%
+%%The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like Orrido di Botri, with The Canyon Park \footnote{\url{ https://www.canyonpark.it/}}, and other service companies and associations for experiential tourism.
+%
+%\begin{table}
+%    \centering
+%    \begin{tabular}{|p{70mm}||p{45mm}|} \hline 
+%         {\bf Touristic promotion }&  {\bf Kind of service}\\
+%         \hline \hline 
+%         Aguarajà River Experience \newline \url{https://aguaraja.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
+%         \hline 
+%         Rafting H2O \newline \url{https://www.raftingh2o.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
+%         \hline 
+%         RocKonda \newline \url{https://www.rockonda.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ \hline 
+%         A.S.D. Garfagnana Rafting \newline \url{https://www.garfagnanarafting.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ 
+%        \hline 
+%         Val di Lima Off-Road \newline \url{https://www.valdilimaoffroad.com} & Quad, enduro\\ \hline 
+%         Canyon Park \newline \url{https://www.canyonpark.it/canyon-sup}& Canyoning, aqua-trekking, stand-up padding, mental care and yoga, eco laboratory\\
+%         \hline 
+%         Pro Loco Bagni di Lucca \newline  \url{https://www.turismobagnidilucca.com/} \newline \url{https://www.valdilima.org/}& Sportfishing, mental care, trekking, survival, paragliding, horse riding, diving\\
+%         \hline 
+%         eBike Adventure Tour \newline \url{http://www.ebikeadventuretour.it/}& Ebike, Mountain bike\\ \hline 
+%         Val di Lima eBike \newline  \url{https://valdilimaebike.it/} & Mountain e-bike, Quad, Enduro, Paintball\\
+%         \hline 
+%        Terme Bagni di Lucca \newline \url{https://www.termebagnidilucca.it/} &Spa, beauty and wellness experience\\
+%        \hline
+%    \end{tabular}
+%    \caption{\textbf{Tourism experiences’ offer in \emph{Val di Lima} area \tablefootnote{The list might not include all private stakeholders of the territory, since some touristic operators might have an Internet presence and others might operate as public-private players.
 The companies and associations in the list have been individually contacted but some did not respond to our survey on touristic information.}}}
-    \label{tab:experience}
-\end{table}
-
-The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality.
+%    \label{tab:experience}
+%\end{table}
+%
+%The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality.
 It is 7,22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters of altitude.
-
-\emph{Casoli} is a fascinating village settled on the green and rolling hill landscapes crossed by the river \emph{Lima}, in the \emph{Media Valle del Serchio}.
+%
+%\emph{Casoli} is a fascinating village settled on the green and rolling hill landscapes crossed by the river \emph{Lima}, in the \emph{Media Valle del Serchio}.
 Its territory features multiple kinds of touristic sights: from soft and hard outdoor experiential attractions (specifically the Canyon park activities like canyoning, rafting, and trekking experiences)) to Romanesque churches’ remains and huts.
 Definitely, despite its variety of tourist attractions, \emph{Casoli} exhibits unique features in its comprehensive tourism.
-
-%There aren’t touristic information at the administrative level of \emph{Casoli} , so that the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT.
-
-Given the context and the local vocation, we conclude that \emph{Casoli} is a small destination for tourists practicing heterogeneous touristic circuits in individual and combined ways.
+%
+%%There aren’t touristic information at the administrative level of \emph{Casoli} , so that the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT.
+%
+%Given the context and the local vocation, we conclude that \emph{Casoli} is a small destination for tourists practicing heterogeneous touristic circuits in individual and combined ways.
 According to qualitative and quantitative analysis \cite{loz09} on the touristic and social context of \emph{Bagni di Lucca}, as the municipal area encompassing \emph{Casoli} village, the mentioned area boasts a good touristic appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities.
@@
 
-As explained in the previous section, the area around \textit{Casoli} is a suitable case study location due to its long history, which is reflected in its architectural and geological features.
-In the following, we define the relevant features and report on their accessibility.
-
-The cave named {\em Buca La Piella} is the foremost site from an archeological point of view.
+This section defines the points of interest selected to position the QR tags, and reports on their accessibility and state of conservation.
+This may be relevant in a perspective to spot the segment of users that may take advantage of the tag.
+
+The cave named {\em Buca La Piella} is the foremost site from an archaeological point of view.
 It is located in a fascinating environment outside conventional circuits, which is why it is less frequented and consequently less compromised.
@@
 However, the application records the URL so that the user can visit it when entering a covered area.
-Each Web page contains site-specific information, an interactive map with the location of all the QR tags (see figure \ref{fig:linkmap}), and a form for user feedback (see figure \ref{fig:linkpoll}).
-The map is hosted by the UMap Web service (\url{https://umap.openstreetmap.fr/it/}) \citeWeb{umap}, and displays the location and content of all QR-tags (as shown in figure \ref{fig:qrcodecontent}).
-The server hosting the site is configured to collect aggregated statistics about visitors' activity.
+Each Web page contains site-specific information, an interactive map hosted by the uMap web service (\url{https://umap.openstreetmap.fr/it/}) with the location and content of all the QR tags (see the left box of figure \ref{fig:editing}), and a form for user feedback.
 \end{itemize}
@@
 
-Its {\em performance} is currently difficult to evaluate.
+Collecting statistics about user perception is not a primary interest for us at this stage.
+However, we provided two ways to collect such data.
+One is a 5-questions poll on the web pages linked to each tag.
+The user is asked to agreement a level of agreement with the following sentences:
+\begin {itemize}
+\item the QR tag was useful
+\item a plaque would be preferable
+\item I scanned other QR codes in the vicinity
+\item I will come back to visit the sites I did not visit today
+\item I am satisfied about my visit
+\end{itemize}
+In addition, the hits on the pages linked to the QR-codes were counted.
+Both ways proved to be ineffective.
+In the case of the form, we collected one positive feedback that has no statistical significance.
+The hit count was polluted by spurious access to the page (for instance for maintenance, or reaching the page from the map) and by the fact that scanning the tag was not necessarily linked to a hit.
+%Its {\em performance} is currently difficult to evaluate.
 The recording of the number of times a given tag is scanned, in theory, a significant performance index, is unreliable for technical reasons, one being that the QR tag cannot record or transmit read events.
@@
 The measurement of the efficacy in financial terms is currently beyond the scope of this paper and will be assessable only in the long term depending on the collaboration of the involved stakeholders.
+
+Geo-itinerary
 
@@
 Here too the primary drawback compared with our QR-tag based solution is the need for an ad-hoc application.
-In addition the event of data pollution is not excluded.
+In addition, the event of data pollution is not excluded.
 
@@
 On the technical side, the discussion regarding the deployed experiment indicates directions for future investigation.
-The definition of a potential application will foster research aiming at passive or very low-energy devices that are able to effectively advertise their presence and collect statistics.
+The definition of a potential application will foster research aiming at passive or very low-energy devices that can effectively advertise their presence and collect statistics.
 

### 2023-11-07 paper.tex
Integra modifiche durante la call 6/11, modifiche di Letizia e di Maria Grazia

@@
 Throughout the year, local stakeholders offer a variety of services, often incorporating sustainable tourism practices.
-After a detailed analysis of the tourist and social context of Bagni di Lucca, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{x} dating back to the ‘90s, when some associations began organizing mainly trekking excursions and rafting experiences (\citeWeb{agu,r2o,ronda,garfraft,valdilima,canyonpark,experience,ebikea,ebikel,bdlt2,bdl}).
+After a detailed analysis of the tourist and social context of Bagni di Lucca, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{loz09} dating back to the ‘90s, when some associations \citeWeb{agu,r2o,ronda,garfraft,valdilima,canyonpark,experience,ebikea,ebikel,bdlt2,bdl} began organizing mainly trekking excursions and rafting experiences .
 
@@
 It is important to ensure that any development actions taken are sustainable, in order to preserve its unique social fabric.
-This means paying attention to conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{x}.
+This means paying attention to conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{kuh10a}.
 
@@
 \section{Conclusions}
+
+Based on qualitative and quantitative analysis \cite{loz09} carried out in the social and touristic context of Bagni di Lucca, including the Casoli village, it can be argued that the area under investigation has unique features that need to be capitalized on to enhance its tourism potential.
+Given the touristic trends and the local vocation, we conclude that Casoli is a small destination for potential tourists practicing heterogeneous touristic circuits in individual and combined ways.
+Our study aims to involve stakeholders in sustainable planning for new heritage itineraries, despite accessibility limitations on site.
 

### 2023-11-07 compress.sh
Aggiunta bibliografia a git index!

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+if [ $# -eq 0 ]; then
+ echo "no arguments"
+ exit 1
+fi
+
+bn=`basename $1 .pdf`
+gs -sDEVICE=pdfwrite -dCompatibilityLevel=1.4 -dPDFSETTINGS=/ebook -dNOPAUSE -dQUIET -dBATCH -sOutputFile=$bn-compress.pdf $1 

### 2023-11-07 paper.tex
Revisione introduzione e conclusioni e altro

@@
 \usepackage[caption = false]{subfig}
-\usepackage{tablefootnote}
+%\usepackage{tablefootnote}
 \usepackage{multibib}
@@
 
-Leisure activities play are relevant in bringing profit to niche businesses.
-This creates a reinforcing effect that follows the negative dynamics seen above: a leisure activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact.
+Leisure and cultural activities, like tourism, bring profit to niche businesses.
+This creates a reinforcing effect that follows the negative dynamics seen above: a activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact.
 Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a} relating to outdoor winter sports.
@@
 
-This document addresses the specific aspect of heritage trail signage, and describes a solution based on QR code tags.
-The design gathers suggestions from several disciplines to take into account the many aspects that contribute to sustainability.
-The result is a practical signage technology that is adaptable to other environments sharing defined traits.
+This paper illustrates a concrete exercise in designing for sustainability.
+We introduce an instance of cultural sustainability, an inland region with a legacy to defend, and design a concrete tool to address the issue.
+To reach fully sustainable result, we need to practice from the early design stages an holistic approach to the environmental, social and financial facets of the problem.
+
+The concrete aspect we address is a signage system, and the solution we propose is based on QR code tags. The result is applicable or adaptable to other environments sharing defined traits.
+
+The next section illustrates signage as a topic of interest.
+The discussion targets low cost, low impact technologies that are appropriate for outdoor signage of natural or cultural heritage, and finally indicates QR code tags as a consistent technology.
+The section features a survey of relevant literature.
+
+Section \ref{sec:methods} provides the multidisciplinary framework.
+Section \ref{sec:legacy} illustrates the legacy emphasizing historical facts and locations.
+Section \ref{sec:tourism} describes the social and touristic context of the target location and provides the guidelines for integrating the initiative in the socio-political fabric.
+Such guidelines are used to define an agenda, in section \ref{sec:agenda}, and the section is concluded by a reasoned description of the area, highlighting the points of interest and their reachability \ref{sec:soa}.
+
+Section \ref{sec:implementation} describes the concrete implementation that follows the defined requirements.
+The description of the process to produce the printed tags starting from editable content on a geographic database allows to reproduce and reuse the solution in other contexts.
+
+Section \ref{sec:results} merges the various aspects explored in the previous sections to provide a methodological view of the process, focusing on the interplay between the fourth pillars of sustainability.
+
+The conclusive section summarizes the results, the missing parts, and the potential of the solution.
 
@@
 
-Such a solution is severely limited by the power supply dependency.
+Such a solution is severely limited by the dependency from a power supply.
 A radio transmitter is a rather power-consuming device, in the range of Watts.
@@
 
-The NFC technology is currently very diffused, although reading an NFC tag as text requires the installation of an appropriate application.
-Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the {\em user's} device as a chink of text.
+The NFC technology is available without installing ad-hoc applications.
+Its operation consists in approaching the smartphone to the tag, and the user is advertised that a content is available.
 The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues.
-Such capabilities do not need an Internet connection.
+Such capabilities do not need an Internet connection but currently depend on ad-hoc applications.
 
 Another solution in the family of passive devices is the QR tag (QR stands for Quick-Response).
-Such a technology does not require radio communication but uses the smartphone camera to acquire a graphical code that is translated into text.
+Such a technology does not require radio communication but uses the smartphone camera to acquire text encoded in a graphical code.
 The capacity of a QR tag depends on the number of dots in the image, which in turn depends on the size of the code and the smartphone camera characteristics.
@@
 
-A preliminary check verifies compliance with the Universal Design Principles \cite{udi97a}:
+A preliminary check verifies compliance of the passive technologies described above with the Universal Design Principles \cite{udi97a}:
 
@@
 
-Such minimal solutions (see figure \ref{fig:qrnfccoin}) compare well with other more complex ones that make use of the user's smartphone.
-There is a trade-off concerning capacity, but in many circumstances, a capacity of 2-300 words is sufficient to convey the description of the site or provide directions for the visit.
+Compared with other smartphone-based technologies, there is a trade-off concerning capacity.
+However, a signage applications a capacity of 2-300 words is sufficient to provide directions for the visit and convey text-based educational contents.
 If capacity limits are not an issue, a solution based on NFC or QR tags exhibits several advantages:
@@
 \begin{itemize} 
-\item does not require any power supply,
+\item does not require power supply,
 \item has a limited impact on the landscape,
@@
 \item produces a limited quantity of waste when disposed of,
-\item content can be stored for later usage; for instance, to visit a URL once the user reaches a zone covered by the Internet.
+\item content can be stored for later usage; for instance, to visit a URL once the user reaches a zone covered by the Internet,
+\item is inclusive regarding user inabilities and nationality,
+\item requires basic technical skills from the user.
 \end{itemize}
@@
 
-For a signage application, a QR tag is preferable because a noticeable size is needed. In addition, keeping the tag out of reach prevents vandalism and misuse.
-
-In conclusion, we have reasons to select a QR-tag-based solution as a good candidate for smartphone-based signage.
-We now consider how it copes with the limits of a traditional board-based approach (as listed in table \ref{sec:signage}):
+For a signage application, a QR tag is preferable because a noticeable size is needed.
+In addition, keeping the tag out of reach prevents vandalism and misuse.
+Therefore we have reasons to select a QR-tag-based solution for smartphone-based signage.
+
+Its use is especially appropriate when the cultural heritage extends across a scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging.
+In these cases, QR tags could improve the visitors' experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos.
+Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot.
+The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.
+
+We now examine how such technology copes with the limits of a traditional board-based approach (as listed in table \ref{sec:signage}):
 
@@
 
-\section{Materials and Methods}
+\section{Materials and Methods \label{sec:methods}}
 
@@
 
-\subsection{The natural and cultural resources of \emph{Casoli} and its vicinity}
+\subsection{The natural and cultural assets of \emph{Casoli} and its vicinity \label{sec:legacy}}
 
@@
 
-\subsection{An integrated view of tourism in the \emph{Casoli} region}
+\subsection{An integrated view of tourism in the \emph{Casoli} region \label{sec:tourism}}
 
@@
 
-The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like the \emph{Orrido di Botri} with the \emph{Canyon Park\footnote{\url{ https://www.canyonpark.it/}}}, and other service companies and associations for experiential tourism.
+The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like the \emph{Orrido di Botri} \citeWeb{botri} with the \emph{Canyon Park} \citeWeb{canyonpark}, and other service companies and associations for experiential tourism.
 
 In reason of a whole insight of the mentioned touristic area, a socio-cultural approach is needed, by considering also tourism flows.
-In this perspective, it must be considered that there isn’t touristic information at the administrative level of \emph{Casoli}, so the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT (National Statistical Institute).
+In this perspective, it must be considered that there isn’t touristic information at the administrative level of \emph{Casoli}, so the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT (National Statistical Institute) \citeWeb{istat}.
  The demographic trend (Figure 4, a) enlightens a relevant aspect of \emph{Bagni di Lucca} society and economic aspects of its touristic ecosystem (Figure 4, b).
@@
 
-\emph{Bagni di Lucca} stretches on 164.70 $km^2$ area with a population density \footnote{Index referring to the number of people living in a territory per square kilometer.
-It is the rate between the annual number of residents and the surface area.
-In our case and for 2022, the density is 5593 people over 164.70 $km^2$.} of 3395 inhabitants per square kilometer and a job occupancy rate of 33.53\%.
+\emph{Bagni di Lucca} stretches on 164.70 $km^2$ area with a population density of 33.95 inhabitants per square kilometer and a job occupancy rate of 33.53\%.
 Regarding human growth at the territorial level, it is relevant to consider the \emph{Bagni di Lucca} population trend during the 2001-2022 period.
@@
 
-The Tourism Density Index \footnote{The Tourism Density Index shows the tourism concentration in the higher touristic season.
-It is the rate between the number of tourists and the surface area.} is 106.41 tourists per $km^2$ in 2022.
+The Tourism Density Index is the rate between the number of tourists and the surface area in the higher touristic season.
+It is 106.41 tourists per $km^2$ in 2022.
 The arrivals exhibit a drop from 2019 (from  4339 in 2017 to 2363 in 2021), and touristic presences have suffered a slight decrease during the same period, as shown in figure \ref{fig:tourismflows}, while a further increase has been registered in 2022 (5.876 arrivals).
@@
         Accommodation density (accommodations/{$km^2$}) & 10.14 & 19.30 \\
-        Coexistence Index (100 $\times$ Foreign/Italian) & 103.74\% &  69.87\% \\ 
+        Coexistence (100 $\times$ Foreign/Italian) & 103.74\% &  69.87\% \\ 
          \hline
     \end{tabular}
-    \caption{Tourist indexes in \emph{Bagni di Lucca} in 2022\tablefootnote{Each index gives a different type of touristic information about \emph{Bagni di Lucca} Municipality: the touristic density represents the number of tourists per $km^2$; the Average stay index is the average number of nights spent in town by tourists; the Accommodation Density Index is related to the number of touristic presences on the number of beds occupied by tourists all over the year; the coexistence index shows the relevance of foreign tourists per 100 Italian tourists.}}
+    \caption{Tourist indexes in \emph{Bagni di Lucca}
+%.
+The {\em touristic density} is the number of tourists in a year per $km^2$.
+The {\em average stay} is the average number of nights spent in town by tourists.
+The {\em accommodation density} is the rate between the number of touristic presences and the number of beds occupied by tourists in a year.
+The {\em Coexistence} is the percent rate between foreign and Italian tourists.
+}
     \label{tab:tourism}
@@
 
-\subsubsection{Our activity}
+\subsubsection{Our activity \label{sec:agenda}}
 
@@
 
-\subsubsection{Sites and trail’s state of art}
+\subsubsection{Sites and trail’s state of art \ref{sec:soa}}
 
@@
 
-\subsection{Implementing a QR-based signage}
+\subsection{Implementing a QR-based signage \label{sec:implementation}}
 
@@
 
-We aim to define a design approach that will enhance the tourist experience of an inland area.
-The approach is conceptual and therefore applicable to a wide range of use cases.
-First, the requirements for sustainability in design are defined, taking into account the three pillars of sustainability: social, environmental, and economic.
-
-To put our approach into practice we used a small-scale use case: a signage system addressing the guidance of visitors in an inland area.
-The area has a cultural and environmental heritage to be protected and valorized at the same time.
+We aim to enhance the tourist experience in an inland area located around the village of Casoli.
+Our goal is tightly related with the fourth pillar of sustainability: there are natural and cultural assets that need to be nurtured or at least preserved.
+We concentrate on a signage system, the realization of which is finalized to reaching such cultural result.
+
+To avoid the over-tourism pitfall, we need to take into account the other three pillars: social, environmental, and economic.
+To this end we analyze the social fabric that will host our initiative, the environment where signage will be deployed, and the economic framework that will support its implementation.
+From this study we derive the requirements for our signage system: low environmental impact, open to participation of the residents, easy to use from the visitors, and economically sustainable with a minimal touristic flow.
+
+The approach we applied to reach such conclusions is summarized below: it is conceptual and therefore applicable to a wide range of use cases, and merely consists in designing the system with sustainability in mind from the early stages.
 
@@
 The environmental aspect is characterized by natural formations and human artifacts tracing back to the Paleolithic Era.
-As for tourism-related economics, currently the dominant component in the area, we noticed that the location selected for our use case exhibits unexpressed potentialities although the area surrounding the location has a recognized touristic vocation.
-From the social point of view, we observed that the area is exposed to depopulation due to the shortage of productive activities.
-
-One conclusion after the above analysis is that, in order to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable.
-Regarding the potential reasons for a visit, we found a not easily accessible cave inhabited since pre-history, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
-
-We investigated a technique for guiding the visitor and telling the history of the place. The fragile economy limits financial investment, while the natural heritage requires non-intrusive techniques.
+As for tourism-related economics, a qualitative and quantitative analysis \cite{loz09} reveals unique features that need to be capitalized on to enhance the tourism potential.
+Given the touristic trends and the local vocation, we conclude that Casoli is a small destination for visitors practicing heterogeneous touristic circuits in individual and combined ways.
+Cultural sustainability mandates the involvement of local stakeholders in planning new heritage itineraries and supporting initiatives, including signage.
+
+An outcome of the analysis of the social fabric is that, in order to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable.
+Regarding the potential reasons for a visit, we found a cave inhabited since pre-history but reachable with some effort, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
+Such targets offer a variety of access ways, from those requiring performance to the relaxing ones.
+For this reason we opted for not suggesting a geo-itineray, allowing the user to follow personal inclinations.
+In this way we expect, justified by the results of \cite{sol18a}, that the educational impact of the signage is maximized.
+
+We investigated a technique for guiding the visitor and telling the history of the place.
+The small number of expected visitors limits financial investment, while the natural heritage requires non-intrusive techniques.
 
@@
 
-The preparation of the signage required a solid knowledge of the area that was gained by carrying out archaeological surveys and consulting historical documents and cartography.
-The evaluation of touristic vocation was based on the available statistical data and person-to-person informal interviews.
+Previous investigations, including archaeological surveys and access to historical documents and cartography, allowed to design the contents to be delivered through signage.
 Two on-site surveys provided further practical indications and the opportunity to place the QR tags.
 
-On the technical side, the proposed signage is a benchmark difficult to improve in terms of environmental impact and cost.
+The main result from our activity is the real scale deployment of the signage, a non trivial exercise in holistic sustainable design.
+
+As a side result we have a signage technology based on QR code tags, a benchmark difficult to improve in terms of environmental impact and cost.
+
+%Its use is especially appropriate when the cultural heritage extends across a scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging.
+In these cases, QR tags could improve the visitors' experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos.
+Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot.
+The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.
 
@@
 
-In areas like our use case, where the cultural heritage extends across scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging.
-In these cases, QR tags could improve the visitors' experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos.
-Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot.
-The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.
-
-Collecting statistics about user perception is not a primary interest for us at this stage.
-However, we provided two ways to collect such data.
-One is a 5-questions poll on the web pages linked to each tag.
-The user is asked to agreement a level of agreement with the following sentences:
-\begin {itemize}
+Collecting statistics about user perception is not a primary interest for us at this stage: the initiative started as an exercise and was not adequately advertised to the potential users.
+However, we provided two ways to collect data as part of the exercise.
+
+One is a poll in five points on the web pages linked to each tag. The user is asked to quantify an agreement (0-do not agree, 5-strongly agree) with four sentences, and finally provide an answer:
+\begin{itemize}
 \item the QR tag was useful
 \item a plaque would be preferable
-\item I scanned other QR codes in the vicinity
 \item I will come back to visit the sites I did not visit today
 \item I am satisfied about my visit
+\item I scanned other QR codes in the vicinity
 \end{itemize}
-In addition, the hits on the pages linked to the QR-codes were counted.
-Both ways proved to be ineffective.
-In the case of the form, we collected one positive feedback that has no statistical significance.
-The hit count was polluted by spurious access to the page (for instance for maintenance, or reaching the page from the map) and by the fact that scanning the tag was not necessarily linked to a hit.
+
+The other is an hits count on the pages linked to the QR-codes.
+
+Both of them proved to be ineffective.
+We collected only one form with a positive feedback but with no statistical significance.
+The hit count was polluted by spurious accesses to the page (for instance for maintenance, or reaching the page from the map) and by the fact that scanning the tag was not necessarily linked to a hit.
 %Its {\em performance} is currently difficult to evaluate.
@@
 
-Geo-itinerary
-
 Although we are satisfied that the solution complies with the sustainability principles that inspired this work, we identified two issues that indicate directions for future research:
@@
 \section{Conclusions}
-
-Based on qualitative and quantitative analysis \cite{loz09} carried out in the social and touristic context of Bagni di Lucca, including the Casoli village, it can be argued that the area under investigation has unique features that need to be capitalized on to enhance its tourism potential.
-Given the touristic trends and the local vocation, we conclude that Casoli is a small destination for potential tourists practicing heterogeneous touristic circuits in individual and combined ways.
-Our study aims to involve stakeholders in sustainable planning for new heritage itineraries, despite accessibility limitations on site.
 

### 2023-11-07 diff.sh

@@ -1,0 +1,8 @@
+rev="e8af57"
+
+latexdiff-vc -r $rev paper.tex
+pdflatex paper-diff$rev.tex
+bibtex paper-diff$rev
+pdflatex paper-diff$rev.tex
+pdflatex paper-diff$rev.tex
+./compress.sh paper-diff$rev.pdf

### 2023-11-08 paper.tex
Post call 8/11

@@
 $^{2}$ \quad University of Pisa; mariagrazia.deri@sp.unipi.it\\
-$^{3}$ \quad University of Pisa; l.chiti2@studenti.unipi.it}
+$^{3}$ \quad University of Pisa; letizia.chiti@}
 
@@
 Section \ref{sec:tourism} describes the social and touristic context of the target location and provides the guidelines for integrating the initiative in the socio-political fabric.
-Such guidelines are used to define an agenda, in section \ref{sec:agenda}, and the section is concluded by a reasoned description of the area, highlighting the points of interest and their reachability \ref{sec:soa}.
+Such guidelines are used to motivate and define an agenda, in section \ref{sec:agenda}, and the section is concluded by a reasoned description of the area, highlighting the points of interest and their reachability \ref{sec:soa}.
 
@@
 This approach can involve the community, especially in the southern part of the municipality, which is still less frequented, more hilly, and therefore less traveled by the network of trails.
-We envision the creation of geo-itineraries characterized by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.
+We envision the creation of heritage trails characterized by the rediscovery of the historical roads, partly well-preserved, which connected the villages with the valley bottom and between them, including those sites of interest that encapsulate the history of this area, starting with the caves.
 
@@
     \end{tabular}
-    \caption{Tourist indexes in \emph{Bagni di Lucca}
+    \caption{Tourist indexes in \emph{Bagni di Lucca} in 2022
 %.
@@
 
-\subsubsection{Our activity \label{sec:agenda}}
+\subsubsection{How top implement the guidelines? \label{sec:agenda}}
 
@@
 
-\subsubsection{Sites and trail’s state of art \ref{sec:soa}}
+\subsubsection{Sites and trail’s state of the art \ref{sec:soa}}
 
@@
 However this requires the installation of the app on the visitor, and the design of the IT product including the device and the application, which may cost far more than the physical devices.
-
-A different application-based solution privileging passive devices is based on a writable NFC tag close to the one delivering the content.
-To update the number of visitors that have read the informational NFC, the user application reads and increments the count recorded on the writable NFC.
-Here too the primary drawback compared with our QR-tag based solution is the need for an ad-hoc application.
-In addition, the event of data pollution is not excluded.
 

### 2023-11-08 paper.tex
Post update 8/11 (serale)

@@
 Its use is especially appropriate when the cultural heritage extends across a scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging.
-In these cases, QR tags could improve the visitors' experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos.
+In these cases, QR tags could improve the visitors’ experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos, even if they don’t have access to a specific spot.
 Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot.
@@
 For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study with the purpose of highlighting the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}.
-As concerns tourist indexes shown in Chart 3, the touristic density counts 1866.96 people per square kilometer (1773 $km^2$), with an average stay of 3.36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2.98 nights).
+As concerns tourist indexes shown in Table \ref{tab:tourism}, the touristic density counts 1866.96 people per square kilometer (1773 $km^2$), with an average stay of 3.36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2.98 nights).
 The provincial area of \emph{Lucca} shows an accommodation density of 19.30 on 10.14 of \emph{Bagni di Lucca}, while the coexistence index of \emph{Lucca} (with 69.87 foreign tourists on 100 Italian tourists) is lower than \emph{Bagni di Lucca}’s coexistence index (with 103.74 foreign tourists in 100 Italian tourists).
@@
 
-Several promotional websites advertise Bagni di Lucca attractions such as these two destination websites: www.turismobagnidilucca.com and www.valdilima.
-org; the sitography closing this paper contains a long yet incomplete list that confirms such a statement.
-
-From the point of view of our research, the context of Bagni di Lucca helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well as touristic services).
-Indeed, Val di Lima can promote a very identifying destination brand straddling the town of Lucca and the Garfagnana mountain area.
-The Municipality of Bagni di Lucca and the tourist area of the Val di Lima are the context for the small and fascinating village of Casoli, which is one of the 31 villages belonging to the Municipality.
-It is 7.22 km far from Bagni di Lucca, at an altitude of 500 meters.
-
-Casoli village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as: canyoning, rafting, and trekking experiences in the Canyon park.
+Several promotional websites, like \citeWeb{tbdl} and \citeWeb{valdilima} advertise \emph{Bagni di Lucca} attractions.
+The sitography closing this paper contains a longer yet incomplete list of touristic initiatives  \citeWeb{agu,r2o,ronda,garfraft,valdilima,canyonpark,experience,ebikea,ebikel,bdlt2,bdl}.
+
+From the point of view of our research, the context of \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well as touristic services).
+Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of \emph{Lucca} and the \emph{Garfagnana} mountain area.
+The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small and fascinating village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality.
+It is 7.22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters.
+
+\emph{Casoli} village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as: canyoning, rafting, and trekking experiences in the Canyon Park.
 Additionally, it features Romanesque churches and huts attracting visitors looking for various experiences such as thermal, environmental, outdoor, sport and cultural activities.
 Throughout the year, local stakeholders offer a variety of services, often incorporating sustainable tourism practices.
-After a detailed analysis of the tourist and social context of Bagni di Lucca, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{loz09} dating back to the ‘90s, when some associations \citeWeb{agu,r2o,ronda,garfraft,valdilima,canyonpark,experience,ebikea,ebikel,bdlt2,bdl} began organizing mainly trekking excursions and rafting experiences .
-
-As part of Bagni di Lucca's ecosystem, with its focus on tourism, Casoli is a destination for slow travel.
+After a detailed analysis of the tourist and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{loz09} dating back to the ‘90s, when some associations began organizing mainly trekking excursions and rafting experience .
+
+As part of \emph{Bagni di Lucca}'s ecosystem, with its focus on tourism, \emph{Casoli} is a destination for slow travel.
 It is important to ensure that any development actions taken are sustainable, in order to preserve its unique social fabric.
@@
 
-%Summing up, the quantitative data give evidence of a vibrant and attractive tourist reality constantly attracting tourists during the last decade.
-Thanks to such tourist flows \emph{Bagni di Lucca} is considered one of the main tourist areas of the \emph{Val di Lima} and the Serchio Valley.
-Several promotional websites advertise \emph{Bagni di Lucca} attractions\footnote{Among local websites, these two destination websites are: \url{www.turismobagnidilucca.com} and \url{www.valdilima.org}}: the sitography closing this paper contains a long yet incomplete list that confirms such a statement, and Table \ref{tab:experience} exposes the promotional touristic system offering varied experiences in the \emph{Val di Lima} area.
-The variety of outdoor touristic services dates back to the ‘90s when some associations began organizing mainly trekking excursions and rafting experiences.
-%
-%From the point of view of our research, the context of the \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well touristic services).
-%
-%After a detailed analysis of the touristic and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities  \cite{loz09}.
-%
-%Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of Lucca and the \textit{Garfagnana} mountain area. 
-%
-%%The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like Orrido di Botri, with The Canyon Park \footnote{\url{ https://www.canyonpark.it/}}, and other service companies and associations for experiential tourism.
-%
-%\begin{table}
-%    \centering
-%    \begin{tabular}{|p{70mm}||p{45mm}|} \hline 
-%         {\bf Touristic promotion }&  {\bf Kind of service}\\
-%         \hline \hline 
-%         Aguarajà River Experience \newline \url{https://aguaraja.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
-%         \hline 
-%         Rafting H2O \newline \url{https://www.raftingh2o.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
-%         \hline 
-%         RocKonda \newline \url{https://www.rockonda.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ \hline 
-%         A.S.D. Garfagnana Rafting \newline \url{https://www.garfagnanarafting.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ 
-%        \hline 
-%         Val di Lima Off-Road \newline \url{https://www.valdilimaoffroad.com} & Quad, enduro\\ \hline 
-%         Canyon Park \newline \url{https://www.canyonpark.it/canyon-sup}& Canyoning, aqua-trekking, stand-up padding, mental care and yoga, eco laboratory\\
-%         \hline 
-%         Pro Loco Bagni di Lucca \newline  \url{https://www.turismobagnidilucca.com/} \newline \url{https://www.valdilima.org/}& Sportfishing, mental care, trekking, survival, paragliding, horse riding, diving\\
-%         \hline 
-%         eBike Adventure Tour \newline \url{http://www.ebikeadventuretour.it/}& Ebike, Mountain bike\\ \hline 
-%         Val di Lima eBike \newline  \url{https://valdilimaebike.it/} & Mountain e-bike, Quad, Enduro, Paintball\\
-%         \hline 
-%        Terme Bagni di Lucca \newline \url{https://www.termebagnidilucca.it/} &Spa, beauty and wellness experience\\
-%        \hline
-%    \end{tabular}
-%    \caption{\textbf{Tourism experiences’ offer in \emph{Val di Lima} area \tablefootnote{The list might not include all private stakeholders of the territory, since some touristic operators might have an Internet presence and others might operate as public-private players.
-The companies and associations in the list have been individually contacted but some did not respond to our survey on touristic information.}}}
-%    \label{tab:experience}
-%\end{table}
-%
-%The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality.
-It is 7,22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters of altitude.
-%
-%\emph{Casoli} is a fascinating village settled on the green and rolling hill landscapes crossed by the river \emph{Lima}, in the \emph{Media Valle del Serchio}.
-Its territory features multiple kinds of touristic sights: from soft and hard outdoor experiential attractions (specifically the Canyon park activities like canyoning, rafting, and trekking experiences)) to Romanesque churches’ remains and huts.
-Definitely, despite its variety of tourist attractions, \emph{Casoli} exhibits unique features in its comprehensive tourism.
-%
-%%There aren’t touristic information at the administrative level of \emph{Casoli} , so that the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT.
-%
-%Given the context and the local vocation, we conclude that \emph{Casoli} is a small destination for tourists practicing heterogeneous touristic circuits in individual and combined ways.
-According to qualitative and quantitative analysis \cite{loz09} on the touristic and social context of \emph{Bagni di Lucca}, as the municipal area encompassing \emph{Casoli} village, the mentioned area boasts a good touristic appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities.
-Local stakeholders operating in the tourism system offer a variety of services all year round.
-Whenever possible the offer is enriched with original touristic approaches paying attention to sustainability.
-As part of the \emph{Bagni di Lucca}’s ecosystem with its touristic vocation, \emph{Casoli} is defined as a well-known destination.
-Considering that \emph{Casoli} is a destination for slow tourism, ensuring the sustainability of development actions is crucial to maintaining its unique social fabric.
-The latter aspect suggests paying attention to environmental conservation, cultural heritage conservation, intergenerational continuity, and economic equity \cite{kuh10}.
-
 \subsubsection{Guidelines for a sustainable development}
 
-As described by \cite{dre22} Dianne Dredge and in light of the above analysis, private and public entities can contribute to sustainable tourism  by developing the following actions:
-
-%In this perspective, policy-makers play a key role in the planning of sustainable measures and in the monitoring of impacts; in particular, in the context of tourism, sustainability is a topic discussed in its many aspects, including resident ecosystem resilience approaches and regeneration actions by policy-makers and/or private tour operators \cite{dre22}.
+Based on the tourism trend of \emph{Casoli} and its overall tourism, and in alignment with the research goals focused on sustainable tourism, it is crucial to consider the role of local policy-makers.
+In fact, they play a vital role in planning sustainable measures and monitoring their impact.
+
+In these terms, as described by [50] Dianne Dredge, and in light of the above analysis, private and public entities can contribute to sustainable tourism by developing the following actions:
 
@@
 
-Regarding residents, we had the chance to meet a few people living in \emph{Casoli}, which did not provide useful information; local entrepreneurs we met informed us about the trekking tourism flows passing through \emph{Casoli}  and the Piella Cave path, which usually starts from the Canyon Park experience.
+Regarding residents, we had the chance to meet a few people living in \emph{Casoli}, which did not provide useful information; local entrepreneurs we met informed us about the trekking tourism flows passing through \emph{Casoli} and the \emph{Piella} cave path, which usually starts from the Canyon Park experience.
 
@@
 
-\subsubsection{Sites and trail’s state of the art \ref{sec:soa}}
+\subsubsection{Sites and trail’s state of the art \label{sec:soa}}
 
@@
 An ad-hoc application referenced in the paper allows to transform the uMap file downloaded from the server into a batch of printable QR tags (top left).
-The tag is highlighted in red in the picture on the Web display}
+The tag is highlighted in red in the picture on the Web display \label{fig:editing}}
 \end{figure}
@@
 
-Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the UMap user interface already fit the purpose (see figure \ref{fig:umap-editing}).
+Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the UMap user interface already fit the purpose (see the bottom left box in figure \ref{fig:editing}).
 The resulting \emph{GeoJSON} object describing a tagged feature is shown in figure \ref{fig:geojson-sample}.
 
-The master document is rendered by UMap on the page linked to the tags (as in figure \ref{fig:linkmap}).
+The master document is rendered by UMap on the page linked to the tags (as shown in figure \ref{fig:editing}).
 
@@
 To this end we analyze the social fabric that will host our initiative, the environment where signage will be deployed, and the economic framework that will support its implementation.
-From this study we derive the requirements for our signage system: low environmental impact, open to participation of the residents, easy to use from the visitors, and economically sustainable with a minimal touristic flow.
+From this study we derive the requirements for our signage system: low environmental impact, open to participation of the residents, responsive to the visitor's information needs, and economically sustainable with a minimal touristic flow.
 
 The approach we applied to reach such conclusions is summarized below: it is conceptual and therefore applicable to a wide range of use cases, and merely consists in designing the system with sustainability in mind from the early stages.
+The simplicity of the approach description must not mislead: its implementation requires collaboration among experts in distinct disciplines converging to the final result.
 
 Our design started by defining a profile for the area according to the \emph{three pillars}.
-The environmental aspect is characterized by natural formations and human artifacts tracing back to the Paleolithic Era.
+The environmental aspect is characterised by natural formations and special habitats hosting a vast biodiversity, placed under special protection according to European legislation.
+
 As for tourism-related economics, a qualitative and quantitative analysis \cite{loz09} reveals unique features that need to be capitalized on to enhance the tourism potential.
-Given the touristic trends and the local vocation, we conclude that Casoli is a small destination for visitors practicing heterogeneous touristic circuits in individual and combined ways.
-Cultural sustainability mandates the involvement of local stakeholders in planning new heritage itineraries and supporting initiatives, including signage.
+In this light, community-involved tourism
+represents a virtuous approach for destination management in inland areas that need a renewed sustainability-oriented design.
+Therefore, public and private stakeholders are crucial for managing spatial planning and addressing local issues through investments, conservation, promotion and monitoring actions.
+
+In this perspective, our research is oriented to stimulate local policymakers and tourism professionals towards a better environmental and cultural sustainability approach by taking advantage of our study results to realize new potential heritage trails in Casoli and its vicinity.
+
+%As for tourism-related economics, a qualitative and quantitative analysis reveals unique features that need to be capitalized on to enhance the tourism potential.
 
@@
 Such targets offer a variety of access ways, from those requiring performance to the relaxing ones.
-For this reason we opted for not suggesting a geo-itineray, allowing the user to follow personal inclinations.
-In this way we expect, justified by the results of \cite{sol18a}, that the educational impact of the signage is maximized.
+For this reason we opted for not suggesting a geo-itineray, allowing the user cherry-picking locations based on personal interests.
+In this way we expect, legitimated by the results of \cite{sol18a}, that the educational impact of the signage is maximized.
+Furthermore, our choice aims at a conciseness and communicative effectiveness of the tourist information, by inserting within the QR code a text of an average of 460 characters that facilitates readability and immediately provides the information without further steps to web pages, which may slow down the visit or discourage the less motivated visitor.
 
@@
 We show how a balanced use of technology may help in the task, with a process that is necessarily multidisciplinary, joining skills coming from the diverse domains: sociology, economics, humanities, and engineering.
-The paper demonstrates how their interplay returns ad-hoc tools targeting sustainability and efficacy.
+Taking into account all such aspects is a challenging task, even when applied to a limited use case.
+Reporting about the design process is also a complex task, as witnessed by this paper.
+However, the result is rewarding and we demonstrates how the multidisciplinary interplay returns ad-hoc tools targeting sustainability and efficacy.
+
+In the case of \emph{Casoli}, the touristic trends and the local vocation portraits a small destination for visitors practicing heterogeneous touristic circuits in individual and combined ways.
+Cultural sustainability mandates the involvement of local stakeholders in planning new heritage itineraries and supporting initiatives, including signage.
 
@@
 
-\nociteWeb{*}
+%\nociteWeb{*}
 \bibliographystyleWeb{abbrv}
@@
 % If authors have biography, please use the format below
-%\section*{Short Biography of Authors}
-%\bio
-%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author1.pdf}}}
-%{\textbf{Firstname Lastname} Biography of first author}
-%
-%\bio
-%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author2.jpg}}}
-%{\textbf{Firstname Lastname} Biography of second author}
-
+\section*{Short Biography of Authors}
+\bio
+{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/mariagrazia.png}}}
+{\textbf{Maria Grazia Deri} PhD in Migration and Tourism of the roots at the Department of Political Science of the University of Pisa.
+Researcher on the Geography of Tourism and Human Geography, as well as Destination and Destination Marketing, with interests in cultural and sustainable tourism.
+Researcher in the framework of European projects about tourism and member of the European Romea Strata Association (AERS).}
+
+\bio
+{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/letizia.png}}}
+{\textbf{Letizia CHiti} Research fellow at the University of Pisa, Department of Civilization and Forms of Knowledge.
+She graduated in Archaeology at the University of Pisa and obtained a specialization diploma at SISBA with an essay on the archaeological map of the municipality of Bagni di Lucca (LU).
+He participated in projects and surface reconnaissance in mountain areas in Liguria and Tuscany, specializing in non-invasive analysis methods (GIS and study of elevations) with a diachronic approach.}
+
+\bio
+{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/augusto.png}}}
+{\textbf{Augusto Ciuffoletti} Former teacher and researcher at the University of Pisa, retired on September 2023, with recent achievements on the Internet of Things and distributed control, with an interest on low-power devices.
+Previously active on Grid and, later, Cloud computing in collaboration with INFN.
+Until 2023, teaching activity at the Degree Course in Computer Humanities of the University of Pisa with two courses respectively on Network Protocols and the Angular 2+ framework.}
 % For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references

### 2023-11-09 paper.tex
In seconda revisione

@@
 % Full title of the paper (Capitalized)
-\Title{A sustainable approach to tourist signage on heritage trails\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}}
+\Title{A sustainable approach to tourist signage on heritage trails
+%\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}
+}
 % MDPI internal command: Title for citation in the left column
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{Guiding the visitor to appreciate the sights during an excursion is crucial to building a memorable experience.
-In the literature, we find many articles exploring the use of sophisticated technologies towards such a goal.
-The case of tourism proposals for inland areas exhibits a noteworthy aspect, since the presence of human artifacts and the signage may damage the experience, and introduce pollutants as well.
-This paper analyzes the signage for a natural trail putting into practice a holistic methodology that takes into account all aspects of the problem, starting from the touristic vocation of the area and the social fabric.
-The analysis indicates slow, community-involved tourism as the preferable target, with extreme requirements of sustainability in the social, economic, and environmental domains for the signage solution.
-Appropriate governance guidelines are given to implement the proposal.
-Footprint evaluations address the QR-code technology, which is used to design a real-scale solution finally deployed on the field.}
+\abstract{Understanding the cultural aspects of an area rich in heritage is crucial to building a lasting educational experience from an excursion.
+Many articles in literature explore the use of sophisticated technologies to achieve such a goal.
+Tourism proposals for inland areas are significant because the presence of human artifacts and signage can harm the experience and create pollutants.
+Through a holistic methodology and analysis, this paper examines the signage for an area rich in cultural and natural assets: the study encompasses history, touristic vocation, and the environmental context.
+According to the analysis, slow, community-involved tourism is the preferred destination, and the signage solution must meet strict sustainability requirements in the social, economic, and environmental realms.
+After applying the appropriate governance guidelines provided, the QR code technology is selected for a thoroughly documented experimental deployment.}
 %\abstract{A single paragraph of about 200 words maximum.
@@
 Although an official decision is still pending, there is a widely agreed proposal for the naming of the time that we are living as the Anthropocene.
-Behind this proposal is the recognition that humans have the potential to significantly alter on a global scale the environment in which they live.
+Behind this proposal is the recognition that humans have the potential to significantly alter, on a global scale, the environment in which they live.
 
@@
 Leisure and cultural activities, like tourism, bring profit to niche businesses.
-This creates a reinforcing effect that follows the negative dynamics seen above: a activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact.
+This creates a reinforcing effect that follows the negative dynamics seen above: an activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact.
 Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a} relating to outdoor winter sports.
@@
 
-This research stems from such a tough contradiction in trying to answer the question: is there a way to unlock the potential of cultural heritage while at the same time limiting the damage to a social, economic, and environmental equilibrium?
+This research stems from such a tough contradiction in trying to answer the question: which is the way to unlock the potential of cultural heritage while at the same time limiting the damage to a social, economic, and environmental equilibrium?
 
@@
 We introduce an instance of cultural sustainability, an inland region with a legacy to defend, and design a concrete tool to address the issue.
-To reach fully sustainable result, we need to practice from the early design stages an holistic approach to the environmental, social and financial facets of the problem.
+To reach fully sustainable results, we need to practice from the early design stages a holistic approach to the environmental, social, and financial facets of the problem.
 
@@
 The next section illustrates signage as a topic of interest.
-The discussion targets low cost, low impact technologies that are appropriate for outdoor signage of natural or cultural heritage, and finally indicates QR code tags as a consistent technology.
+The discussion targets low-cost, low-impact technologies that are appropriate for outdoor signage of natural or cultural heritage, and finally indicates QR code tags as a consistent technology.
 The section features a survey of relevant literature.
@@
 Section \ref{sec:legacy} illustrates the legacy emphasizing historical facts and locations.
-Section \ref{sec:tourism} describes the social and touristic context of the target location and provides the guidelines for integrating the initiative in the socio-political fabric.
+Section \ref{sec:tourism} describes the social and touristic context of the target location and provides the guidelines for integrating the initiative into the socio-political fabric.
 Such guidelines are used to motivate and define an agenda, in section \ref{sec:agenda}, and the section is concluded by a reasoned description of the area, highlighting the points of interest and their reachability \ref{sec:soa}.
-
 Section \ref{sec:implementation} describes the concrete implementation that follows the defined requirements.
-The description of the process to produce the printed tags starting from editable content on a geographic database allows to reproduce and reuse the solution in other contexts.
-
-Section \ref{sec:results} merges the various aspects explored in the previous sections to provide a methodological view of the process, focusing on the interplay between the fourth pillars of sustainability.
+The description of the process to produce the printed tags starting from editable content on a geographic database allows reproduction and reuse of the solution in other contexts.
+
+Section \ref{sec:results} merges the various aspects explored in the previous sections to provide a methodological view of the process, focusing on the interplay between the four pillars of sustainability.
 
@@
  \item a sign should explain the reasons for interest in a site.
-This generic characterization includes all sorts of semantic characterization of the site, and may have educational or motivational intents.
-This includes describing historical facts related with the site, as well as advertising nearby points of interest (see figure \ref{fig:pole-direction}).
+This generic characterization includes all sorts of semantic characterization of the site and may have educational or motivational intents.
+This includes describing historical facts related to the site, as well as advertising nearby points of interest (see figure \ref{fig:pole-direction}).
 In this sense, the CAI signal above is not sufficient.
@@
 In our study, we consider a range of solutions that support sustainable tourism, specifically the dynamic provision of information during a visit along a natural trail not covered by Internet connectivity.
-We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.
+We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts information from signing devices.
 
@@
 
-One is that smartphones although widely available, are not equally familiar to everybody.
+One is that smartphones, although widely available, are not equally familiar to everybody.
 This is related to {\em usability}, with a term borrowed from P.
@@
 The second is that several functions depend on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach the Internet.
-So the {\em applicability limits} of a solution need to be defined.
-
-Starting from the two aspects above we envision the guidelines for a successful smartphone-based strategy, keeping in mind its sustainability.
+So the {\em applicability limits} of a smartphone-based solution need to be defined.
+
+Starting from the two aspects above we envision the guidelines for a successful strategy, keeping in mind its sustainability.
 
@@
 
-A popular research direction covers the social aspects related to the use of the smartphones.
+A popular research direction covers the social aspects related to the use of smartphones.
 W.
 Tan \cite{tan17a} covers all aspects of a touristic experience related to the smartphone, from the definition of travel destination to assistance during the visit.
-Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is considered on the Internet.
-Such an assumption is in contrast with the title, which indicates a nature-based destination, where notably the Internet is not always reachable.
+Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is on the Internet.
+Such an assumption is in contrast with the title, which indicates a outdoor-based destination, where notably the Internet is not always reachable.
 
@@
 Slavec et al.
-investigated the use of cameras \cite{sla21a} while on travel in locations with a relevant cultural heritage to sustain its preservation and engage the tourists using location-based games, similar to Pokemon Go or geo-caching.
+investigated the use of cameras \cite{sla21a} while on travel in locations with a relevant cultural heritage to sustain preservation and engage the tourists using location-based games, similar to Pokemon Go or geo-caching.
 
@@
 published a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}.
-Although the title evokes a one-way contribution, i.e., the impact of digitalization is assumed to be positive on the quality and the sustainability of touristic offers, in the conclusions the authors reveal the awareness for the need to address {\em "the preservation of tourism attractions/sites"} and call for a {\em "a holistic approach ...
+Although the title evokes a one-way contribution, therefore assuming digitalization improves quality and sustainability of touristic offers, in the conclusions the authors reveal the awareness for the need to address {\em "the preservation of tourism attractions/sites"} and call for a {\em "a holistic approach ...
 to support a concept that still lacks conceptual and empirical clarification"}.
@@
 
-Regarding the specific technology we refer to, QR codes have been invented in 1994 for industrial applications.
+Regarding the QR code, the specific technology are going to use, it have been invented in 1994 for industrial applications.
 The current version, described in \cite{isoqr}, is dated February 2015.
@@
 The authors proposed to use the QR code as a link to a panel in PDF format viewed on the user's portable device.
-A similar work in a urban environment is presented in \cite{fin13a} and \cite{tat15a}, the latter illustrating the QR system serving the monuments in the Serbian town of Niš.
-
-Since then the use of the QR code technology for heritage promotion has been suggested in other papers.
+A similar work in an urban environment is presented in \cite{fin13a} and \cite{tat15a}, the latter illustrating the QR system serving the monuments in the Serbian town of Niš.
+
+Since then, the use of the QR code technology for heritage promotion has been suggested in other papers.
 In \cite{sol18a} the authors describe its application in a museum: after defining the points of interest, tags have been placed and linked to related documents.
@@
 
-There are examples of urban outdoor applications of QR tags, like
-
-All the found publications share a common attitude, which can be summarized in the motto "analog portal to digital world", as in the title of \cite{bai12a}.
+All the referenced publications share a common attitude, which can be summarized in the motto "analog portal to the digital world", as in the title of \cite{bai12a}.
 None of them evaluates the possibility of storing text in the tag itself, thus eliminating the requirement of an Internet connection.
 
-The present paper promotes the use of QR codes with sustainability and as the primary or only way of providing information to the visitor.
-In the literature, QR codes are considered ancillary to other forms of communication, like physical installations or the Web.
+The present paper promotes the use of QR codes as the primary or only way of providing information to the visitor.
+Unlike in the literature, where QR tags are considered ancillary to other forms of communication, like physical installations or linked Web contents.
 Therefore, in our case the informational content of the tag, which is commonly ignored, becomes relevant.
-It is a new and original way of using QR codes for the delivery of information.
-
-We emphasize that this article is not meant to quantify user satisfaction or the economic revenue associated with the specific solution.
+It is a new and original way of using a QR code for the delivery of information.
+
+%We emphasize that this article is not meant to quantify user satisfaction or the economic revenue associated with the specific solution.
 Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve a thorough investigation.
@@
 
-The wireless capabilities of a smartphone are the focus of a straightforward approach.
+The wireless capabilities of a smartphone are the focus of a straightforward approach which however shows critical issues.
 Given the premise that the Internet is assumed to be unreachable, the {\em host} stakeholder might provide a local network of low-power radios covering the region of interest.
@@
 
-Such a solution is severely limited by the dependency from a power supply.
+Such a solution is severely limited by the dependency on a power supply.
 A radio transmitter is a rather power-consuming device, in the range of Watts.
@@
 
-The NFC technology is available without installing ad-hoc applications.
-Its operation consists in approaching the smartphone to the tag, and the user is advertised that a content is available.
+Reading a tag does not require an ad-hoc application, and is performed approaching the smartphone to the tag: the user is advertised that content is available.
 The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues.
@@
 
-Another solution in the family of passive devices is the QR tag (QR stands for Quick-Response).
+Another solution in the family of passive devices is the QR code (QR stands for Quick-Response).
 Such a technology does not require radio communication but uses the smartphone camera to acquire text encoded in a graphical code.
@@
 	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone,
-	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation,
+	\item{\em Perceptible information} is a critical requirement, which contrasts with the need to keep the low environmental intrusion.
+This point will be further discussed in the section devoted to the implementation,
 	\item{\em Tolerance for error}: there is no chance of accidentally causing a hazardous situation using the QR tag.
@@
 Compared with other smartphone-based technologies, there is a trade-off concerning capacity.
-However, a signage applications a capacity of 2-300 words is sufficient to provide directions for the visit and convey text-based educational contents.
+However, a signage application with a capacity of 2-300 words is sufficient to provide directions for the visit and convey text-based educational content.
 If capacity limits are not an issue, a solution based on NFC or QR tags exhibits several advantages:
@@
 \begin{itemize} 
-\item does not require power supply,
+\item does not require a power supply,
 \item has a limited impact on the landscape,
@@
 \item is inclusive regarding user inabilities and nationality,
-\item requires basic technical skills from the user.
+\item requires only basic technical skills from the user.
 \end{itemize}
@@
 
-Its use is especially appropriate when the cultural heritage extends across a scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging.
-In these cases, QR tags could improve the visitors’ experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos, even if they don’t have access to a specific spot.
-Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot.
-The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.
+Its use is especially appropriate when the cultural heritage extends across scarcely accessible land, considering that preserving and accessing sites located on private property or in hazardous areas can be challenging.
+In such cases, QR tags could improve the visitors’ experience by providing, when Internet coverage is available, the Web URLs for videos, aerial shots, and photos, even if the specific spot is not accessible.
+Such resources allow visitors to immerse themselves in the area's history and gain an experiential understanding.
 
@@
 Ceramic fragments dated between the 3rd century BC and the 1st-2nd century AD prove that Ligurian populations occupied the region scattered and in small nuclei.
-The dedication of the Latin colony of \textit{Lucca} in 180 BC marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.
+The dedication of the Latin colony of \textit{Lucca} in 180 BC marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by rapid acculturation \cite{gia96, cia05}.
 
@@
 
-The Tourism Density Index is the rate between the number of tourists and the surface area in the higher touristic season.
+The Tourism Density Index is the rate between the number of tourists and the surface area in the higher tourist season.
 It is 106.41 tourists per $km^2$ in 2022.
@@
 
-As shown in \ref{tab:tourism}, the number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay a number of nights more than a weekend and less than a week.
-On the other hand, this data means that people like staying in this tourist comprehensory and they may have found attractions, services and activities they need.
+As shown in \ref{tab:tourism}, the number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay more than a weekend and less than a week.
+On the other hand, this data means how long people likes staying in this tourist region and that they may have found attractions, services, and activities they need.
 Consequently, a well-organized touristic system attracts potentially touristic presences widespread on the municipality territory.
@@
 
-For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study with the purpose of highlighting the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}.
+For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study to highlight the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}.
 As concerns tourist indexes shown in Table \ref{tab:tourism}, the touristic density counts 1866.96 people per square kilometer (1773 $km^2$), with an average stay of 3.36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2.98 nights).
 The provincial area of \emph{Lucca} shows an accommodation density of 19.30 on 10.14 of \emph{Bagni di Lucca}, while the coexistence index of \emph{Lucca} (with 69.87 foreign tourists on 100 Italian tourists) is lower than \emph{Bagni di Lucca}’s coexistence index (with 103.74 foreign tourists in 100 Italian tourists).
-Definitely, it can be claimed that \emph{Bagni di Lucca} is well positioned in the whole touristic context of \emph{Lucca} since it represents one of the less populated municipalities in the territory with interesting tourist indexes.
+It can be claimed that \emph{Bagni di Lucca} is well positioned in the whole touristic context of \emph{Lucca} since it represents one of the less populated municipalities in the territory with interesting tourist indexes.
 In fact, \emph{Bagni di Lucca} boasts its historical thermal tourism tradition together with a well-equipped environment for sport and adventure tourism, on a surface of 164.70 $km^2$, that is one of the biggest municipality areas in the province of \emph{Lucca}.
@@
 
-\emph{Casoli} village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as: canyoning, rafting, and trekking experiences in the Canyon Park.
-Additionally, it features Romanesque churches and huts attracting visitors looking for various experiences such as thermal, environmental, outdoor, sport and cultural activities.
+\emph{Casoli} village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as canyoning, rafting, and trekking experiences in the Canyon Park.
+Additionally, it features Romanesque churches and huts attracting visitors looking for various experiences such as thermal, environmental, outdoor, sports, and cultural activities.
 Throughout the year, local stakeholders offer a variety of services, often incorporating sustainable tourism practices.
-After a detailed analysis of the tourist and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{loz09} dating back to the ‘90s, when some associations began organizing mainly trekking excursions and rafting experience .
+After a detailed analysis of the tourist and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{loz09} dating back to the '90s when some associations began organizing mainly trekking excursions and rafting experience.
 
 As part of \emph{Bagni di Lucca}'s ecosystem, with its focus on tourism, \emph{Casoli} is a destination for slow travel.
-It is important to ensure that any development actions taken are sustainable, in order to preserve its unique social fabric.
-This means paying attention to conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{kuh10a}.
+It is important to ensure that any development actions taken are sustainable, to preserve its unique social fabric.
+This means paying attention to the conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{kuh10a}.
 
@@
 
-Based on the tourism trend of \emph{Casoli} and its overall tourism, and in alignment with the research goals focused on sustainable tourism, it is crucial to consider the role of local policy-makers.
-In fact, they play a vital role in planning sustainable measures and monitoring their impact.
-
-In these terms, as described by [50] Dianne Dredge, and in light of the above analysis, private and public entities can contribute to sustainable tourism by developing the following actions:
+Based on the tourism trend of \emph{Casoli}, and aligned with the research goals focused on sustainability, it is crucial to consider the role of local policy-makers.
+They play a vital role in planning and monitoring sustainable measures and their impact.
+
+In these terms, as described by \cite{dre22} Dianne Dredge, and in light of the above analysis, private and public entities can contribute to sustainable tourism by developing the following actions:
 
@@
 
-In order to create a successful strategic plan, it is crucial to have a thorough understanding of the destination's morphology, environment, history, culture, society, and economy, with sustainability being the key value.
+To create a successful strategic plan, it is crucial to have a thorough understanding of the destination's morphology, environment, history, culture, society, and economy, with sustainability being the key value.
 Policymakers should recognize such needs and act in this way, both politically and socially as described by Beatrix et al.
@@
 
-\subsubsection{How top implement the guidelines? \label{sec:agenda}}
-
-We include three stakeholder communities in the evaluation of the response to the initiative.
+\subsubsection{How to implement the guidelines? \label{sec:agenda}}
+
+According with the above guidelines for tourism valorization, we explain how the research has been conducted dealing with local stakeholders.
+Although information is still incomplete, we were able to collect sufficient opinions to confirm that \emph{Casoli} has the potential to be promoted through a community-involved tourism approach.
+
+We include four stakeholder communities in the evaluation of the response to the initiative.
 We do not speak about {\em success} since the initiative is deliberately soft, an outcome expected in years, and the definition of the evaluation criteria raises new questions.
@@
 	\item {\em the administrators} that are in charge of managing the site resources and that, at due time, will respond to the two stakeholders above, and finally
-	\item {\em the users}, those that come to \emph{Casoli} to visit the Cave of La Piella and that we find the QR-codes on their way
+	\item {\em the users}, those that come to \emph{Casoli} to visit the Cave of \emph{La Piella} and that we find the QR-codes on their way
 \end{itemize}
 
+Considering that each stakeholder category has specific concerns that need to be addressed, our focus has been on understanding the unique characteristics of each category, including their needs and expectations for tourism and environment development.
+
+Regarding residents, we had the chance to meet a few people living in \emph{Casoli}, which did not provide useful information; local entrepreneurs we met informed us about the trekking tourism flows passing through \emph{Casoli} and the \emph{Piella} cave path, which usually starts from the Canyon Park experience.
+
+The administrators of \emph{Bagni di Lucca} Municipality were highly involved in addressing the study's concerns and provided valuable information regarding archaeological and cultural aspects that align with our results.
+
 The QR tag does not record reading operations, preventing an evaluation of the user category.
-A URL was added to the text message in the QR tag to monitor the number of hits.: however, this feedback largely underestimates the number of times the tag is read.
-
-Considering that each stakeholder category has specific concerns that need to be addressed, our focus has been on understanding the unique characteristics of each category, including their needs and expectations for tourism and environment development.
-
-Regarding residents, we had the chance to meet a few people living in \emph{Casoli}, which did not provide useful information; local entrepreneurs we met informed us about the trekking tourism flows passing through \emph{Casoli} and the \emph{Piella} cave path, which usually starts from the Canyon Park experience.
-
-The administrators of \emph{Bagni di Lucca} Municipality were highly involved in addressing the study's concerns and provided valuable information regarding archaeological and cultural aspects that align with our study goals, namely to enhance \emph{Casoli}  area as a sustainable area for hi-tech oriented tourists.
-
-In light of the above assumptions, a reflection on \textit{community-involved tourism} perspective is needed.
-As shown by this empirical study, a small village like \emph{Casoli}  can require a valorization strategy taking advantage of the wider local tourism context.
+A URL was added to the text message in the QR tag to monitor the number of hits: however, this is not an effective usage estimator, as explained in section \ref{sec:results}.
+
+In light of the above, a reflection on \textit{community-involved tourism} perspective is needed.
+As shown by this empirical study, a small village like \emph{Casoli} requires a valorization strategy which takes advantage of the wider local tourism context.
 
@@
 
-Literature on this scientific focus, as described by Richards G. and Derek H. \cite {ric03} gives evidence of local stakeholders' role as ‘designers’ of their living territory.
+Literature on this scientific focus, as described by Richards G.
+and Derek H.
+\cite {ric03} gives evidence of local stakeholders' role as designers of their living territory.
 This is particularly important in inland and marginal tourist areas that require a sustainable approach to development, rather than relying on mass tourism.
@@
 A systemic vision is necessary to manage these areas effectively.
-With these concerns, inner area governance needs a sustainable-led community and policy approach considering socio-cultural local aims for tourism system implementation and environment development.
+With these concerns in mind, inner area governance needs a sustainable-led community and policy approach considering socio-cultural local aims for tourism system implementation and environment development.
 In these terms, local awareness is fundamental for private and public stakeholders enabling sustainable decision-making processes.
@@
 In this regard, we have analyzed the potentiality of involving local administrations and professional tourism operators to create new tourist itineraries by valorizing existing hiking trails.
+
 Firstly, the Municipal Structural Plan should include an archaeological survey; secondly, local tourism stakeholders will benefit from a destination management action to promote environmental and socio-cultural sustainability.
@@
 
-This section defines the points of interest selected to position the QR tags, and reports on their accessibility and state of conservation.
-This may be relevant in a perspective to spot the segment of users that may take advantage of the tag.
+This section defines the points of interest where to position the QR tags, and reports on their accessibility and state of conservation.
+This may be relevant in the perspective of spotting the segment of users that may take advantage of the tag.
 
@@
 
-We placed QR tags both along the trail and at sites of interest; the former are at detours or at close range so that from one you can see the next, as per the CAI standard \cite{cai10}.
+We placed QR tags both along the trail and at sites of interest; the former is at detours or close range so that from one you can see the next, as per the CAI standard \cite{cai10}.
 They can therefore have the function of a simple signpost to visually indicate the continuity of the trail.
@@
 In our prototype, the maximum length of the text amounts to 700 characters.
-A Huawei FIG-LX1 (2017) decodes the tag from a distance of more than 50cm using an Android application available on the Play Store.
+A Huawei FIG-LX1 (2017) decodes the tag from a distance of more than 50cm.
 The same phone can read aloud the content (in Italian) without Internet connectivity;
@@
 \caption{Generation of the QR tag and UMap visualization from UMap content.
-The uMap editor (bottom left) provides a simplified access to the GeoJSON description.
+The uMap editor (bottom left) provides simplified access to the GeoJSON description.
 The same interface allows to specify the template controlling the Web visualization of the map (right side).
-An ad-hoc application referenced in the paper allows to transform the uMap file downloaded from the server into a batch of printable QR tags (top left).
+An ad-hoc application referenced in the paper translates the uMap file downloaded from the server into a batch of printable QR tags (top left).
 The tag is highlighted in red in the picture on the Web display \label{fig:editing}}
@@
 The finalized master document contains a single \emph{GeoJSON} \emph{FeatureCollection} object containing one \emph{Feature} object for each tag.
-Each \emph{Feature} contains a \emph{Geometry} object of type \emph{Point}, and a \emph{properties} object with fields containing information needed to create the tag: its unique identifier \emph{uid}, a random string of six lowercase letters used to produce the associated URL, the \emph{title} printed on the tag, and the text to be encoded in the QR code (see figure \ref{fig:geojson-sample})
+Each \emph{Feature} contains a \emph{Geometry} object of type \emph{Point}, and a \emph{properties} object with fields containing information needed to create the tag: its unique identifier (\emph{fid}) is a random string of six lowercase letters used to produce the associated URL, the \emph{title} printed on the tag, and the text to be encoded in the QR code (see figure \ref{fig:geojson-sample}).
 
@@
 
-The \emph{properties} are also shown on the UMap map.
+%The \emph{properties} are also shown on the UMap map.
 
 Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the UMap user interface already fit the purpose (see the bottom left box in figure \ref{fig:editing}).
-The resulting \emph{GeoJSON} object describing a tagged feature is shown in figure \ref{fig:geojson-sample}.
-
-The master document is rendered by UMap on the page linked to the tags (as shown in figure \ref{fig:editing}).
+
+%The resulting \emph{GeoJSON} object describing a tagged feature is shown in figure \ref{fig:geojson-sample}.
+
+The master document is rendered by UMap on the page linked to the tags as shown in figure \ref{fig:editing}.
 
@@
 
-A \emph{Bash} script running on a Linux system makes straightforward the conversion of the \emph{GeoJSON} file into a printable array of QR tags.
-The code, a total of 90 source lines written in Bash and Python makes use of the \texttt{ogr2ogr} and \texttt{qrencode} commands, and is on GitHub \citeWeb{github-qr}.
+A cross-platform HTML+JavaScript application provides an app that makes straightforward the conversion of the master document into a printable array of QR tags.
+The code does not depend on an Internet connection and is on GitHub \citeWeb{github-qr}.
+
+%A \emph{Bash} script running on a Linux system makes straightforward the conversion of the \emph{GeoJSON} file into a printable array of QR tags.
+The code, a total of 90 source lines written in Bash and Python makes use of the \texttt{ogr2ogr} and \texttt{qrencode} commands and is on GitHub \citeWeb{github-qr}.
 
@@
 
-We aim to enhance the tourist experience in an inland area located around the village of Casoli.
+We aim to enhance the tourist experience in an inland area located around the village of \emph{Casoli}.
 Our goal is tightly related with the fourth pillar of sustainability: there are natural and cultural assets that need to be nurtured or at least preserved.
-We concentrate on a signage system, the realization of which is finalized to reaching such cultural result.
+We concentrate on a signage system, the realization of which is finalized to reach such a cultural result.
 
 To avoid the over-tourism pitfall, we need to take into account the other three pillars: social, environmental, and economic.
-To this end we analyze the social fabric that will host our initiative, the environment where signage will be deployed, and the economic framework that will support its implementation.
-From this study we derive the requirements for our signage system: low environmental impact, open to participation of the residents, responsive to the visitor's information needs, and economically sustainable with a minimal touristic flow.
-
-The approach we applied to reach such conclusions is summarized below: it is conceptual and therefore applicable to a wide range of use cases, and merely consists in designing the system with sustainability in mind from the early stages.
-The simplicity of the approach description must not mislead: its implementation requires collaboration among experts in distinct disciplines converging to the final result.
+To this end, we analyze the social fabric that will host our initiative, the environment where signage will be deployed, and the economic framework that will support its implementation.
+From this study, we derive the requirements for our signage system: low environmental impact, open to participation of the residents, responsive to the visitor's information needs, and economically sustainable with a minimal tourist flow.
+
+The approach we applied to reach such conclusions merely consists of designing the system with sustainability in mind from the early stages.
+The simplicity of the description must not mislead: its implementation is challenging, requiring the collaboration among experts in distinct disciplines converging to a final result.
 
 Our design started by defining a profile for the area according to the \emph{three pillars}.
-The environmental aspect is characterised by natural formations and special habitats hosting a vast biodiversity, placed under special protection according to European legislation.
-
-As for tourism-related economics, a qualitative and quantitative analysis \cite{loz09} reveals unique features that need to be capitalized on to enhance the tourism potential.
-In this light, community-involved tourism
-represents a virtuous approach for destination management in inland areas that need a renewed sustainability-oriented design.
-Therefore, public and private stakeholders are crucial for managing spatial planning and addressing local issues through investments, conservation, promotion and monitoring actions.
+The environmental aspect is characterized by natural formations and special habitats hosting vast biodiversity, placed under special protection according to European legislation.
+
+As for tourism-related economics, a qualitative and quantitative analysis \cite{loz09} reveals unique features that need to be capitalized on to enhance tourism potential.
+In this light, community-involved tourism represents a virtuous approach to destination management in inland areas that need a renewed sustainability-oriented design.
+Therefore, public and private stakeholders are crucial for managing spatial planning and addressing local issues through investments, conservation, promotion, and monitoring actions.
 
@@
 An outcome of the analysis of the social fabric is that, in order to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable.
-Regarding the potential reasons for a visit, we found a cave inhabited since pre-history but reachable with some effort, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
-Such targets offer a variety of access ways, from those requiring performance to the relaxing ones.
-For this reason we opted for not suggesting a geo-itineray, allowing the user cherry-picking locations based on personal interests.
+Regarding the potential reasons for a visit, we found a cave, reachable with some effort, inhabited since pre-history, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
+Such targets offer a variety of access ways, from those requiring performance to relaxing ones.
+For this reason, we opted not to suggest a geo-itinerary, allowing the user cherry-picking locations based on personal interests.
 In this way we expect, legitimated by the results of \cite{sol18a}, that the educational impact of the signage is maximized.
-Furthermore, our choice aims at a conciseness and communicative effectiveness of the tourist information, by inserting within the QR code a text of an average of 460 characters that facilitates readability and immediately provides the information without further steps to web pages, which may slow down the visit or discourage the less motivated visitor.
+Furthermore, our choice aims at concise and effective communication with the tourist  by inserting within the QR code a text of an average of 460 characters that facilitates readability and immediately provides the information without further steps to web pages, which may slow down the visit or discourage the less motivated visitor.
 
@@
 \hfill
-\subfloat[Google Earth view of the surroundings of the tag. The map is north-oriented, and shows the village of \emph{Casoli} in the left-high corner]{
+\subfloat[Google Earth view of the surroundings of the tag. The map is north-oriented and shows the village of \emph{Casoli} in the left-high corner]{
 	\includegraphics[width = 0.62 \linewidth]{figure/GoogleEarth.png}
@@
 
-Previous investigations, including archaeological surveys and access to historical documents and cartography, allowed to design the contents to be delivered through signage.
+Previous investigations, including archaeological surveys and access to historical documents and cartography, allowed to design of the contents to be delivered through signage.
 Two on-site surveys provided further practical indications and the opportunity to place the QR tags.
 
-The main result from our activity is the real scale deployment of the signage, a non trivial exercise in holistic sustainable design.
-
-As a side result we have a signage technology based on QR code tags, a benchmark difficult to improve in terms of environmental impact and cost.
+The main result of our activity is the real-scale deployment of the signage, a non-trivial exercise in holistic sustainable design.
+
+As a side result, we have a signage technology based on QR code tags, a benchmark difficult to improve in terms of environmental impact and cost.
 
@@
 \item I will come back to visit the sites I did not visit today
-\item I am satisfied about my visit
+\item I am satisfied with my visit
 \item I scanned other QR codes in the vicinity
@@
 
-The other is an hits count on the pages linked to the QR-codes.
+The other is a hit count on the pages linked to the QR tags.
 
 Both of them proved to be ineffective.
-We collected only one form with a positive feedback but with no statistical significance.
+We collected only one form with positive feedback but no statistical significance.
 The hit count was polluted by spurious accesses to the page (for instance for maintenance, or reaching the page from the map) and by the fact that scanning the tag was not necessarily linked to a hit.
@@
 
-One technology that fits the requirements of our task is Bluetooth Low Energy (BLE): the autonomy of a battery-operated BLE device is compatible with the periodic maintenance of other forms of signage.
+One technology that fits the requirements of our task is Bluetooth Low Energy (BLE): the autonomy of a battery-operated BLE device is comparable with the periodic maintenance of signage based on passive devices.
 A Bluetooth device can send to the visitor's smartphone a beacon to indicate the presence of a passive sign, like a not-so-visible QR tag, or directly deliver information.
-However this requires the installation of the app on the visitor, and the design of the IT product including the device and the application, which may cost far more than the physical devices.
+However this requires the installation of an ad-hoc on the visitor's device, and the design of the IT product including the BLE device and the application, which may cost far more than the physical devices.
 
 The demonstration of the existence of such needs biases the technological evolution: for instance, if counting NFC reads were a frequent requirement, tags with such a feature would be produced at low costs, and the increment operation might become a default in NFC reading apps.
-The research on low-energy and low-cost devices is also that of demonstrating that the market for such devices is expanding.
+The many research results on low-energy and low-cost devices also indicates an expanding market.
 
@@
 
-Tourism, like other productive activities, is at risk of being unsustainable and damaging the very resources that support it.
-To avoid heavy drawbacks the design of a tourism initiative has to consider the complex framework where it will operate according to an \emph{holistic} approach.
-A methodology made of examples and guidelines helps cope with such a challenging task.
-
-This paper is a step in this direction.
-We address an ordinary problem, that of providing the visitor with directions, with the aim of promoting slow tourism and preserving distinguishing traits of an inland area.
+Tourism, like other productive activities, is at risk of being unsustainable, damaging the very resources that support it.
+To avoid heavy drawbacks the design of a tourism initiative has to consider the complex framework where it will operate according to an holistic approach.
+A methodology made of examples and guidelines helps to cope with such a challenging task.
+
+This paper is a step in that direction. We address an ordinary problem, that of providing the visitor with directions, to promote slow tourism and preserve distinguishing traits of an inland area.
 
@@
 Reporting about the design process is also a complex task, as witnessed by this paper.
-However, the result is rewarding and we demonstrates how the multidisciplinary interplay returns ad-hoc tools targeting sustainability and efficacy.
-
-In the case of \emph{Casoli}, the touristic trends and the local vocation portraits a small destination for visitors practicing heterogeneous touristic circuits in individual and combined ways.
+However, the result is rewarding and we demonstrate how the multidisciplinary interplay returns ad-hoc tools targeting sustainability and efficacy.
+
+In the case of \emph{Casoli}, the touristic trends and the local vocation portray a small destination for visitors practicing heterogeneous touristic circuits in individual and combined ways.
 Cultural sustainability mandates the involvement of local stakeholders in planning new heritage itineraries and supporting initiatives, including signage.
 
-The resulting implementation is minimalistic in its deployment. The detailed description enables its reuse as a starting point in future initiatives.
-
-On the technical side, the discussion regarding the deployed experiment indicates directions for future investigation.
-The definition of a potential application will foster research aiming at passive or very low-energy devices that can effectively advertise their presence and collect statistics.
+The resulting implementation is minimalistic in its deployment.
+This may encourage the direct participation of the residents community in its implementation: from the production of the QR tags to their deployment and maintenance.
+The detailed description of the production process enables its reuse as a starting point in future initiatives.
+
+On a medium term perspective, the description of the deployed experiment gives suggestions for the development of new devices or features for existing ones.
+In our case, a new potential application may foster research aiming at passive or very low-energy devices that can effectively advertise their presence or simply collect statistics in \emph{dead-zones}.
 
@@
 {\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/mariagrazia.png}}}
-{\textbf{Maria Grazia Deri} PhD in Migration and Tourism of the roots at the Department of Political Science of the University of Pisa.
-Researcher on the Geography of Tourism and Human Geography, as well as Destination and Destination Marketing, with interests in cultural and sustainable tourism.
+{\textbf{Maria Grazia Deri} PhD Candidate in Migration and Roots Tourism at the Department of Political Science of Pisa University.
+Researcher on Geography of Tourism and Human Geography, as well as Destination Management and Destination Marketing, with interests in cultural and sustainable tourism.
 Researcher in the framework of European projects about tourism and member of the European Romea Strata Association (AERS).}
@@
 {\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/letizia.png}}}
-{\textbf{Letizia CHiti} Research fellow at the University of Pisa, Department of Civilization and Forms of Knowledge.
+{\textbf{Letizia Chiti} Research fellow at the University of Pisa, Department of Civilization and Forms of Knowledge.
 She graduated in Archaeology at the University of Pisa and obtained a specialization diploma at SISBA with an essay on the archaeological map of the municipality of Bagni di Lucca (LU).
-He participated in projects and surface reconnaissance in mountain areas in Liguria and Tuscany, specializing in non-invasive analysis methods (GIS and study of elevations) with a diachronic approach.}
+She participated in projects and surface reconnaissance in mountain areas in Liguria and Tuscany, specializing in non-invasive analysis methods (GIS and study of elevations) with a diachronic approach.}
 

### 2023-11-09 diff.sh

@@ -1,4 +1,5 @@
 rev="e8af57"
+#rev="9d0fd1"
 
 latexdiff-vc -r $rev paper.tex
 pdflatex paper-diff$rev.tex

### 2023-11-15 paper.tex
Pre trasporto di QR-code

@@
 $^{2}$ \quad University of Pisa; mariagrazia.deri@sp.unipi.it\\
-$^{3}$ \quad University of Pisa; letizia.chiti@}
+$^{3}$ \quad University of Pisa; letizia.chiti@cfs.unipi.it}
 
@@
 \abstract{Understanding the cultural aspects of an area rich in heritage is crucial to building a lasting educational experience from an excursion.
-Many articles in literature explore the use of sophisticated technologies to achieve such a goal.
+Many articles in the literature explore the use of sophisticated technologies to achieve such a goal.
 Tourism proposals for inland areas are significant because the presence of human artifacts and signage can harm the experience and create pollutants.
@@
 % Keywords
-\keyword{Inland areas; heritage trail; digital sustainability; slow tourism; destination governance; community-involved tourism; cultural heritage; holistic approach; QR-code; trail signage}
+\keyword{Inland areas; digital sustainability; slow tourism; destination governance; community-involved tourism; cultural heritage; holistic approach; QR-code}
 
@@
 Several promotional websites, like \citeWeb{tbdl} and \citeWeb{valdilima} advertise \emph{Bagni di Lucca} attractions.
-The sitography closing this paper contains a longer yet incomplete list of touristic initiatives  \citeWeb{agu,r2o,ronda,garfraft,valdilima,canyonpark,experience,ebikea,ebikel,bdlt2,bdl}.
+The sitography closing this paper contains a longer yet incomplete list of touristic initiatives  \citeWeb{valdilima,agu,r2o,ronda,garfraft,canyonpark,experience,ebikea,ebikel,bdlt2,bdl}.
 
@@
 We aim to enhance the tourist experience in an inland area located around the village of \emph{Casoli}.
-Our goal is tightly related with the fourth pillar of sustainability: there are natural and cultural assets that need to be nurtured or at least preserved.
-We concentrate on a signage system, the realization of which is finalized to reach such a cultural result.
+Our goal is closely related to the fourth pillar of sustainability: there are cultural assets that need to be nurtured or at least preserved.
+The proposed signage system aims to achieve such a goal by promoting the documentation and touristic fruition of these assets.
 
@@
 To this end, we analyze the social fabric that will host our initiative, the environment where signage will be deployed, and the economic framework that will support its implementation.
-From this study, we derive the requirements for our signage system: low environmental impact, open to participation of the residents, responsive to the visitor's information needs, and economically sustainable with a minimal tourist flow.
+From this study, we derive the requirements for our signage system: low environmental impact, open to the participation of the residents, responsive to the visitor's information needs, and economically sustainable with minimal tourist flow.
 
 The approach we applied to reach such conclusions merely consists of designing the system with sustainability in mind from the early stages.
-The simplicity of the description must not mislead: its implementation is challenging, requiring the collaboration among experts in distinct disciplines converging to a final result.
+The simplicity of the description must not mislead: its implementation is challenging, requiring collaboration among experts in distinct disciplines to converge to a final result.
 
@@
 
-An outcome of the analysis of the social fabric is that, in order to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable.
+An outcome of the analysis of the social fabric is that, to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable.
 Regarding the potential reasons for a visit, we found a cave, reachable with some effort, inhabited since pre-history, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
 Such targets offer a variety of access ways, from those requiring performance to relaxing ones.
-For this reason, we opted not to suggest a geo-itinerary, allowing the user cherry-picking locations based on personal interests.
+For this reason, we opted not to suggest a geo-itinerary, allowing the user to cherry-pick locations based on personal interests.
 In this way we expect, legitimated by the results of \cite{sol18a}, that the educational impact of the signage is maximized.
@@
 %\nociteWeb{*}
-\bibliographystyleWeb{abbrv}
+\bibliographystyleWeb{unsrt}
 \bibliographyWeb{biblio/sitography.bib}

### 2023-11-16 paper.tex
Finale, pre-grammarly

@@
 
-The conclusive section summarizes the results, the missing parts, and the potential of the solution.
+The conclusive section summarizes the results, the pending tasks, and the potential of the solution.
 
@@
 
-Regarding the QR code, the specific technology are going to use, it have been invented in 1994 for industrial applications.
-The current version, described in \cite{isoqr}, is dated February 2015.
-In 2013 an early paper by Martinez et al.
-\cite{mar13a} suggests its utilization to implement a virtual tour in the {\em Las Quilamas Natural Park} in Spain.
-At that time the availability of QR-code readers was not as diffused as today, and therefore the work may be considered among the seminal ones.
-The authors proposed to use the QR code as a link to a panel in PDF format viewed on the user's portable device.
-A similar work in an urban environment is presented in \cite{fin13a} and \cite{tat15a}, the latter illustrating the QR system serving the monuments in the Serbian town of Niš.
-
-Since then, the use of the QR code technology for heritage promotion has been suggested in other papers.
-In \cite{sol18a} the authors describe its application in a museum: after defining the points of interest, tags have been placed and linked to related documents.
-The authors conducted a poll about user perception on two segments of users: Italians and Polish.
-The QR tags were not unanimously considered useful: in particular half of the Italian users considered them as not useful.
-It is interesting to note that the motivation of the other segment of users, less skeptical about usefulness, shared a solid educational motivation to access QR-linked content (70\% in the Polish group against 50\% in the Italian one).
-This fact sheds light on the necessity to provide content the user needs: curiosity is not a sufficient motivation.
-
-Recent publications about the use of QR codes for heritage valorization do not show significant novelties.
-In 2023, the authors of \cite{pan23a} introduce the installation of six QR codes in a museum and analyze the reactions of 42 visitors.
-The tags are created using the Taplink \url{https://taplink.at/} web service and are linked to content on the Web.
-
-All the referenced publications share a common attitude, which can be summarized in the motto "analog portal to the digital world", as in the title of \cite{bai12a}.
-None of them evaluates the possibility of storing text in the tag itself, thus eliminating the requirement of an Internet connection.
-
-The present paper promotes the use of QR codes as the primary or only way of providing information to the visitor.
-Unlike in the literature, where QR tags are considered ancillary to other forms of communication, like physical installations or linked Web contents.
-Therefore, in our case the informational content of the tag, which is commonly ignored, becomes relevant.
-It is a new and original way of using a QR code for the delivery of information.
-
 %We emphasize that this article is not meant to quantify user satisfaction or the economic revenue associated with the specific solution.
@@
 
-Its use is especially appropriate when the cultural heritage extends across scarcely accessible land, considering that preserving and accessing sites located on private property or in hazardous areas can be challenging.
+The QR code have been invented in 1994 for industrial applications. The current version, described in \cite{isoqr}, is dated February 2015.
+
+In our use case it is especially appropriate when the cultural heritage extends across scarcely accessible land, considering that preserving and accessing sites located on private property or in hazardous areas can be challenging.
 In such cases, QR tags could improve the visitors’ experience by providing, when Internet coverage is available, the Web URLs for videos, aerial shots, and photos, even if the specific spot is not accessible.
@@
 
-%As for tourism-related economics, a qualitative and quantitative analysis reveals unique features that need to be capitalized on to enhance the tourism potential.
-
 An outcome of the analysis of the social fabric is that, to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable.
-Regarding the potential reasons for a visit, we found a cave, reachable with some effort, inhabited since pre-history, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
+
+A rationale for the above attitude may be found in an article by Solima et al.
+\cite{sol18a}.
+The authors conducted a poll about user perception on two segments of users: Italians and Polish.
+The feelings about a QR based signage were not unanimous, and half of the Italian users considered it as not useful.
+Another result from the poll reveals that Polish users showed a stronger motivation to use tags for educational purposes (70\% versus 50\%), while Italians were mainly driven by curiosity.
+We conclude that, to provide a useful service, the signage must address a motivated user.
+
+So we identified the potential reasons for a visit.
+We found a cave reachable with some effort inhabited since pre-history, a village with historical buildings, and nearby architecture dating back to the Middle Ages.
 Such targets offer a variety of access ways, from those requiring performance to relaxing ones.
 For this reason, we opted not to suggest a geo-itinerary, allowing the user to cherry-pick locations based on personal interests.
-In this way we expect, legitimated by the results of \cite{sol18a}, that the educational impact of the signage is maximized.
-Furthermore, our choice aims at concise and effective communication with the tourist by inserting within the QR code a text of an average of 460 characters that facilitates readability and immediately provides the information without further steps to web pages, which may slow down the visit or discourage the less motivated visitor.
 
@@
 The impact on the environment is extremely limited, as shown in figure \ref{fig:TagPlacement}.
+
+To allow readability in {\em dead zones} and quickly provide the information without further steps to web pages, we encode the content within the QR code itself.
+The content consists of a text of an average of 460 characters aiming at a concise and effective communication with the tourist.
+
+Therefore, in our design, the information encoded in the QR tag is the primary or only way of communication with the visitor.
+It is a new and original way of using such technology, since QR tags are instead considered ancillary to other forms of communication, like physical installations or linked Web contents, according with the motto "analog portal to the digital world", as in the title of \cite{bai12a}.
+
+Such an attitude has not much changed since the first reported utilization of QR tags for heritage promotion.
+In 2013 an early paper by Martinez et al.
+\cite{mar13a} suggests its utilization to implement a virtual tour in the {\em Las Quilamas Natural Park} in Spain.
+At that time the availability of QR-code readers was not as diffused as today, and therefore the work may be considered among the seminal ones.
+The authors proposed to use the QR code as a link to a panel in PDF format viewed on the user's portable device.
+Similar works in an urban environment are presented in \cite{fin13a} and \cite{tat15a}, the latter illustrating the QR system serving the monuments in the Serbian town of Niš.
+Among indoor applications we already referenced \cite{sol18a}, which describe the use of QR tags linked to Web resources in two museums in different countries.
+
+The role of QR tags did not change from those early days: in 2023, the authors of \cite{pan23a} introduce the installation of six QR codes in a museum and analyze the reactions of 42 visitors.
+The tags are created using the Taplink \url{https://taplink.at/} web service and are linked to content on the Web.
+
+We conclude that storing informational content in the tag itself is not considered as an option, which justifies the novelty of our approach.
 
@@
 
-Previous investigations, including archaeological surveys and access to historical documents and cartography, allowed to design of the contents to be delivered through signage.
+The results of previous multi-disciplinary investigations, including archaeological surveys and access to historical documents and cartography, were used to define such informational content.
 Two on-site surveys provided further practical indications and the opportunity to place the QR tags.
@@
 %\nociteWeb{*}
-\bibliographystyleWeb{unsrt}
+\bibliographystyleWeb{abbrv}
 \bibliographyWeb{biblio/sitography.bib}

### 2023-11-16 diff.sh

@@ -1,5 +1,6 @@
-rev="e8af57"
+#rev="e8af57"
 #rev="9d0fd1"
+rev="d53b296" # solo trasferimento QR-code
 
 latexdiff-vc -r $rev paper.tex
 pdflatex paper-diff$rev.tex

### 2023-11-22 paper.tex
aggiunto source combo_gen

@@
 Section \ref{sec:implementation} describes the concrete implementation that follows the defined requirements.
-The description of the process to produce the printed tags starting from editable content on a geographic database allows reproduction and reuse of the solution in other contexts.
+The description of the process to produce the printed tags starting from editable content on a geographic database allows the reproduction and reuse of the solution in other contexts.
 
@@
 Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is on the Internet.
-Such an assumption is in contrast with the title, which indicates a outdoor-based destination, where notably the Internet is not always reachable.
+Such an assumption is in contrast with the title, which indicates an outdoor-based destination, where notably the Internet is not always reachable.
 
@@
 published a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}.
-Although the title evokes a one-way contribution, therefore assuming digitalization improves quality and sustainability of touristic offers, in the conclusions the authors reveal the awareness for the need to address {\em "the preservation of tourism attractions/sites"} and call for a {\em "a holistic approach ...
+Although the title evokes a one-way contribution, therefore assuming digitalization improves the quality and sustainability of touristic offers, in the conclusions the authors reveal the awareness for the need to address {\em "the preservation of tourism attractions/sites"} and call for a {\em "a holistic approach ...
 to support a concept that still lacks conceptual and empirical clarification"}.
@@
 
-Reading a tag does not require an ad-hoc application, and is performed approaching the smartphone to the tag: the user is advertised that content is available.
+Reading a tag does not require an ad-hoc application, and is performed by approaching the smartphone to the tag: the user is advertised that content is available.
 The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues.
@@
 
-The QR code have been invented in 1994 for industrial applications. The current version, described in \cite{isoqr}, is dated February 2015.
-
-In our use case it is especially appropriate when the cultural heritage extends across scarcely accessible land, considering that preserving and accessing sites located on private property or in hazardous areas can be challenging.
+The QR code was invented in 1994 for industrial applications. The current version, described in \cite{isoqr}, is dated February 2015.
+
+In our use case, it is especially appropriate when the cultural heritage extends across scarcely accessible land, considering that preserving and accessing sites located on private property or in hazardous areas can be challenging.
 In such cases, QR tags could improve the visitors’ experience by providing, when Internet coverage is available, the Web URLs for videos, aerial shots, and photos, even if the specific spot is not accessible.
@@
 
-From a historical-geographical point of view, this area is identified with the \emph{Lima} stream and its tributaries, which impress the area with a peculiar geomorphology impervious despite the modest hilly elevations.
+From a historical-geographical point of view, this area is identified with the \emph{Lima} stream and its tributaries, which impress the area with peculiar geomorphology, impervious despite the modest hilly elevations.
 
@@
 The village of \emph{Casoli} is located south of the \emph{Lima} stream on a hill named {\em "Tanette"}.
-In the local small lair, the name reveals the presence of karstic cavities, some inhabited between the Paleolithic and the Iron Age and used also as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.
+In the local small lair, the name reveals the presence of karstic cavities, some inhabited between the Paleolithic and the Iron Age and also used as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.
 
@@
 We have little evidence of Roman settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of \textit{Buca La Piella}, investigated in 1975 by the Centre for Archaeological Studies of \emph{Lucca}; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside.
-In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found \cite{gia96, men81, cia03}.
+In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins from the 3rd century AD, bronze and lead objects were found \cite{gia96, men81, cia03}.
 
@@
 As shown in \ref{tab:tourism}, the number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay more than a weekend and less than a week.
-On the other hand, this data means how long people likes staying in this tourist region and that they may have found attractions, services, and activities they need.
+On the other hand, this data means how long people like staying in this tourist region and that they may have found attractions, services, and activities they need.
 Consequently, a well-organized touristic system attracts potentially touristic presences widespread on the municipality territory.
@@
 
-From the point of view of our research, the context of \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well as touristic services).
+From the point of view of our research, the context of \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as anthropic, socio-cultural, and historical features, as well as touristic services).
 Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of \emph{Lucca} and the \emph{Garfagnana} mountain area.
@@
 
-According with the above guidelines for tourism valorization, we explain how the research has been conducted dealing with local stakeholders.
+According to the above guidelines for tourism valorization, we explain how the research has been conducted dealing with local stakeholders.
 Although information is still incomplete, we were able to collect sufficient opinions to confirm that \emph{Casoli} has the potential to be promoted through a community-involved tourism approach.
@@
 In light of the above, a reflection on \textit{community-involved tourism} perspective is needed.
-As shown by this empirical study, a small village like \emph{Casoli} requires a valorization strategy which takes advantage of the wider local tourism context.
+As shown by this empirical study, a small village like \emph{Casoli} requires a valorization strategy that takes advantage of the wider local tourism context.
 
@@
 \includegraphics[width =\linewidth]{figure/combo_gen.jpeg}
-\caption{Generation of the QR tag and UMap visualization from UMap content.
+\caption{Generation of the QR tag and uMap visualization from uMap content.
 The uMap editor (bottom left) provides simplified access to the GeoJSON description.
@@
 
-Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the UMap user interface already fit the purpose (see the bottom left box in figure \ref{fig:editing}).
+Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the uMap user interface already fit the purpose (see the bottom left box in figure \ref{fig:editing}).
 
@@
 
-The master document is rendered by UMap on the page linked to the tags as shown in figure \ref{fig:editing}.
+The master document is rendered by uMap on the page linked to the tags as shown in figure \ref{fig:editing}.
 
@@
 The authors conducted a poll about user perception on two segments of users: Italians and Polish.
-The feelings about a QR based signage were not unanimous, and half of the Italian users considered it as not useful.
+The feelings about a QR-based signage were not unanimous, and half of the Italian users considered it as not useful.
 Another result from the poll reveals that Polish users showed a stronger motivation to use tags for educational purposes (70\% versus 50\%), while Italians were mainly driven by curiosity.
@@
 To allow readability in {\em dead zones} and quickly provide the information without further steps to web pages, we encode the content within the QR code itself.
-The content consists of a text of an average of 460 characters aiming at a concise and effective communication with the tourist.
+The content consists of a text of an average of 460 characters aiming at concise and effective communication with the tourist.
 
 Therefore, in our design, the information encoded in the QR tag is the primary or only way of communication with the visitor.
-It is a new and original way of using such technology, since QR tags are instead considered ancillary to other forms of communication, like physical installations or linked Web contents, according with the motto "analog portal to the digital world", as in the title of \cite{bai12a}.
+It is a new and original way of using such technology since QR tags are instead considered ancillary to other forms of communication, like physical installations or linked Web contents, according with to motto "analog portal to the digital world", as in the title of \cite{bai12a}.
 
@@
 A Bluetooth device can send to the visitor's smartphone a beacon to indicate the presence of a passive sign, like a not-so-visible QR tag, or directly deliver information.
-However this requires the installation of an ad-hoc on the visitor's device, and the design of the IT product including the BLE device and the application, which may cost far more than the physical devices.
+However, this requires the installation of an ad-hoc on the visitor's device, and the design of the IT product including the BLE device and the application, which may cost far more than the physical devices.
 
 The demonstration of the existence of such needs biases the technological evolution: for instance, if counting NFC reads were a frequent requirement, tags with such a feature would be produced at low costs, and the increment operation might become a default in NFC reading apps.
-The many research results on low-energy and low-cost devices also indicates an expanding market.
+The many research results on low-energy and low-cost devices also indicate an expanding market.
 
@@
 Tourism, like other productive activities, is at risk of being unsustainable, damaging the very resources that support it.
-To avoid heavy drawbacks the design of a tourism initiative has to consider the complex framework where it will operate according to an holistic approach.
+To avoid heavy drawbacks the design of a tourism initiative has to consider the complex framework where it will operate according to a holistic approach.
 A methodology made of examples and guidelines helps to cope with such a challenging task.
@@
 
-On a medium term perspective, the description of the deployed experiment gives suggestions for the development of new devices or features for existing ones.
+From a medium-term perspective, the description of the deployed experiment gives suggestions for the development of new devices or features for existing ones.
 In our case, a new potential application may foster research aiming at passive or very low-energy devices that can effectively advertise their presence or simply collect statistics in \emph{dead-zones}.
@@
 Previously active on Grid and, later, Cloud computing in collaboration with INFN.
-Until 2023, teaching activity at the Degree Course in Computer Humanities of the University of Pisa with two courses respectively on Network Protocols and the Angular 2+ framework.}
+Until 2023, he was a teacher at the Degree Course in Computer Humanities at the University of Pisa, where he taught two courses, one on Network Protocols and the other on the Angular 2+ framework.}
 % For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references

### 2023-11-22 diff.sh

@@ -1,7 +1,8 @@
 #rev="e8af57"
 #rev="9d0fd1"
-rev="d53b296" # solo trasferimento QR-code
-
+rev=0d1563d # in prima revisione
+#rev="d53b296" # solo trasferimento QR-code
+rev=0d1563d
 latexdiff-vc -r $rev paper.tex
 pdflatex paper-diff$rev.tex
 bibtex paper-diff$rev
